### 2024-01-24 Markdown/Manuscript/Child_RMDs/Clippings.Rmd
first draft of reworked ethics paper

---
title: "Clippings"
output: html_document
date: "2023-04-18"
---

<!-- . The authors of that framework propose 13 questions to evaluate ethical risk across the four criteria of  -->
<!-- . However, while our six assessment criteria are specific to three attributes (TRU) of the computational implementation of the model, the prior ethical framework includes questions relevant to the conceptual and mathematical models and the potential impacts of model use. Examples of these more general evaluation questions include (for the justice criterion) “is any lack of knowledge about important parameters attributable to uncertainty or variability?” and (for the beneficence criterion) “if a policy is based on the model evidence, is it more likely to be effective and beneficial than a decision made in the absence of the model?”.  -->

<!-- Multiple ethical considerations of computational modeling have previously been identified [@10.3389/fpubh.2017.00068; @pliakos2021ethics; @thompson2022escape; @calder2018computational],  -->

<!-- For example, to reduce the risk of inequitable policy implementation, it may be important for of a model to predict the benefits of harms of an intervention for different sub-populations [@10.3389/fpubh.2017.00068], but model developers may prefer to allocate a project budget to implementing other model features [@thompson2022escape]. -->

<!-- Reuse of models as components of other models can also make model implementation more efficient [@Arnold2010; @garcia2021cost] but the reusability of models is significantly shaped by computational implementation choices such as how model code and data are authored, documented, licensed and disseminated.  -->

<!-- Common barriers to model re-use include commercial and legal considerations [@Feenstra2022; @Pouwels2022], as well as challenges related to model transferability across jurisdictions [@garcia2021cost]. -->

<!-- Compared to using commercial modelling software, authoring CHEMs in an open-source language like R [@RCORE2022] makes it easier to -->
<!-- Using open-source development platforms and licenses can aid both generalizability and transferability. -->
<!-- The choices that CHEM developers make about model implementation and licensing will also shape who can use a model and for what purposes.  -->
<!-- are relatively easier to selectively modify. -->

<!-- Currently, the core development team for a CHEM will be typically be funded to produce a project end-point deliverable whose specifications are well defined early in the project. For more complex and multi-purpose CHEMs, particularly those designed to be incorporated into decision support systems, it may be better for development teams to adopt Agile Software Development [@beck2001manifesto], an approach that has been recommended for complex public health software projects [@10.3389/fpubh.2022.899874]. An Agile model will be less clearly specified in the initial project plan, but will instead continually develop in response to the requirements and feedback of users, who are provided with an initial, simplified working version of the model at the earliest feasible opportunity. -->

<!-- that have the potential to be met with the assistance of software frameworks -->

<!-- We have designed this framework to enable ethical computational implementations of health economic models that are both open-source (for transparency and reusability) and modular (for reusability and updatability). The framework aims to support the standardization and interoperability that are core enablers of implementing a modular and open source approaches.  -->

<!-- Functions to implement model algorithms can be associated with a module via . -->

<!-- to provide tools for authoring CHEMs, supplying those CHEMs with data and using CHEMs to implement reproducible modelling analyses. -->

<!-- The ready4pack library depends on two other module authoring libraries. <!-- MOVED CONTENT -->

<!-- Each CHEM module authored with the ready4 framework will include both a data structure (specifying the required properties of data that can validly be supplied to a module) and a set of algorithms (specifying the operations that can be performed on data contained in a module instance). -->

<!-- development version -->

<!-- To facilitate its application to our youth mental health project, we established and configured accounts with these required services (see Availability of data and materials). -->

<!-- Talk about DARTH (https://doi.org/10.1007/s40273-019-00837-x )/ BCEA -->

<!-- The less numerous and more focused assessment criteria we propose may potentially be more tractable to implement in reviews of models authored by third parties and as the basis for designing software frameworks to support ethical computational model implementation. -->

<!-- Two issues for funders - allow support for beneficial use, foster viable communities. -->

#################################

Computational models, particularly those addressing economic topics, have become essential tools for health policy development [@dakin2015influence; @Erdemir2020]. Although influential and widely used, health economic models typically have a number of limitations that at best restrict their usefulness and which in some cases may facilitate harmful misuse.

There is a strong case for making health economists more accountable for the appropriate use and social acceptability of their models. Potential users of models should be able to assess their adequacy for a particular purpose [@thompson2019escape].  This goal is difficult to achieve in the current context of the poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017] and frequently insufficient validation [@Ghabri2019] of health economic models. Many health economic models are released with undeclared errors [@Radeva2020]. The value judgments that strongly shape health economic analyses are rarely made explicit, omissions that may lead to socially unacceptable policy recommendations [@duckett2022journey]. A modelling team's value judgments about what questions to address, the most important features of a system to represent and the weighting of different types of evidence may be poorly aligned with those of the people impacted by decisions informed by model analyses [@thompson2022escape]. As health economic models adopt more sophisticated techniques, the need for accountability grows. More complex models may be more prone to propagation errors [@Saltelli2019] and models designed to address multiple questions should be expected to meet more onerous verification and validation obligations [@Eddy2012; @Feenstra2022]. The nature and extent of individual model authorship contributions may be less clear in models implemented over longer time-frames with a large and changing group of collaborators [@thompson2022escape].

The resource and skills intensive undertaking of modelling could be made more tractable if it was easier for health economists to re-use each other's models [@Arnold2010]. However, as many health economic models are owned by pharmaceutical companies and consultancies, commercial considerations can limit the reuse of models and their constituent code and data [@Feenstra2022]. Legal and ethical issues such as privacy and confidentiality requirements further limit public release of some model artefacts [@Pouwels2022]. Transferring a health economic model developed for one jurisdiction for application in another typically involves retaining some features and updating others [@barbieri2010international]. However, models do not always make implementation choices (for example, use of concepts with standardised meanings across jurisdictions<!-- [@garcia2021cost] -->, supplying context specific data as replaceable data-packs and distributing source code under licenses that allow derivative works) that facilitate model transferability.

Health economic models should be updated and refined as new evidence emerges and decision contexts change [@Jenkins2021], but this occurs infrequently [@Sampson2017]. Funding for health economic modelling projects rarely extend to provision of medium term support for model updates and improvements. The career trajectories of health economists can also mitigate against adequate maintenance of a model, it being relatively common for model authors to have moved on from the team that owns the model and / or from working on the health condition for which the model was developed.

A potential strategy for improving model accountability, reusability and updatability is to make models both modular and open-source.

A modular model is constructed from multiple reusable and replaceable sub-models (modules) [@pan2021modular]. Modules can share inputs and outputs with other modules or can be run as independent models [@barros2023empowering]. Advantages of modular models include feasibility (large projects are broken into smaller tasks, with each component independently developed and tested) and flexibility (making it easier to selectively replace or update specific parts of a model and to scale up or down the level of granularity) [@pan2021modular]. Modular approaches are currently being used to facilitate the development of complex computational models in disciplines such as biology [@pan2021modular], ecology [@barros2023empowering] and neuroscience [@frazier2022robust]. In health economics the related and enabling concept of reference models has been recommended [@Afzali2013], but peer reviewed studies describing modular health economic models remain relatively rare, though examples exist in infectious disease [@Trauer2017] and cardiology [@Urach2013].

Modular models provide an opportunity for multiple modelling teams to contribute to, test and reuse models. To enhance this capacity, modular models may be implemented as open source projects that give others liberal permissions to access and use model source code and data [@pan2021modular; @barros2023empowering; @frazier2022robust]. Although there appears to be in principle support from many health economists for greater use of open source health economic models (OSHEMs)  [@Pouwels2022], actual implementations are rare [@Emerson2019; @Michalczyk2018; @Feenstra2022]. Barriers to adoption of OSHEMs include concerns about intellectual property, confidentiality, model misuse and the resources required to support open source implementations [@Pouwels2022; @Wu2019]. Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020], but guidelines for implementing OSHEMs remain scarce, piecemeal and need improving [@Sampson2019].

Our interest in modular and open source approaches developed when we began seeking am appropriate framework for undertaking and validly synthesising diverse types of economic research in mental health. Mental disorders impose high health, social and economic burdens worldwide [@RN8;@GBD2019]. Much of this burden is potentially avertable [@RN25], but poorly financed and organised mental health systems are ill-equipped for this challenge [@RN22;@RN23]. A substantial economic literature already exists to assess the affordability and value for money of mental health interventions [@RN34]. This economic evaluation work is an essential prerequisite for improving allocative efficiency in mental health, but could be of greater value to systems planners if integrated with a broader program of economic research.

We are developing ready4 (https://www.ready4-dev.com), a modular OSHEM (MOSHEM) on the mental health of young people aged 12 to 25. We are initially applying ready4 to four of the twelve domains of health economics identified by Wagstaff and Culyer [@wagstaff2012four]:
<!-- There is therefore an opportunity for health economists to undertake mental health research of significant potential public benefit. -->

- *health and its value* (our projects: utility mapping models);

- *determinants of health and ill-health* (our projects: models for creating synthetic household populations with key risk and protective factors for mental disorders);

- *demand for health and health care* (our projects: spatial epidemiology and help-seeking choice models); and

- *supply of health services* (our projects: a model of primary mental health care services).

Once these projects are completed, our aim is to flexibly combine these models to answer questions in two additional Wagstaff and Culyer domains:

- *efficiency and equity* (our goal: assess the distributional impacts and identify the optimal targeting of care provision); and

- *economic evaluation* (our goal: assess the cost-utility of competing policy options for improving the mental health of young people).

Standardisation is an essential enabler of automation, collaboration, interoperability and transferability in modular and open source approaches. An early requirement to implement ready4 is therefore to identify an appropriate set of standards. However, we are not aware of any consolidated source of guidance about recommended standards for MOSHEMs. Once identified, the standards specified for ready4 will need to be implemented by software, some of which will require custom development.

<!-- When planning how to undertake and synthesise this program of work, we became concerned with issues relating to accountability, reusability and updatability. If mistakes are common in individual health economic modelling projects [@Radeva2020], then combining multiple projects may be particularly error prone. The risk of errors may also depend on whether a model is too simple or too complex [@Saltelli2019], so the level of granularity appropriate for one model application may be ill-suited to another. Undertaking a multiannual program of interdependent work raises the risk that models developed in the first phase of a project may need to be updated by the time the project is approaching conclusion. -->

<!-- These concerns led us to seek the MOSHEM approach that provides the transparency to make it easier for errors to be uncovered, the adaptability to fix these errors and the flexibility to facilitate interchangeability between representations of different levels of granularity.  -->

### Standards for an accountable MOSHEM
Guidance on transparency in health economic modelling published over ten years ago [@Eddy2012] made recommendations on documenting models but notably did not include recommendations on sharing model code and data. However, more recent and multidisciplinary healthcare modelling guidance [@Erdemir2020] recommends using existing digital repository services to make these types of digital model artefacts publicly available. Some repositories such as GitHub [@github2007] provide tools for disseminating work in progress code and providing highly transparent records of the complete development history and individual authorship contributions of a software project. Other repositories such as Zenodo [@Zenodo2013] and Dataverse [@Dataverse2007] provide persistent storage solutions that generate a Digital Object Identifier (DOI) for each code and data collection.

Model code and data should be clearly documented, potentially with different versions for technical and non-technical users [@Eddy2012]. Consistent use of meaningful naming conventions when authoring code is recommended [@Wilson_2017; @Alarid2019]. Code can be made easier to follow by using the practices of abstraction [@8717448], where only simple, high level commands are routinely exposed to reviewers, and polymorphism [@7181447], where the same command (e.g. "simulate") can be reused to implement different algorithms of the same type. Programs to implement model analyses can be made comprehensible to even non-technical users through the use of literate programming techniques and tools like RMarkdown [@xie2018r] that integrate computer code with plain English descriptions.

An essential component of quality assuring health economic models is verification - ensuring that calculations are correct and consistent with model specifications [@techver2019]. One useful concept for informing model users about the extensiveness of verification checks is code coverage [@ERICWONG2010188] - the proportion of model code that has been explicitly tested. <!-- In R, the testthat [@testthat2011] and covr [@covr2020] tools can be used in conjunction with GitHub to define tests and report coverage metrics. -->Transcription errors - mistakes introduced when transferring data between sources, models and reports - are very common in health economic models [@Radeva2020]. The risk of these errors might be lower if there was full transparency across all steps in a study workflow. Scientific computing tools now make it relatively straightforward to author programs that reproducibly execute all steps in data ingest, processing and reporting [@Wilson_2017].

Code and data should be distributed with tools that make it easy for potential users to appropriately cite each model artefact.

### Standards for a reusable MOSHEM
```{r echo = F}
gpl_pc_dbl <- get_license_share("GPL")
```
To make model code and data widely re-usable by others, it is important to provide users with appropriate and explicit permissions. In the context of open source models, there are two broad categories of licensing options. Some guidance strongly recommends the use of permissive licensing [@Wilson_2017] that provides users with great flexibility as to the purposes (including commercial) for which the content could be re-used. An alternative approach is to use copyleft licenses [@copyleft2022] that can require content users to distribute any derivative works they create under similar open source arrangements. For code, it may be appropriate to adopt the prevailing open source licensing practice within the programming language being used. For data, it may not be sufficient to simply choose between a permissive license like the Public Domain Dedication (CC0) [@cc02022] or a copyleft option such as the Attribution-Share Alike (CC-BY-SA) [@bysa2022]. In addition to ensuring that data is ethically appropriate for disseminate in open access repositories, responsible custodianship of some de-identified or aggregated data may involve using or adapting template terms of use [@sampleterms2022] which have a number of ethical clauses (for example, prohibiting efforts to re-identify research participants).

Storing model code and data in separate files and locations (as opposed to hard coding - embedding data into source code) can make it easier to apply models to different decision contexts and to selectively restrict access to data that are confidential, while disseminating all other model artefacts. Clear distinctions should be made between model modules (code that defines abstract data structures and the algorithms that can be applied to data described by these structures), model datasets (digital information such as parameter values, unit records, etc) and model analyses (code that links model datasets to model modules and specifies the algorithms to apply to data associated with each module).

The software development practice of encapsulation [@8717448] can be used to help ensure that model modules continue to work as intended when they are combined [@ready4oop2022]. In some cases, combining modules may mean new versions of modules have to be created to better account for interaction effects. The concept of inheritance [@8717448] can be used to write code that efficiently achieves this objective as well as to facilitate selective editing of modules when transferring models to different decision contexts [@ready4oop2022]. Writing algorithms as collections of functions (short, self-contained and reusable software routines that each perform a discrete task) is recommended as good practice for scientific computing [@Wilson_2017]. Functions to implement model algorithms can be associated with data structures (also known as a class) via a special type of function called a method. Model modules of a similar type or purpose can be efficiently distributed and documented by bundling them as code libraries. It is good practice to make available test or toy data to demonstrate the use of model algorithms [@Wilson_2017].

Statistical models are a common output of health economic evaluations, but they are often not reported in a format that enables others to confidently and reliably re-use them for out of sample prediction [@Kearns2013]. Open source approaches can help address this by disseminating code artefacts that enable easy and appropriate use of a statistical model to make predictions with new data. However, great care must be exercised when publicly releasing model artefacts derived from data on human subjects as they may by default embed a copy of the source dataset. Sensitive dataset copies must therefore be replaced (for example, with synthetic data) and the amended artefact's predictive performance then retested before any public release. Another way to make MOSHEMs easier to use is to develop simple user-interfaces for non-technical users.

### Standards for an updatable MOSHEM
To avoid MOSHEMs going stale - losing validity and usefulness with time - they should be routinely updated.  Each update of code and data should be uniquely identifiable and retrievable, a goal that can be facilitated by use of version control tools [@Erdemir2020] such as git [@git20XX].

Potential users of model code and data should be able to easily identify the appropriateness of a version to their needs. Using semantic versioning [@semver20xx] conventions can signal the potential importance of an update to users of model code and data. Also informative is to clearly label the type of code versions hosted in a repository as either:

- "development" - typically the most comprehensive and up to date source code, but potentially not yet sufficiently well documented and tested for potential users to apply to purposes other than testing;

- "production" - code that has been released with a view to it being applied to its stated purposes and therefore typically required to meet defined documentation and testing standards; or

- "archive" - permanent copies of code at key milestones in its development that can aid study reproductions by making available the version of code used in a specific analysis.

Continuous integration [@CI2017] tools can help verify that each code update continues to pass a standardised battery of quality tests. Finally, using deprecation conventions that take an informative and staged approach to retiring old code and data reduces the risk of model revisions having unintended consequences on third party users.

We developed a toolkit to help us develop and use MOSHEM modules, datasets and analyses that meet the standards listed in Table \@ref(tab:timelygls). The toolkit is comprised of online repositories and software.

## Repositories <!-- Existing ref -->
We created a GitHub organisation (a collection of code repositories) where all our development code is stored and version controlled [@ready4gh2022]. We configured the repositories in our GitHub organisation to use GitHub actions to support continuous integration. Some of the continuous integration checks we have defined assess each library's compliance with policies specified by the Comprehensive R Archive Network (CRAN) [@CRAN2022], to which we plan submitting future production releases. To track our code coverage, we linked our GitHub organisation to an account we established at codecov [@codecov_2022].

We enabled GitHub Pages in each repository we used for code library development to facilitate the creation and hosting of library documentation websites. We also developed a consolidated and versioned project documentation website [@rfwn2022] using the Hugo framework [@hugo_2023] and Docsy theme [@docsy_2023]. To host that website we established an account with Netlify [@netlify_2023] and linked that service to our GitHub organisation so that the website would automatically update whenever the source code in its GitHub repository was edited.

We also created a Zenodo community [@ready4zen2022] - a collection of permanent, uniquely identified repositories. We then linked our Zenodo community and GitHub organisation so that every time we specify a version of code in one of our GitHub repositories as a "release", a copy of that code is automatically created on Zenodo with a DOI. Finally, to manage model datasets, we created a dedicated collection within the Harvard Dataverse installation [@ready4dv2022].

## Software <!-- Novel ref -->
We created six development version R libraries to help us author model modules, supply those modules with data and implement reproducible modelling analyses. The six libraries, their primary focus, the standards they support and the third-party packages they depend on are summarised in Table \@ref(tab:cpkgs).

A library called ready4 [@ready42022] defines a template module (using R's S4 class system) from which all model module data structures will inherit features and a novel syntax for attaching algorithms to those data structures. The ready4 library also contains tools for retrieving web based information on model modules, datasets and analysis programs and for partially automating updates to the project documentation website.

Three R libraries are designed to standardise and partially automate workflows for authoring new model modules. The ready4pack library [@ready4pack2022] is designed to integrate with our GitHub organisation and provides tools for authoring module libraries that are:

- documented (with a website, a manual itemising selected contents and a manual itemising all contents);

- licensed (using GNU GPL-3 [@GNUGPL2022] by default);

- easily citable (citation information can be retrieved within an R session or from hosting repositories); and

- quality assured (each update triggers continuous integration workflows, including any unit tests created by module library authors).

The ready4pack library depends on two other module authoring libraries. Methods from the ready4fun library [@ready4fun2022] are used to verify that functions for implementing module algorithms are written in a consistent house style. That standardised format is then used by ready4fun methods to automatically generate basic documentation for each function. Methods from the ready4class [@ready4class2022] library are used to streamline and standardise the authoring of module data structures and the linking of methods to these data structures. Like ready4fun, the ready4class library uses standardised code implementation to automatically generate basic documentation for each module data structure.

The ready4use library  [@ready4use2022] contains modules for ingesting model datasets from online repositories (hosted on a Dataverse installation or on GitHub), labelling model datasets and sharing model datasets via online repositories. The ready4show library [@ready4show2022] contains tools to help author analyses programs that are either self-documenting or which trigger the creation of a scientific summary.

When used in conjunction with toolkit repositories, the six R libraries provide support for implementing 17 out of 20 framework standards (Table \@ref(tab:timelygls)). Standards relating to safe dissemination of statistical models (R8), user-interface development (R9) and deprecation conventions (U4) are better met through using existing third party R libraries. Preparing statistical models for dissemination can be accomplished with standard R data management tools like the dplyr [@dplyr2022] and purrr [@dplyr2022] libraries. User-interfaces are typically developed with the shiny [@shiny2022] library, for which a tutorial aimed at health economists is available [@SmithR2020]. The library lifecycle [@lifecycle2021] provides tools for R developers to consistently deprecate their code.

We applied the framework to develop an initial set of ready4 modules, supply those modules with data and implement modelling analyses. These outputs were created as part of a previously described study [@Hamilton2021.07.07.21260129] to develop utility mapping models appropriate for use in samples of young people presenting to primary mental health services. The ready4 framework's modelling toolkit created the following artefacts:

 - development version module libraries for describing and validating youth mental health human record datasets [@hamilton_matthew_2022_6084467], scoring health utility [@hamilton_matthew_2022_6084824], specifying utility mapping models [@hamilton_matthew_2022_6116701] and implementing reproducible utility mapping studies [@gao_caroline_2022_6130155];

 - a development version library of functions for finding and using utility mapping models developed with these tools [@matthew_p_hamilton_2021_5646669];

 - data collections of synthetic populations for testing model modules [@DVN/HJXYKQ_2021] and study input and results data [@DVN/DKDIB0_2021];

 - programs for replicating all steps from data ingest to manuscript reporting [@hamilton_matthew_2022_6129906], applying utility mapping models to new data [@hamilton_matthew_2022_6416330] and generating a synthetic representation of the study dataset [@hamilton_matthew_p_2022_6321821];

 - subroutines for creating a catalogue of utility mapping models [@hamilton_matthew_2022_6116385] and generating a draft scientific manuscript [@matthew_p_hamilton_2022_5976988] for studies implemented with these modules.

We created a checklist (Table \@ref(tab:checktb)) that we used to assess the extent to which study outputs met framework standards. We assessed these outputs as wholly or mostly meeting 18 out of 20 standards. The two standards where the study outputs currently fall short are in reporting code coverage and including a user-interface. Both these items are scheduled to be addressed when we release production versions of the code libraries.

In this article we described a framework that we developed to help us implement ready4 - a MOSHEM in youth mental health. We outlined framework standards for an accountable, reusable and updatable MOSHEM and described the modelling toolkit we created for applying those standards to the development and use of the ready4 MOSHEM. We also provided an overview of an initial set of MOSHEM modules developed with the framework to implement a utility mapping study. We reviewed the modules, datasets and analyses generated by that study against framework standards. The work we have described has potential implications for the development of the ready4 MOSHEM and for health economic modelling in mental health. A number of issues have more general relevance to health economic modellers and funders of health economic research.

## Implications for implementing ready4
The most direct implication of the development of the ready4 framework is that it makes it feasible for us to implement a MOSHEM in youth mental health. The standards specified by the framework have enabled us to partially automate workflows for developing and applying ready4 through use of the framework's modelling toolkit. We have demonstrated the practical utility of the modelling toolkit by applying it to authoring, documenting and disseminating ready4 module libraries  [@hamilton_matthew_2022_6084467; @hamilton_matthew_2022_6084824; @hamilton_matthew_2022_6116701; @gao_caroline_2022_6130155], datasets [@DVN/HJXYKQ_2021; @DVN/DKDIB0_2021]; analyses [@hamilton_matthew_2022_6129906; @hamilton_matthew_2022_6416330; @hamilton_matthew_p_2022_6321821], reporting templates [@hamilton_matthew_2022_6116385; @matthew_p_hamilton_2022_5976988] and prediction tools  @matthew_p_hamilton_2021_5646669] used in a utility mapping study [@Hamilton2021.07.07.21260129]. The standardised and partially automated workflows used in creating and sharing these artefacts has the potential to generate significant efficiencies as we apply the ready4 framework to undertaking new economic studies.

We have also been able to demonstrate the interoperability of the initial ready4 modules developed with the modelling toolkit. The program used to implement the utility mapping analysis [@hamilton_matthew_2022_6129906] combines modules from four module libraries ([@hamilton_matthew_2022_6084467; @hamilton_matthew_2022_6084824; @hamilton_matthew_2022_6116701; @gao_caroline_2022_6130155]) and two framework libraries [@ready4use2022; @ready4show2022]. Example literate programs published on the ready4 documentation website [@rfwn2022] use toy data [@DVN/HJXYKQ_2021] to illustrate the potential for ready4 modules to facilitate study replication and transferability. As demonstrated by the checklist we developed (Table \@ref(tab:checktb)), our framework's standards also provide a mechanism for to assess the extent to which the ready4 MOSHEM meets explicit objectives.

However, having features that facilitate accountability, reuse and updating is not the same as being accountable, reused and updated. If diverse groups of stakeholders do not review model components, suggest improvements and develop alternatives, then little progress is made towards enhancing model legitimacy. Similarly, making code and data publicly available does not guarantee that others will know of the existence of these tools, trust their validity and find them easy to use. Without reuse, errors in model artefacts are more likely to remain undetected. Even when errors are detected, they still need to be fixed, but maintaining code and data requires ongoing resourcing through a combination of centralised infrastructure and an active open source community.

To progress from a technical capability to behavioural outcomes, both our framework and MOSHEM need further work. Currently all the framework and model module libraries we have developed are available only as "development" releases. An early priority for us is to undertake the additional development, testing and documenting of these libraries so that we can submit production versions of each library to CRAN [@CRAN2022]. Making an R library available on CRAN is normally a prerequisite for a high level of use.

The transferability claims we make for our existing modules are to date supported only by example programs using toy data. Our future work aims to address this with real world studies that apply modules to different concepts and contexts. Our current work program also aims to create new ready4 modules for modelling help-seeking choice, spatial epidemiology, household populations and primary mental health services that we hope will provide others with more reasons to use ready4 and contribute to its development. To facilitate code contributions by third parties, our libraries for authoring modules [@ready4pack2022: @ready4fun2022; @ready4use2022] require some additional development to make them easier to use by third parties without knowledge of the naming and directory structure conventions we use in authoring code.

## Implications for economic modelling in mental health
Open source approaches have been recommended to help develop the mental health modelling field [@RN73] but only one mental health related model (in Alcohol Use Disorder [@Basu2018]) is currently indexed in the Open Source Models Clearinghouse [@OSMC_20xx;@Emerson2019]. We are aware of just one other open source mental health model - a reference model in Major Depressive Disorder - that is currently in development [@IVIMDD2022]. Of the known barriers to adoption of open source models by health economists (including issues like intellectual property and confidentiality [@Pouwels2022; @Wu2019]), our experience suggests that the biggest challenges may be the enormous effort required to first prepare model code and data for public release in formats that facilitate appropriate reuse by third parties and to then maintain and continually improve potentially large numbers of digital artefacts.

Automated tools such as those we developed in our modelling toolkit can help reduce the burden associated with some of these tasks. However, we think the current low rates of adoption by health economists of open source approaches [@Emerson2019; @Michalczyk2018; @Feenstra2022] (which in turn facilitate the collaboration that make modular models more attractive) will only change slowly unless there is significant and strategic investments made by research funders. Currently, incentive structures for health economists do not promote the dedication of large quantities of time to enable peers to reuse their work.<!-- MH Research priorities (WP3 - Research, WP6 - Economic) https://doi.org/10.1016/S2215-0366(15)00332-6 -->

Reducing waste in research is a responsibility of research funders [@chalmers2014increase] and the poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017], limited reusability [@Emerson2019; @Michalczyk2018; @Feenstra2022] and uncertain validity [@Ghabri2019; @Radeva2020] of health economic models is wasteful. Approximately 4,000 mental health focused economic evaluation reports were produced between 2000 to 2019 [@RN34]. The intellectual asset represented by this literature could be enhanced if many of the models described in these reports could be brought and kept up to date and made available in formats that maximised transferability to diverse decision contexts. We believe that modular and open source approaches would be well suited to accomplishing this goal and that the framework we have developed could act as an early prototype for solving some of the technical challenges of this task. Ideally such a program of research would be resourced to be sustained over the medium to long term and to engage a diverse network of investigators, contributors and advisers from high, middle and low income countries.

In addition to extracting more value from the existing health economic knowledge base in mental health, there is an opportunity for research funders to shape how future health economic models in mental health are undertaken. Mental health topics accounted for 268 of the 2829 (10%) peer reviewed economic evaluations undertaken during 26 month period in 2012-2014 identified by a review [@pitt2016economic]. The ongoing annual output of economic research in mental health that focuses on the other 11 domains identified by Wagstaff and Culyer [@wagstaff2012four] is probably also substantial. Funders should provide support for the projects and infrastructure to promote greater collaboration, interoperability, transferability and maintenance of future mental health modelling projects.

Developing networks of modellers working on common health conditions has been recommended as a strategy for improving model validity [@Sampson2019] and some of us are part of a nascent initiate of this type in mental health [@whiteford_bagheri_2022]. Collaboration between teams of health economists can make some complex modelling projects more feasible [@Arnold2010] and the significant deficits in our understanding of the systems in which mental disorders emerge and are treated [@Fried2020] suggest that there are a number of candidate topics in mental health that might benefit from pooling of efforts. The weak theoretical underpinnings for understanding complex mental health systems [@RN2111] may be a place to start. It remains unclear why increased investments in mental health care have yet to discernibly reduce the prevalence and burden of mental disorders[@RN26].  The  literature, and evidence base, regarding how the requirements, characteristics and performance of mental health services are shaped by spatiotemporal context needs to be further developed [@RN42]. There is also a need for better evidence to identify the social determinants of mental disorders most amenable to preventative interventions, and for which population sub-groups such interventions would be most effective [@RN43].

Ideally health economists would explore these complex topics in partnership with modellers from other disciplines (in particular epidemiology and health services research) and a wide range of stakeholders such as other researchers, policymakers, service planners and community members. Modular and open source approaches would facilitate such investigations by breaking down ambitious and long term goals into manageable time-bound discrete projects, each progressed by different teams. To facilitate such an approach, a common framework of standards and tools would be needed. To be suitable for such a task our framework would need additional development, with its overall architecture reviewed for scalability and suitability and to provide better integration with and use of other open source languages (particularly python) and repositories. Whatever MOSHEM infrastructure is developed, its resilience would depend on a community of open source contributors sufficiently large and active to ensure that all core modules are maintained even after their original authors cease their involvement.

## General issues for health economists and health research funders
Some of the issues we have discussed in the context of the development of our model or health economic modelling in mental health are potentially relevant to health economists and funders of health economic research more generally. Proactive measures by funders to encourage more accountable, reusable and updatable health economic models is not a need confined to mental health. For example, funders have been encouraged to support methodological innovation to improve model transferability [@craig2018taking]. However, funders also need credible proposals to support and this is an area for health economists interested in MOSHEMs to prioritize. Health economists could use existing and new special interest groups to identify opportunities and enablers of more collaborative approaches to model development, potentially as the basis for future funding proposals.

Adopting MOSHEMs will expand the type of skillset typically engaged in health economic modelling projects, with a much greater role for data-scientists, software engineers and online community builders. The requirement for these roles should be incorporated into project proposals. Not all efforts by health economists to promote MOSHEMs need to depend on the decisions of research funders. Releasing selected subsets of unmaintained model artefacts in open source repositories is still better than not providing access to any code and data and can typically be accomplished within existing project budgets. Developing knowledge and skills of MOSHEMS can be advanced by making small contributions (e.g. improvements to documentation, code contributions) to open source projects. Our project website [@rfwn2022] includes details of multiple ways to contribute to ready4.

<!-- Estimating the potential uptake of these tools is difficult.  -->
<!-- Open source is more than open access (limited discussion). -->
 <!-- A coding framework for OSHEMs developed in the language R includes standardised approaches to directory structure and naming conventions [@Alarid2019]. -->
<!-- http://hdl.handle.net/1893/27205--><!-- CONTEXT: Funders should also support methodological work on context, including the development of reporting templates to facilitate accurate, comprehensive and systematic reporting of context, approaches to evidence synthesis that can capture contextual variation adequately and modelling methods that can leverage the accumulating evidence on contextual variation to predict the reach, implementation and effectiveness of promising interventions in new contexts." --

<!-- (described in more detail on the project documentation website [@rfwn2022])  -->
<!-- Sustained funding -->

 <!-- - One shot at big changes -->
 <!-- - Our analysis - do it in stages, be transparent, continually improve, get help, be flexible (modelling toolkit / monolithic model), have a framework -->
<!-- Diversify skills - move to tools means software engineering, -->
<!-- More collaboration -->
<!-- Role for SIGs -->
<!-- Make grant proposals that give funders the opportunity to fund this work. -->
<!-- Share some code and data at small scale. -->

<!-- Standardized and transferable approaches to health economic modelling have been recommended. [doi:10.1017/S0266462317000666] -->
<!-- As modelling methods have become more sophisticated, some have argued that more complex models are required to adequately explore certain types of public health topics [@SQUIRES2016588].  -->
<!-- Modellers should be faithful to reaity and values of users DOI: 10.1002/asi.23697 -->
<!-- transferability - [https://doi.org/10.1017/S026646232200321X].  -->
<!-- Output by health economists [@pitt2016economic; @wagstaff2012four; @munoz2020health] -->
<!-- - Can develop guidelines and funding proposals, make small amounts of artefacts online -->

<!-- Reducing waste in research is responsibility of research funder: https://doi.org/10.1016/S0140-6736(13)62229-1 -->

<!-- greater use of these types of models may require adaptation on the part of funders, modellers and decision-makers. T -->

<!-- Link to Wagstaff - Demand forecasting, valuation of health and healthcare, economic evaluation, utility mapping?? -->

<!-- # Motivation -->

<!-- ## Why develop OSHEMs in mental health -->

<!-- Major mental health reform programs, <!-- such as those currently being implemented in Victoria, Australia [@VicRC2021],  -->
<!-- can involve the identification, prioritisation, sequencing, targeting and monitoring of multiple interdependent initiatives.   -->

<!-- The development, validation and updating of more complex mental health economic models implemented over longer time frames may be too onerous a burden for a single modelling team.  -->

<!-- Decision makers are frequently left to rely on economic analyses that become less valid and relevant each year. A model can grow stale when changes in the real world, whether sudden (e.g. new pandemics) or gradual (e.g. demographic), is not represented by corresponding updates to relevant model features. Even when the system being modelled remains relatively stable, new research findings may potentially invalidate aspects of a model's original conceptualisation and implementation.  -->

<!-- that rarely explores complex dynamic systems  -->
<!-- [PECUNIA] -->

<!-- Computational modelling could play an important role in developing policies to improve population mental health but this may require significant changes in the way mental health modelling projects are funded, conceptualised and implemented.  -->
 <!-- but, as with health economics more generally, OSHEMs remain rare. Greater use of open source approaches could help improve the scope, validity and usefulness of mental health economic models. -->

<!-- Repositories such as Zenodo [@Zenodo2013] and Dataverse [@Dataverse2007] provide persistent storage solutions that generate a Digital Object Identifier (DOI) for each unique item. These repositories are a preferable solution for sharing citable code and data than transitory repositories such as corporate websites or GitHub where items can be deleted or relocated at any time [@sseditors2022]. Zenodo includes tools that automate integration with GitHub, which makes it easy for developers to maintain parallel code repositories - one for disseminating the most up to date development code and the other for archiving citable code releases. --> <!-- Developers storing data in a Dataverse installation have access to multiple meta-data fields to document both a data collection and its individual constituent files.  --><!-- In R, code manuals and websites can be created with the aid of tools such as devtools [@devtools2021], sinew [@sinew2022], roxygen2 [@roxygen2021] and pkgdown [@pkgdown2022].  -->

<!-- and the use of version control systems. <!-- such as Git [@git20XX].  -->

<!-- Applying a previously published algorithm [@kross2016] to analyse the most comprehensive archive of released R packages [@CRAN2022] finds that `r gpl_pc_dbl`% are distributed under various forms of General Public License (GPL) [@GNUGPL2022], a copyleft license. -->
<!-- In R, including a CITATION file in the `inst` directory of a package will enable users of that package to retrieve citation information by running a command of the format `citation("Package Name")` in the R console [@Salmon2021]. More generally, including a CITATION.cff file at the top level of your code repository will enable GitHub and Zenodo repositories hosting that item to include the relevant information in their citation tools [@Druskat_Citation_File_Format_2021]. Datasets hosted on Dataverse installations have metadata fields that, once completed by authors, generate citation files for dataset viewers. -->

<!-- A key challenge to generalising health economic models is that they are typically developed to inform a decision problem with a highly specific jurisdictional context. However, a number of choices about how these models are implemented can significantly increase the re-usability of model code in other contexts.  -->
<!-- For example, when generalising a model developed for the Australian context to a UK context, one could create a class that initially inherits all of the methods defined for the Australian model and then write new or replacement methods as needed for the UK model.  -->

<!-- A number of tools and approaches can make the process of implementing and curating changes to model code and data more coherent and efficient.   <!-- Repositories such as Zenodo [@Zenodo2013] and Dataverse [@Dataverse2007] provide persistent access to all published versions of a dataset, each uniquely identifiable. For code, use of version control tools like Git [@git20XX] can ensure that the entire development history of a project is organised so that each version is distinguishable and retrievable by developers. The online platform GitHub [@github2007] can make this version history accessible to anyone. -->

<!-- For R code, the usethis [@usethis2021] package can be used to partially automate version number increments using the convention Major.Minor.Patch.Development (e.g. ExampleSoftware v1.1.1.1). Datasets stored on the Harvard Dataverse use the simpler Major.Minor convention.  -->
<!-- MOSHEMs developed in R can take advantage of templates provided by devtools [@devtools2021] and pkgdown [@pkgdown2022] to run continuous integration checks on GitHub.  --><!-- These tests can include those of units (do individual functions produce expected output?), documentation (does documentation render correctly?; can all example workflows be executed?) and installation (can the software be successfully deployed on multiple types of operating system?).  -->
<!-- MH systems design is not a pharma led project - less concerns about commercial ownership -->

<!-- and for mental health specifically [@Occhipinti2021],  -->
<!-- could provide greater insight about inter-dependencies between candidate policies and the dynamic nature of the mental health systems planning context. Dynamic systems methods might provide the foundations for developing   -->
<!-- and models designed for multiple purposes r -->

 <!-- as part of the Open Value Initiative [@Jansen2019].  -->
<!-- Similarly, developing partnerships between modellers and decision-makers across the life-cycle of a modelling project can help ensure models are appropriately conceptualised, implemented and have practical utility as decision aids [@Zabell2021; @SQUIRES2016588]. -->
 <!-- Yet single purpose models that assume static systems may be inadequate for the decision support needs of policymakers and service planners [@PC2020; @Occhipinti2021].  -->
<!-- The large and widespread additional mental health burdens recently observed during the COVID-19 pandemic[@20211700] and predicted as a potential future consequence of climate change [@page_howard_2010], highlight the need to improve the resilience and adaptability of these systems. To help stem growing demand for mental health services, policymakers have also been encouraged to place greater emphasis on tackling the social determinants of mental disorders [@RN11].  -->

<!-- Our approach to model development is to undertake a number of discrete modelling projects of the people, places, platforms and programs that shape the mental health and wellbeing of young people and to progressively link them together by means of a common framework. To model people we are developing synthetic representations of populations of interest [@DVN/HJXYKQ_2021] that describe relevant individual characteristics and their relationships, algorithms that map psychological measures to health utility [@Hamilton2021.07.07.21260129] and choice models for predicting the helpseeking behaviour of young people [@hamilton_matthew_2022_6627995]. Our in-development model of places [@DVN/V3OKZV_2022] has the aim of synthesising geometry and spatial attribute data to characterise the geographic distribution of relevant demographic, environmental, epidemiological and service infrastructure features. We are in the early stages of a multi-annual project to develop a service platform model that will represent the processes and operations of a complex primary youth mental health service. We also plan to extend and update our prior work reviewing economic evidence relating to youth mental health programs [@RN33] so that it can be integrated with this model. Our initial work on *readyforwhatsnext* is focused on Victoria, Australia but the framework we are using to develop it is designed to facilitate extension by ourselves and others to different jurisdictional decision contexts. Progress is reported on a project website [@rfwn2022]. -->

### 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
first draft of reworked ethics paper

---
title: "MS_Aims"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---

In this paper, we <!--  MOVED_EDITED_START -->consider literature on modelling practice and draw on our prior experience with developing and using CHEMs to propose<!--  MOVED_EDITED_START -->:  (i) a set of ethical responsibilities for CHEM project teams<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.  <!--  MOVED_EDITED_END -->  
<!-- (i) some core ethical responsibilities of CHEM project teams;

 <!-- -->

<!-- (ii) describe a prototype software framework we have developed to support the ethical implementation of CHEMs; and -->

<!-- (iii)	assess an initial set of youth mental health CHEMs developed with our software framework against our proposed ethical CHEM implementation criteria.  -->

### 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Body.Rmd
first draft of reworked ethics paper

---
title: "Manuscript Body"
description: "Main body of paper"
date: "`r Sys.Date()`"
output: html_document
---

```{r child="../Child_RMDs/XX_Funs.Rmd", echo=FALSE}
```
```{r child="../Child_RMDs/MS_Tables.Rmd", echo=FALSE}
```
```{r child="../Child_RMDs/MS_Figures.Rmd", echo=FALSE}
```
<!-- Add child RMDs for results content -->

# Introduction

```{r child="../Child_RMDs/MS_Intro_Short.Rmd", echo=FALSE}
```
```{r child="../Child_RMDs/MS_Aims.Rmd", echo=FALSE}
```

<!-- # Methods -->
```{r child="../Child_RMDs/MS_Method.Rmd", echo=FALSE}
```

<!-- # Results -->
```{r child="../Child_RMDs/MS_Results.Rmd", echo=FALSE}
```

# Discussion

```{r child="../Child_RMDs/MS_Discussion.Rmd", echo=FALSE}
```

# Conclusion

```{r child="../Child_RMDs/MS_Conclusion.Rmd", echo=FALSE}
```

<!-- ## Acknowledgement {-} -->
<!-- The authors would like to acknowledge the contribution of John Gillam who provided advisory input to this research. -->

<!-- ## Availability of data and materials {-#availability-of-data-and-materials} -->
<!-- # ```{r child="../Child_RMDs/MS_Available_Data_And_Resources.Rmd", echo=FALSE} -->
<!-- # ``` -->

## Ethics approval {-}
```{r child="../Child_RMDs/MS_Ethics.Rmd", echo=FALSE}
```

## Funding {-#funding}
```{r child="../Child_RMDs/MS_Funding.Rmd", echo=FALSE}
```

## Conflict of Interest {-#conflict-of-interest}
```{r child="../Child_RMDs/MS_COI.Rmd", echo=FALSE}
```

### 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Cndl_Tabs_Not_In_Body.Rmd
first draft of reworked ethics paper

---
title: "MS_Cndl_Child_Not_In_Body"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---

\newpage
<!-- `r if(!is.null(child_tabs_1L_chr)){"## Tables {-}"}` -->

<!-- and figures -->

```{r, child = '../Child_RMDs/MS_Tables_1.Rmd', echo=FALSE}
```

\newpage

```{r, child = '../Child_RMDs/MS_Tables_2.Rmd', echo=FALSE}
```

\newpage

<!-- # ```{r, child = '../Child_RMDs/MS_Tables_3.Rmd', echo=FALSE} -->
<!-- # ``` -->
<!-- # \newpage -->
<!-- # ```{r, child = '../Child_RMDs/MS_Tables_4.Rmd', echo=FALSE} -->
<!-- # ``` -->

### 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Conclusion.Rmd
first draft of reworked ethics paper

---
title: "MS_Conclusion"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---
`r ifelse(is.na(params$X@conclusion_1L_chr),"",params$X@conclusion_1L_chr)` 
We have identified criteria that can be used to systematically assess the extent to which the computational implementation of health economic models adheres to transparency, reusability and updatability criteria. <!-- We have developed a prototype software framework that can support the implementation of CHEMs that largely meet these criteria. Our framework can be used as a prototype for developing future software frameworks to support ethical implementation of CHEMs.  --> Barriers to ethical CHEM practice arising from resource constraints should be addressed by research funders.

### 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Discussion.Rmd
first draft of reworked ethics paper

---
title: "MS_Discussion"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---

Ethical practice is a core expectation of health researchers and computational methods underpin most quantitative research, yet an understanding of what constitutes ethical public health modelling practice is underdeveloped [@10.3389/fpubh.2017.00068]. We have addressed this gap with our proposed responsibilities for model developers and funders, enabling CHEM attributes and CHEM implementation assessment criteria.

Currently, many if not most existing CHEMs are insufficiently transparent [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], not reusable [@Feenstra2022; @Emerson2019], nor updatable [@Sampson2017; @10.3389/fpubh.2022.899874]. Addressing these issues will require changes to the tools, practices and funding underpinning health economic model development.

<!-- As illustrated by Table \@ref(tab:checktb), we have developed a prototype software framework that can be used to implement youth mental health CHEMs that largely satisfy our TRU criteria. However, in its current form our software framework is too fragile to be recommended as anything more than a prototype for supporting the development needs of other modelling teams and projects. We developed our prototype software framework primarily to address the user requirements of one group – ourselves. Factors such as user enjoyment, usability, active user-community and supporting resources influence adoption of software frameworks [@myllarniemi2018development]. We currently lack the resources required to  address these issues for our software framework. -->

<!-- However, our prototype has a number of features that future work on CHEM software frameworks may find useful to incorporate. Firstly, developing a software framework to work within an existing and widely used open source programming language such as R, can keep framework scope relatively narrow. This makes it more tractable to develop, maintain and learn, and facilitates integration with other modelling tools written in that language (e.g., the dependency libraries we list in Table \@ref(tab:cpkgs)). Secondly, implementation that combines both object-oriented and functional programming paradigms can avail of the modular and syntactical simplicity benefits of the former, while limiting needless bundling of code artefacts. Thirdly, a sensible trade-off needs to be found between transparent code implementation (which requires clear and sufficiently detailed documentation) and Agile Software Development (for which a foundational principle is prioritising the development of working code over writing documentation [@beck2001manifesto]). Our software framework makes this trade off by enforcing the use of consistent code style conventions and file organisation which in turn enables automated generation of simple documentation at every CHEM module update. All model data-structures and algorithms are therefore always documented (at least minimally, with machine authored content), meaning model developers have a requirement to write customised documentation less frequently. Many of the features of our prototype software framework have the potential to be significantly enhanced through integration with evolving code and documentation authoring capabilities of large language models. -->

<!-- Our software framework provides the outline of a technical solution to some of the challenges of open source CHEM development. However, better tools alone are insufficient and policy and institutional changes are also required.  -->
Reducing waste in research is a core responsibility of research funders [@chalmers2014increase] and funding the development of CHEMs that are not adequately understood, reused or updated is wasteful. Existing incentive structures for health economists generally do not promote facilitating peers to reuse their work and funding is rarely adequate to maintain and routinely update a model. Currently, it can take “an extraordinary amount of idealism” to commit to authoring and maintaining research software [@anzt2020environment]. Previously recommended strategies for more beneficial health economic research investments include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019]. Development of software frameworks to support ethical CHEM implementations could enable and enhance each of these strategies.

<!-- A future software framework for ethical CHEMs would ideally incorporate a base set of features useful to developers of computational models across all domains of public health and health economics, with the capability for community-led extensions that are tailored to the needs of modellers focused on specific health-conditions. Such approaches will likely be important to ensure a sufficient potential user-base to make CHEM software frameworks sustainable. -->

### 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Imp_Example.Rmd
first draft of reworked ethics paper

---
title: "MS_Imp_Guidelines"
output: html_document
date: '2022-07-21'
---

We are using the ready4 software framework to implement multiple transparent, reusable and updatable CHEMs in youth mental health (see Figure \@ref(fig:fig1)).  Some initial outputs from this work are publicly available as development version releases.

```{r, child = child_figs_ls$fig1, echo=FALSE}
```

##	Economic topics
Currently, we are using the ready4 software framework to develop, apply, and share youth mental health CHEMs in four of the twelve domains of health economics identified by Wagstaff and Culyer [@wagstaff2012four]:

-	health and its value (our projects: models to map psychological and functional measures to health utility);

-	determinants of health and ill-health (our projects: models for creating synthetic household populations with key risk and protective factors for mental disorders);

-	demand for health and health care (our projects: spatial epidemiology and help-seeking choice models); and

-	supply of health services (our projects: a model of primary mental health care services for young people).

Potential future directions are to supplement this work with CHEMs in two additional Wagstaff and Culyer domains of: (i) public health (to model the impact of selected fiscal policy and regulation options on young people's mental health); and (ii) human resources (to model the supply and behaviours of the youth mental health workforce). Our ultimate aim is to flexibly combining all our CHEMs in analyses that help answer questions in two additional Wagstaff and Culyer domains:

-	efficiency and equity (our goal: assess the distributional impacts and identify the optimal targeting of youth mental health interventions); and

-	economic evaluation (our goal: assess the cost-utility of competing policy options for improving the mental health of young people).

Although we are principally interested in using CHEMs to answer policy questions relating to the mental health of young people in Australia, we want to facilitate CHEM transferability to other jurisdictions. Our CHEMs are being derived from and applied to real data (which can be empirical, simulated or assumption, so long as it is appropriate for use in analysis intended to inform decision-making) from Australia. To help demonstrate the potential use of CHEMs in other decision contexts, we also create toy datasets. Data created for illustration purposes is prominently labelled as not for use in decision-analysis.

## Case study: health and its value
We have previously described a study [@Hamilton2021.07.07.21260129] to develop utility mapping models for use in samples of young people presenting to primary mental health services. The ready4 software framework was used in that study to develop CHEM modules, supply those modules with data and implement modelling analyses, creating the following artefacts:

 - development version module libraries for describing and validating youth mental health human record datasets [@hamilton_matthew_2022_6084467], scoring health utility [@hamilton_matthew_2022_6084824], specifying utility mapping models [@hamilton_matthew_2022_6116701] and implementing reproducible utility mapping studies [@gao_caroline_2022_6130155];

 - a development version library of functions for finding and using utility mapping models developed with these tools [@matthew_p_hamilton_2021_5646669];

 - collections of real data (study input and results [@DVN/DKDIB0_2021]) and toy data (synthetic populations for testing model modules [@DVN/HJXYKQ_2021]);

 - programs for replicating all steps from data ingest to manuscript production [@hamilton_matthew_2022_6129906], applying utility mapping models to new data [@hamilton_matthew_2022_6416330] and generating a synthetic representation of the study dataset [@hamilton_matthew_p_2022_6321821]; and

 - subroutines for creating a catalogue of utility mapping models [@hamilton_matthew_2022_6116385] and generating a draft scientific manuscript [@matthew_p_hamilton_2022_5976988] for studies implemented with these modules.

We created a checklist (Table \@ref(tab:checktb)) that we used to subjectively assess these study outputs against TRU criteria. For each criterion, we provided a global assessment of whether it was met using the responses “yes”, “no” or “partial”.  We believe our utility mapping study CHEMs have satisfactorily met five of the six criteria (T1, T2, R1, R2 and U1) and have partially met one criterion (U2). The main shortcomings that we identified when applying the assessment criteria was that we have yet to adequately implement unit testing of the R libraries authored as part of this study.

\blandscape

```{r, child = child_docs_ls$four, echo=FALSE}
```

\elandscape

### 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Imp_Guidelines.Rmd
first draft of reworked ethics paper

---
title: "MS_Imp_Guidelines"
output: html_document
date: '2022-07-21'
---

<!-- We considered literature on modelling practice and reflected on our prior experience with developing and using CHEMs to identify: (i) some core ethical responsibilities of CHEM project teams; (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.  -->
<!-- # -->
# Ethical responsibilities of CHEM project teams
Boden and McKendrick [@10.3389/fpubh.2017.00068] propose a framework for ethical public health modelling based on the criteria of independence (concerning how modeller subjectivity shapes model design), beneficence / non-maleficence (concerning model quality and utility), transparency (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and justice (concerning the social obligations of modellers to consider and communicate ethical issues about model use). However, these criteria are not specific to the computational model, but are designed to apply to a modelling project as a whole, including a model's conceptual and mathematical representations. The authors propose 13 questions to evaluate ethical risk across the four criteria, but of these, only "is the model code open source or available on request" (for transparency) and "has the model been verified, i.e., does it do what the modeller wants it to do?" (for beneficence / non-maleficence) are specific to model's representation as software.

To identify ethical issues that are specific to CHEMs, we first considered the distinct phases in a software development project. Software development lifecycle (SDLC) models take different approaches to the inclusion, naming, definition, sequencing and iteration of project phases [@ruparelia2010software]. However, SDLC models typically have components that can principally map to the concepts of *planning* (e.g., identification of system / user requirements and design specifications), *implementation* (e.g., development and testing) and *release* (e.g., deployment, system integration, maintenance and support). For simplicity, we propose one overarching ethical responsibility for CHEM project teams for each phase. These responsibilities are to:

(i) *establish socially acceptable user requirements and design specifications* (during CHEM planning);

(ii) *ensure fitness for purpose* (of CHEM implementation); and

(iii) *support socially beneficial use* (on CHEM release).

Transparency underpins all three responsibilities (Table \@ref(tab:timelygls)). Boden and McKendrick’s other ethical criteria can be selectively mapped to each phase's responsibility: justice and independence to the planning responsibility; independence to the implementation responsibility; and beneficence / non-maleficence and justice to the release responsibility. All three ethical responsibilities have implications for both professional practice and project resourcing. As it is unreasonable to place expectations of modellers that they are not resourced to fulfil, these responsibilities should apply jointly to health economic model developers and funders.

Misalignment between the values of computational model developers and those of the population groups affected by decisions based on their models presents significant ethical risks [@thompson2022escape; @duckett2022journey].  The value judgments of health economic modellers are rarely adequately specified, omissions that may lead to socially unacceptable policy recommendations [@duckett2022journey]. These value judgments influence the assumptions, selection of model features and standards for evidence that shape health economic model projects [@HARVARD2020112975]. For example, as illustrated in Table \@ref(tab:proscons), project resource constraints may require model developers to make trade-offs when agreeing CHEM *user requirements and design specifications*. These trade-offs in turn determine how straightforward it is for a CHEM to be subsequently amended to reflect alternative value judgments relating to model conceptualisation and mathematical formalisation (Table \@ref(tab:timelygls)).

Modellers have duties both to ensure a computational model is *fit for purpose* and to provide potential third-party users with the means of assessing its adequacy for their proposed uses [@Erdemir2020; @Feenstra2022; @thompson2019escape; @10.3389/fpubh.2017.00068]. However, it is common for health economic models to have serious methodological flaws [@carletto_zanuzzi_sammarco_russo_2020; @WONDER2015467]; insufficient validation [@Ghabri2019; @kolovos2017model; @haji2013model], poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017]; and undeclared errors [@Radeva2020]. Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and facilitating manual reviews by third parties.

A CHEM will have limited *social benefit* if it is rarely used, if mis-used or when its acceptability and adequacy rapidly decay. Reuse of CHEMs as components of other models can potentially make model development more efficient [@Arnold2010; @garcia2021cost]. However, health economic models face challenges related to transferability across jurisdictions [@garcia2021cost] that create barriers to reuse. Without ongoing maintenance, a CHEM risks becoming less reliable with time [@garcia2021cost] and is at risk of being deployed for purposes for which it is poorly suited [@calder2018computational]. Currently, health economic models are rarely implemented computationally in a manner that facilitates routine updates [@Sampson2017], thus limiting the temporal window within which a CHEM can be validly applied.

<!-- # -->
# CHEM attributes associated with ethical modelling practice
The responsibilities of model project teams over a CHEM lifecycle are easier to state than to measure.  However, aspects of ethical modelling practice may be inferred from measurable attributes of CHEMs. As described in Table \@ref(tab:timelygls),  we believe that the creators of transparent, reusable and updatable (TRU) CHEMs are likely to have fulfilled a number of their ethical obligations.  We therefore selected these model attributes to use as the basis for deriving ethical assessment criteria.

```{r, child = child_docs_ls$two, echo=FALSE}
```

Taking steps to make computational models *transparent* is an important part of ethical public health modelling practice [@10.3389/fpubh.2017.00068]. Guidance on transparency in health economic modelling recommend that model code and data should be clearly documented [@Eddy2012]. Notably, the same guidelines, published over ten years ago, do not recommend sharing model code and data. However, more recent healthcare modelling guidance does recommend public dissemination of such artefacts [@Erdemir2020]. Online repository services such as Zenodo [@Zenodo2013] and Dataverse [@Dataverse2007] provide persistent storage solutions that generate a Digital Object Identifier (DOI) for each code and data collection. Ensuring that calculations are correct and consistent with model specifications is an essential part of CHEM quality assurance [@techver2019]. The extensiveness of such verification checks can be reported using the concept of code coverage [@ERICWONG2010188] - the proportion of model code that has been explicitly tested. Tests should ideally combine both unit tests (to verify that small, isolated sections of code produce the correct output when run independently) and acceptance tests (to verify that the correct output is produced when multiple code components are run together to perform tasks that meet core user-requirements [@martin2003agile]). The nature and extent of individual model authorship contributions can become unclear when models are implemented over longer time-frames with a large and changing group of collaborators [@thompson2022escape]. This issue can be addressed by use of online repository services such as GitHub [@github2007], that provide citation tools and can transparently record all individual code contributions to a modelling project over its lifecycle.

A CHEM that is *reusable* also signals ethical modelling practice. Although making a CHEM’s code, data and documentation publicly available is increasingly considered good practice, it is insufficient to enable model re-use.  Key concerns for health economists when considering whether to reuse a model are generalisability (application without adaptation) and transferability (selective reuse and/or modification of model components) [@RN39]. Writing model algorithms as collections of functions (short, self-contained and reusable software routines that each perform a discrete task) is good scientific computing practice [@Wilson_2017] and promotes selective reuse. Computational implementations that store model code and data in distinct files and locations (as opposed to embedding data such as parameter values into source code) are easier to selectively modify. Modular implementations enable models to be constructed from multiple independently reusable and replaceable sub-models (modules) [@pan2021modular]. The programming concept inheritance [@8717448], deployed in object-oriented programming approaches, can facilitate the duplication and selective modification of models. Granting permissions to others to use, test and adapt models and their components, can be facilitated by two broad categories of open source licenses. Some guidance strongly recommends the use of permissive licensing [@Wilson_2017] that provides users with great flexibility as to the purposes (including commercial) for which content can be re-used. Alternatively, copyleft licenses [@copyleft2022] can be used to require that derivative works created by content users remain open-source.

Models should be *updatable* so that they remain valid for longer, evolving as new evidence emerges and the systems being modelled change [@Jenkins2021; @garcia2021cost]. Ensuring that a model is regularly reviewed to identify and implement required improvements is a recommended defence against model validity decay [@calder2018computational]. Sustainable maintenance of open source research software requires both a core development team and an active user community [@info:doi/10.2196/20028]. Online communities can be an efficient means of engaging model users in testing each version of a model, identifying issues and suggesting improvements. Services such as GitHub [@github2007] provide collaborative code development tools [@MERGEL2015464] that help elicit, integrate and reconcile contributions from multiple contributors and to ensure each update is uniquely identifiable and retrievable. It is also important that verification checks are rerun with each model update, a task that can be automated using the software development practice of continuous integration [@CI2017].  The risk of software revisions having unintended consequences for third party users can be mitigated through the use of deprecation conventions [@zhou2016api] that take an informative and staged approach to retiring outdated CHEMs.

<!-- # -->
#	Criteria for assessing ethical computational implementations
For each CHEM attribute that suggests ethical CHEM modelling practice, we identified assessment criteria.

How *transparent* a CHEM is can be assessed against the criteria:

-	*T1*: Software files (executable code, testing procedures and outcomes and non-confidential data) are available in open access repositories.

-	*T2*: It is easy to see who funded, developed and tested each part of the CHEM and to identify the modelling team’s assumptions, judgments and theories about CHEM development and use.

How *reusable* a CHEM is can be assessed against the criteria:

-	*R1*: Programming practices facilitate both generalisability and transferability.

-	*R2*: Licenses restrict only unethical use and allow anyone to ethically reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works.

How *updatable* a CHEM is can be assessed against the criteria:

-	*U1*: Maintenance infrastructure is in place to support version control and collaboration with model users.

-	*U2*: Each new release of a CHEM is systematically retested, with changes implemented to minimise disruptions for existing model users.

### 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Imp_Toolkit.Rmd
first draft of reworked ethics paper

---
title: "MS_Imp_Guidelines"
output: html_document
date: '2022-07-21'
---

To support the development of CHEMs that meet TRU assessment criteria, we have created a prototype software framework called ready4. The software framework aims to provide a toolkit for: (i) enabling modular implementation of CHEMs; (ii) authoring and documenting CHEM modules; (iii) managing the labelling and transfer of CHEM input and output data; and (iv) authoring reproducible analyses that apply CHEMs to compatible data.

To achieve these goals, we have implemented the software framework as R [@RCORE2022] code libraries that integrate with a number of online services and which are supported by a documentation website.

## R libraries 
A library in the R language will typically depend on multiple other R libraries, all of which potentially have different authors. As the number of third-party dependencies of an R library grows so does the fragility of that library (e.g., the library may cease to work as intended due to changes in one of its dependency libraries). To reduce the fragility of our framework we implemented it as multiple R libraries rather than one R library. In total we authored six novel R libraries to implement the ready4 framework, all of which have distinct purposes and dependencies (Table \@ref(tab:cpkgs)).

One framework library provides a *foundation for modular CHEM implementations*. In modular model implementations, modules need to be able to share inputs and outputs with each other and to be run as independent models [@barros2023empowering]. To achieve this, the foundation framework library defines a template CHEM module (using R’s S4 class system), which can be used to create other CHEM modules with a common set of inherited properties. One of these inherited properties is a novel syntax of 15 core commands that enable CHEM module algorithms to be consistently named. The foundation library also contains tools for retrieving web-based information on CHEM modules, datasets and analysis programs authored with the framework and tools for partially automating updates to a project documentation website.

Three framework libraries are designed to help streamline and standardise workflows for *authoring CHEM modules* from the template module. The R language supports functional and object-oriented programming paradigms [@10.1214/13-STS452]. Authoring with each of these paradigms is facilitated by a dedicated CHEM module authoring library. One module authoring library contains tools for writing functions in a consistent house style and automatically generating basic documentation for each function. A second module authoring library contains tools to help streamline and standardise the authoring and documenting of novel CHEM modules. A third module authoring library provides tools for disseminating themed bundles of CHEM modules as R libraries that are:

-	documented (with a website and PDF manuals);

-	licensed (using the copyleft GNU GPL-3 [66] by default);

-	easily citable (citation information can be retrieved within an R session or from hosting repositories); and

-	quality assured (each update triggers continuous integration workflows, including any acceptance and unit tests created by module library authors).

A library for *managing CHEM data* contains tools for supplying CHEM modules with input data ingested from local (i.e., a user's computer) or remote (online repositories) locations, labelling CHEM module datasets and exporting CHEM module data to online repositories. A library for *authoring reproducible analyses* contains tools to help write programs that apply CHEM modules to compatible datasets for the purpose of undertaking health economic analyses. These analysis programs can be either self-documenting (code is integrated with plain English explanations of what it does) or trigger the creation of explanatory documents (e.g., a scientific manuscript).

```{r, child = child_docs_ls$three, echo=FALSE}
```

##	Online services 
Framework libraries are designed to be used in conjunction with a number of third-party online services that we established and configured accounts with.

We created a GitHub organisation (a collection of code repositories) where source code that we author is stored and version controlled. We configured individual repositories in our GitHub organisation to implement continuous integration. By default, code libraries authored with our framework will use continuous integration to assess compliance with policies specified by the Comprehensive R Archive Network (CRAN) [@CRAN2022]. To track our code coverage, we linked our GitHub organisation to an account we established at codecov [@codecov_2022]. To facilitate the creation and hosting of documentation websites, we enabled GitHub Pages in each repository used for code library development.

We also created a Zenodo community - a collection of permanent, uniquely identified repositories. We then linked our Zenodo community and GitHub organisation so that every time we specify a version of code in one of our GitHub repositories as a “release”, a copy of that code is automatically created on Zenodo with a DOI. Finally, to manage model datasets, we created a dedicated collection within the Harvard Dataverse installation.

##	Documentation website
We developed a framework documentation website (https://www.ready4-dev.com) that provides guidance to model developers on how to use and contribute improvements to the ready4 software framework and CHEM modules developed with it. The documentation website is versioned, which means documentation relating to prior versions of framework software can be archived, retrieved and viewed.

The documentation website was developed using the Hugo framework [@hugo_2023], Docsy theme [@docsy_2023] and Algolia DocSearch [@algoliadocsearch_2023] and is hosted using the Netlify [@netlify_2023] service.  We used functions from our foundation framework library to partially automate website updates relating to available CHEM modules, datasets and analysis programs. We linked our Netlify account to our GitHub organisation so that the project website would automatically update whenever its  source code (publicly available in a GitHub repository) was edited.

### 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Intro_Short.Rmd
first draft of reworked ethics paper

---
title: "MS_Intro_Short"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---

<!-- `r ifelse(is.na(params$X@background_1L_chr),"",params$X@background_1L_chr)`  -->

Health economics is a discipline concerned with problems that arise due to scarce resources, such as how to value health and healthcare, allocate healthcare budgets and configure health services [@wagstaff2012four]. In seeking to solve these problems, health economists typically use models which are simplified and selective representations of systems that are believed to influence human health. A health economic model should be capable of being described using words and figures (a conceptual model), equations (a mathematical model) and software (a computational model).  Health economic scientific manuscripts typically describe a model using its conceptual and mathematical representations but report results that have been generated by its computational representation (i.e., the format that allows computers to perform calculations). The conceptual, mathematical and computational representations of a model are assumed to be isomorphic. However, the ease of independently assessing the validity of this assumption depends in part on how a health economic model has been implemented computationally.

The set of software files implementing a health economic model's structure, data and algorithms can be called a computational health economic model (CHEM). CHEMs can be developed using commercial modelling software or open-source programming languages such as Python [@python2009] or R [@RCORE2022]. An advantage of commercial modelling software is that users often require limited or no programming skills to develop and apply CHEMs. However, open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022]. The decision about what type of software development platform to use must balance ethical considerations and resource constraints (see Table \@ref(tab:proscons)).

```{r, child = child_docs_ls$one, echo=FALSE}
```

Open-source CHEMs remain rare [@Feenstra2022; @Emerson2019]. The software used to develop CHEMs was identified as a barrier to open-source model implementations by participants in a 2020 survey of health economists [@Pouwels2022]. Other reported barriers relate to difficulties in generalising models, updating models and transferring model data, concerns about the level of public access and lack of interest from decision-makers, and legal, confidentiality and security considerations [@Pouwels2022].

<!-- Better software frameworks could potentially help address some of the barriers to open-source CHEM development. A software framework is a shared common technology used by developers to collaboratively author software and which is not typically visible to software end-users [@myllarniemi2018development]. A software framework provides a foundation for developing multiple software applications with shared resources (e.g., code and data files), that can be modified to suit specific needs. Advantages of using software frameworks include facilitating code reuse and extension, promoting good programming practice and the capability to provide enhanced functionality and performance without additional effort by developers [@edwin2014software]. Software frameworks have widely been developed and implemented in data science, for example, PyTorch [@PyTorch2019] for machine learning using Python.  -->

<!-- A high-level software framework for implementing open-source CHEMs in R has previously been developed with the primary aim of improving model transparency [@alarid2019need]. However, to gain widespread adoption, a software framework for CHEMs may need to provide tools to help model developers meet a broader range of ethical practice requirements. There are multiple challenges to overcome. Firstly, the ethical responsibilities of public health computational modellers remain poorly understood [@10.3389/fpubh.2017.00068] and the computational modelling landscape is changing fast [@calder2018computational]. Secondly, software frameworks can be challenging and time consuming to create and may become excessively complex over time [@edwin2014software]. Thirdly, software frameworks can be difficult to learn, often requiring model developers to undergo specialist training [@myllarniemi2018development]. -->

### 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Tables.Rmd
first draft of reworked ethics paper

---
title: "MA_Results"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---

```{r echo=FALSE}
if(params$X@tables_in_body_lgl){
  child_docs_ls <- list(one = "../Child_RMDs/MS_Tables_1.Rmd",
                        two = "../Child_RMDs/MS_Tables_2.Rmd"
                        # ,
                        # three = "../Child_RMDs/MS_Tables_3.Rmd",
                        # four = "../Child_RMDs/MS_Tables_4.Rmd"
                        )
}else{
    child_docs_ls <- list(one = NULL,
                          two = NULL,
                          three = NULL,
                          four = NULL)
}
```

### 2024-01-24 Markdown/Manuscript/Parent_PDF/Springer.Rmd
first draft of reworked ethics paper

---
# Supported options: 
#   sn-nature:       Style for submissions to Nature Portfolio journals
#   sn-basic:        Basic Springer Nature Reference Style/Chemistry Reference Style
#   sn-mathphys:     Math and Physical Sciences Reference Style
#   sn-aps:          American Physical Society (APS) Reference Style
#   sn-vancouver:    Vancouver Reference Style
#   sn-apa:          APA Reference Style 
#   sn-chicago:      Chicago-based Humanities Reference Style
#   default:         Default

classoptions: 
  - sn-vancouver      
  - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
  # - referee       # Optional: Use double line spacing 
  # - lineno        # Optional: Add line numbers
  # - iicol         # Optional: Double column layour

title: Ethical considerations for computational implementation of health economic models
titlerunning: Ethical computational modelling

# authors:
#   - name: Matthew P Hamilton
#     email: matthew.hamilton@orygen.org.au
#     address: Orygen, Parkville, Australia
#     correspondence: true
#     equal_contributor: "yes"
#   - name: Caroline X Gao
#     address: Orygen, Parkville, Australia, Centre for Youth Mental Health, University of Melbourne, Parkville, Australia and School of Public Health and Preventive Medicine, Monash University, Clayton, Australia
#     equal_contributor: "yes"
#   - name: Kate M Filia
#     address: Orygen, Parkville, Australia and Centre for Youth Mental Health, University of Melbourne, Parkville, Australia
#   - name: Jana M Menssink
#     address: Orygen, Parkville, Australia and Centre for Youth Mental Health, University of Melbourne, Parkville, Australia
#   - name: Sonia  Sharmin
#     address: Orygen, Parkville, Australia and Centre for Youth Mental Health, University of Melbourne, Parkville, Australia
#   - name: Nic Telford
#     address: headspace National Youth Mental Health Foundation, Melbourne, Australia
#   - name: Helen Herrman
#     address: Orygen, Parkville, Australia and Centre for Youth Mental Health, University of Melbourne, Parkville, Australia
#   - name: Ian B Hickie
#     address: Brain and Mind Centre, Youth Mental Health & Technology, Faculty of Medicine & Health, The University of Sydney,  Australia
#   - name: Cathrine  Mihalopoulos
#     address: Deakin Health Economics, Institute for Health Transformation, Deakin University, Geelong,  Australia
#   - name: Debra J Rickwood
#     address: Faculty of Health, University of Canberra, Australia and headspace National Youth Mental Health Foundation, Melbourne, Australia
#   - name: Patrick D McGorry
#     address: Orygen, Parkville, Australia and Centre for Youth Mental Health, University of Melbourne, Parkville, Australia
#   - name: Sue M Cotton
#     address: Orygen, Parkville, Australia and Centre for Youth Mental Health, University of Melbourne, Parkville, Australia
authors: 
  - firstname: Matthew
    particle: P 
    lastname: Hamilton
    email: matthew.hamilton1@monash.edu
    affiliation: [1,2]
    corresponding: TRUE
  # - prefix: Mr.
    #suffix: III
    #naturename: Poet Laureate
    # degrees: MSc #, PhD
    #equalcont: These authors contributed equally to this work.

  # - firstname: Glen
  #   lastname: Wiesner
  #   #     email: glen.wiesner@heartfoundation.org.au
  #   affiliation: [4]
  # 
  # - firstname: Kate  
  #   particle: M
  #   lastname: Filia
  #   #     email: kate.filia@orygen.org.au
  #   affiliation: [2,3]
  # 
  # - firstname: Jana
  #   particle: M
  #   lastname: Menssink
  #   #     email: jana.menssink@orygen.org.au
  #   affiliation: [2,3]
  # 
  # - firstname: Petra
  #   lastname: Plencnerova
  #   #     email: pplencnerova@headspace.org.au
  #   affiliation: [5]
  # 
  # - firstname: David  
  #   lastname: Baker
  #   #     email: david.baker@orygen.org.au
  #   affiliation: [2,3]
  # 
  # - firstname: Patrick 
  #   particle: D 
  #   lastname: McGorry
  #   #     email: pat.mcgorry@orygen.org.au
  #   affiliation: [2,3]
  # 
  # - firstname: Alexandra  
  #   lastname: Parker
  #   #     email: alex.parker@vu.edu.au
  #   affiliation: [6,3]

  - firstname: Jonathan  
    lastname: Karnon
    #     email: jonathan.karnon@flinders.edu.au
    affiliation: [3]

  - firstname: Caroline
    lastname: Gao
    #     email: caroline.gao@orygen.org.au
    affiliation: [2,4,1]

  - firstname: Luis  
    lastname: Salvador-Carulla
    #     email: jonathan.karnon@flinders.edu.au
    affiliation: [5]

  - firstname: Sue
    particle: M
    lastname: Cotton
    #     email: sue.cotton@orygen.org.au
    affiliation: [2,4]

  - firstname: Cathrine
    lastname: Mihalopoulos
    #     email: cathy.mihalopoulos@deakin.edu.au
    affiliation: [1]

affiliations:
  - number: 1
    corresponding: TRUE
    info:
      orgdiv: School of Public Health and Preventive Medicine
      orgname: Monash University
    address:
        #street: "553 St Kilda Road"
        city: Melbourne
        #postcode: 3004
        #state: Victoria
        country: Australia
  - number: 2
    corresponding: FALSE
    info:
      orgname: Orygen
    address:
     #   street: "35 Poplar Road"
        city: Parkville
        # postcode: 3052
        #state: Victoria
        country: Australia
  - number: 3
    corresponding: FALSE
    info:
#      orgdiv: XXXXXXXXXXX
      orgname: Flinders University   
    address:
    #     # street: "182 Victoria Square"     
        city: Adelaide
    #     postcode: 5000
    #  #   state: South Australia
        country: Australia 
  - number: 4
    corresponding: FALSE
    info:
      orgdiv: Centre for Youth Mental Health
      orgname: The University of Melbourne
    address:
     #   street: "35 Poplar Road"
        city: Parkville
        # postcode: 3052
    #   state: Victoria
        country: Australia
#   - number: 4
#     corresponding: FALSE
#     info:
# #      orgdiv: XXXXXXXXXXX
#       orgname: Heart Foundation      
#     address:
#      #   street: "2/850 Collins St"
#         city: Docklands
#         # postcode: 3008
#      #   state: Victoria
#         country: Australia
#   - number: 5
#     corresponding: FALSE
#     info:
# #      orgdiv: XXXXXXXXXXX
#       orgname: headspace National Youth Mental Health Foundation   
#     address:
#     #  #   street: "485 La Trobe St"
#         city: Melbourne
#     #     postcode: 3000
#     #  #   state: Victoria
#         country: Australia     
#   - number: 6
#     corresponding: FALSE
#     info:
#       orgdiv: Institute for Health and Sport
#       orgname: Victoria University 
#     address:
#     #  #   street: "70/104 Ballarat Rd"
#         city: Footscray
#     #     postcode: 3011
#     #  #   state: Victoria
#         country: Australia 
#       # 
  - number: 5
    corresponding: FALSE
    info:
#      orgdiv: XXXXXXXXXXX
      orgname: University of Canberra  
    address:
    #     # street: "182 Victoria Square"     
        city: Canberra
    #     postcode: 5000
    #  #   state: South Australia
        country: Australia 
keywords:  
  - computational models
  - ethics of modelling
  - health economics
  - open-source models

pacs: 
  jel:
    - "C63"
    - "C88"
    - "I10"
  msc:
    - "91-08"

# Sample for unstructured abstract

params:
  X: NULL

abstract: "`r paste0(readLines('../Header/abstract.txt'),collapse=' ')`"

bibliography: bibliography.bib

header-includes: |
  %% Per the spinger doc, new theorem styles can be included using built in style, 
  %% but it seems the don't work so commented below
  %\theoremstyle{thmstyleone}%
  \newtheorem{theorem}{Theorem}%  meant for continuous numbers
  %%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
  %% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
  \newtheorem{proposition}[theorem]{Proposition}%
  %%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

  %% \theoremstyle{thmstyletwo}%
  \theoremstyle{remark}
  \newtheorem{example}{Example}%
  \newtheorem{remark}{Remark}%

  %% \theoremstyle{thmstylethree}%
  \theoremstyle{definition}
  \newtheorem{definition}{Definition}%

  \usepackage{float}
  \let\origfigure\figure
  \let\endorigfigure\endfigure
  \renewenvironment{figure}[1][2] {
      \expandafter\origfigure\expandafter[H]
  } {
      \endorigfigure
  }
  \usepackage{lscape}
  \usepackage{pdflscape}
  \newcommand{\blandscape}{\begin{landscape}}
  \newcommand{\elandscape}{\end{landscape}}
  \usepackage[none]{hyphenat}
  \usepackage[export]{adjustbox}
  \usepackage{multirow}
  \usepackage{longtable}
  \usepackage{array}
  \usepackage{wrapfig}
  \usepackage{colortbl}
  \usepackage{booktabs} 
  \usepackage{colortbl}
  \usepackage{pdflscape}
  \usepackage{tabu}
  \usepackage{threeparttable}
  \usepackage[normalem]{ulem}
  \usepackage{threeparttablex}
  \usepackage{makecell}

# output: rticles::springer_article
output: 
  bookdown::pdf_book: 
    base_format: rticles::springer_article
    # includes:
    #   in_header: preamble.tex
    keep_tex: true
    pandoc_args: 
      - '--lua-filter=../Filters/scholarly-metadata.lua'
      - '--lua-filter=../Filters/author-info-blocks.lua'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      out.extra = "")
#library(flextable)# This forces knitr to label all figures.
```

```{r child="../Child_RMDs/All_Content.Rmd", echo=FALSE, warning=FALSE}
```

### 2024-01-24 setup.R

@@ -7,13 +7,13 @@ X <- Ready4showSynopsis(#background_1L_chr = "Our study is entirely fictional.",
                         coi_1L_chr = "None declared.",
                         #conclusion_1L_chr = "We have identified criteria that can be used to systematically assess extent to which the computational implementation of health economic models adheres to the ethical goals of transparency, reusability and updatability. We have developed an open-source software framework that can support the ethical computational implementation of economic models in youth mental health. Our framework can be used as a prototype for developing future software frameworks to support ethical implementation of CHEMs.",
                         digits_int = 3L,
-                        ethics_1L_chr = "Software framework development did not involve human subject research and was not ethically reviewed. The utility mapping worked example is a previously reported study that was reviewed and granted approval by the University of Melbourne’s Human Research Ethics Committee, and the local Human Ethics and Advisory Group (1645367.1).",
-                        funding_1L_chr = "Software framework development was funded by Orygen, VicHealth, Victoria University and an Australian Government Research Training Program (RTP) Scholarship. The utility mapping study used as a worked example was funded by the National Health and Medical Research Council (NHMRC, APP1076940), Orygen and headspace.",
+                        ethics_1L_chr = "None.",
+                        funding_1L_chr = "Funded by an Australian Government Research Training Program (RTP) Scholarship. ",
                         #interval_chr = "three months",
-                        keywords_chr = c("computational models", "ethics of modelling", "health economics", "mental disorders", "open-source models", "software frameworks"),
+                        keywords_chr = c("computational models", "ethics of modelling", "health economics", "open-source models"),
                         outp_formats_chr = "PDF",
                         #sample_desc_1L_chr = "The study sample is fake data that pretends to be young people aged 12 to 25 years who attended Australian primary care services for mental health related needs between November 2019 to August 2020.",
-                        title_1L_chr = "A prototype software framework for transparent, reusable and updatable computational health economic models")
+                        title_1L_chr = "Ethical considerations for computational implementation of health economic models")
 X <- renewSlot(X,
                "authors_r3", 
                {
@@ -43,16 +43,14 @@ X <- renewSlot(X,
                                                         We propose six criteria for assessing CHEMs: (T1) software files are open access; (T2) project team contributions and judgments are easily identified; 
                                                         (R1) programming practices promote generalisability and transferability; (R2) licenses restrict only unethical reuse; (U1) maintenance infrastructure is in place; and 
                                                         (U2) new releases are systematically retested and appropriately deprecated.
[…]

### 2024-02-14 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
more edits

@@
 ---
-
-In this paper, we <!--  MOVED_EDITED_START -->consider literature on modelling practice and draw on our prior experience with developing and using CHEMs to propose<!--  MOVED_EDITED_START -->:  (i) a set of ethical responsibilities for CHEM project teams<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.
+ <!-- Important rewordings --> 
+In this paper, we <!--  MOVED_EDITED_START -->draw on our prior experience with developing and using CHEMs and consider literature on modelling practice to propose<!--  MOVED_EDITED_END -->:  (i) a set of ethical <!--  EDITED_Start responsibilities-->goals<!--  _EDITED_END --> for CHEM <!--  EDIT -->developers and funders<!--  /EDIT --<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that can reduce ethical risks; and (iii) criteria against which these attributes can be assessed.
 <!--  MOVED_EDITED_END -->

### 2024-02-14 Markdown/Manuscript/Child_RMDs/MS_Discussion.Rmd
more edits

@@
 Ethical practice is a core expectation of health researchers and computational methods underpin most quantitative research, yet an understanding of what constitutes ethical public health modelling practice is underdeveloped [@10.3389/fpubh.2017.00068].
-We have addressed this gap with our proposed responsibilities for model developers and funders, enabling CHEM attributes and CHEM implementation assessment criteria.
+We have addressed this gap with our proposed goals for model developers and funders, enabling CHEM attributes and CHEM implementation assessment criteria.
 
@@
 Previously recommended strategies for more beneficial health economic research investments include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019].
-Development of software frameworks to support ethical CHEM implementations could enable and enhance each of these strategies.
+
+<!-- Development of software frameworks to support ethical CHEM implementations could enable and enhance each of these strategies. --> ????
 

### 2024-02-14 Markdown/Manuscript/Child_RMDs/MS_Imp_Guidelines.Rmd
more edits

@@
 -->
-<!-- # -->
-# Ethical responsibilities of CHEM project teams
+<!-- responsibilities to goals -->
+# Ethical goals for CHEM developers and funders
 Boden and McKendrick [@10.3389/fpubh.2017.00068] propose a framework for ethical public health modelling based on the criteria of independence (concerning how modeller subjectivity shapes model design), beneficence / non-maleficence (concerning model quality and utility), transparency (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and justice (concerning the social obligations of modellers to consider and communicate ethical issues about model use).
-However, these criteria are not specific to the computational model, but are designed to apply to a modelling project as a whole, including a model's conceptual and mathematical representations.
+<!-- However, t -->These criteria are not specific to the computational model, but <!-- are designed to  -->apply to a modelling project as a whole, including a model's conceptual and mathematical representations.
 The authors propose 13 questions to evaluate ethical risk across the four criteria, but of these, only "is the model code open source or available on request" (for transparency) and "has the model been verified, i.e., does it do what the modeller wants it to do?" (for beneficence / non-maleficence) are specific to model's representation as software.
 
-To identify ethical issues that are specific to CHEMs, we first considered the distinct phases in a software development project.
+To identify ethical issues that are specific to CHEMs, we <!-- first  -->considered <!-- Edit -->how Boden and McKendrick's four criteria might apply to <!-- /Edit -->the distinct phases in a software development project.
 Software development lifecycle (SDLC) models take different approaches to the inclusion, naming, definition, sequencing and iteration of project phases [@ruparelia2010software].
 However, SDLC models typically have components that can principally map to the concepts of *planning* (e.g., identification of system / user requirements and design specifications), *implementation* (e.g., development and testing) and *release* (e.g., deployment, system integration, maintenance and support).
-For simplicity, we propose one overarching ethical responsibility for CHEM project teams for each phase.
-These responsibilities are to:
+For simplicity, we propose one overarching ethical <!-- responsibility  for CHEM project team-->goal for each phase.
+These goals<!-- responsibilities  --> are:
+<!-- establish  -->
+(i) *socially acceptable user requirements and design specifications* (during CHEM planning);
+<!-- ensure  -->
+(ii) *fitness for purpose* (of CHEM implementations); and
+<!-- support  -->
+(iii) *socially beneficial use* (of CHEM releases).
 
-(i) *establish socially acceptable user requirements and design specifications* (during CHEM planning);
-
-(ii) *ensure fitness for purpose* (of CHEM implementation); and
-
-(iii) *support socially beneficial use* (on CHEM release).
-
-Transparency underpins all three responsibilities (Table \@ref(tab:timelygls)).
-Boden and McKendrick’s other ethical criteria can be selectively mapped to each phase's responsibility: justice and independence to the planning responsibility; independence to the implementation responsibility; and beneficence / non-maleficence and justice to the release responsibility.
-All three ethical responsibilities have implications for both professional practice and project resourcing.
-As it is unreasonable to place expectations of modellers that they are not resourced to fulfil, these responsibilities should apply jointly to health economic model developers and funders.
+Boden and McKendrick’s transparency criterion underpins the ethical goals we propose for all three phases (Table \@ref(tab:timelygls)).
+Boden and McKendrick’s other <!-- ethical  -->criteria can be selectively mapped to goals for specific phases: our proposed goal for ethical CHEM planning incorporates justice and independence considerations; the goal proposed for CHEM implementation is based on independence and beneficence / non-maleficence criteria; and the goal for ethical CHEM releases reflects beneficence / non-maleficence and justice criteria.
+<!-- /Edit -->
 
 Misalignment between the values of computational model developers and those of the population groups affected by decisions based on their models presents significant ethical risks [@thompson2022escape; @duckett2022journey].
-The value judgments of health economic modellers are rarely adequately specified, omissions that may lead to socially unacceptable policy recommendations [@duckett2022journey].
+The value judgments of health economic modellers are rarely adequately specified [@duckett2022journey].
 These value judgments influence the assumptions, selection of model features and standards for evidence that shape health economic model projects [@HARVARD2020112975].
-For example, as illustrated in Table \@ref(tab:proscons), project resource constraints may require model developers to make trade-offs when agreeing CHEM *user requirements and design specifications*.
-These trade-offs in turn determine how straightforward it is for a CHEM to be subsequently amended to reflect alternative value judgments relating to model conceptualisation and mathematical formalisation (Table \@ref(tab:timelygls)).
+For example, the process of determining CHEM *user requirements and design specifications* will involve making judgments about the competing claims of alternative software development platforms (Table \@ref(tab:timelygls)).
+That decision will in turn determine how straightforward it is for a CHEM to be subsequently amended to reflect alternative value judgments relating to model conceptualisation and mathematical formalisation (Table \@ref(tab:timelygls)).
+<!-- Moved -->
+<!--  , omissions that may lead to socially unacceptable policy recommendations -->
 
-Modellers have duties both to ensure a computational model is *fit for purpose* and to provide potential third-party users with the means of assessing its adequacy for their proposed uses [@Erdemir2020; @Feenstra2022; @thompson2019escape; @10.3389/fpubh.2017.00068].
+Modellers need to do more than self-certify that their <!-- have duties both to ensure a  -->computational model is *fit for purpose* - <!-- Edit -->they also need to<!-- /Edit --> provide potential third-party users with the means of assessing its adequacy
+<!-- for their proposed uses  -->[@Erdemir2020; @Feenstra2022; @thompson2019escape; @10.3389/fpubh.2017.00068].
 However, it is common for health economic models to have serious methodological flaws [@carletto_zanuzzi_sammarco_russo_2020; @WONDER2015467]; insufficient validation [@Ghabri2019; @kolovos2017model; @haji2013model], poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017]; and undeclared errors [@Radeva2020].
-Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and facilitating manual reviews by third parties.
+Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and <!-- Edit -->providing executable code for  <!--facilitating manual  -->review and testing <!-- /Edit -->by third parties.
 
-A CHEM will have limited *social benefit* if it is rarely used, if mis-used or when its acceptability and adequacy rapidly decay.
+The *social benefit* of developing a CHEM may be limited if it is rarely used, if mis-used or when its acceptability and adequacy rapidly decay.
 Reuse of CHEMs as components of other models can potentially make model development more efficient [@Arnold2010; @garcia2021cost].
@@
 Currently, health economic models are rarely implemented computationally in a manner that facilitates routine updates [@Sampson2017], thus limiting the temporal window within which a CHEM can be validly applied.
+<!-- Moved -->
+<!-- Edit --><!-- All three  -->Our proposed ethical <!-- responsibilities  --> goals are most likely to be realised when there is both good<!-- have implications for both  -->professional practice by CHEM developers and <!-- Addition -->adequate <!-- Addition -->resourcing by project funders.
+<!-- /Edit --> The choices CHEM developers make often need to balance ethical considerations with project resource constraints (as illustrated in Table \@ref(tab:proscons)).
+Modelling projects implemented by diligent and conscientious CHEM developers may still be exposed to ethical risks if there are inadequate resources to meaningfully engage relevant stakeholders, prepare code and data artefacts for independent assessment and provide ongoing model maintenance and user support.
+For these reasons, we believe responsibility for achieving our proposed ethical goals is shared between CHEM developers and funders.
+Reflecting the shared nature of this responsibility, we suggest that in most cases where a modelling project fails to meet our ethical goals, it will be more appropriate to describe these shortcomings as "ethical risks" (the language used by Boden and McKendrick) than to use accusatory terms like "unethical practice."<!-- these responsibilities should apply jointly to health economic model developers and funders.
+-->
+<!-- For example, as illustrated in Table \@ref(tab:proscons), project resource constraints may require model developers to make trade-offs -->
+<!-- As it is unreasonable to place expectations of modellers that they are not resourced to fulfil,  -->
+<!-- /Moved -->
 
 <!-- # -->
-# CHEM attributes associated with ethical modelling practice
-The responsibilities of model project teams over a CHEM lifecycle are easier to state than to measure.
-However, aspects of ethical modelling practice may be inferred from measurable attributes of CHEMs.
-As described in Table \@ref(tab:timelygls),  we believe that the creators of transparent, reusable and updatable (TRU) CHEMs are likely to have fulfilled a number of their ethical obligations.
-We therefore selected these model attributes to use as the basis for deriving ethical assessment criteria.
+# CHEM attributes that can reduce ethical risk
+<!-- The responsibilities of model project teams over a CHEM lifecycle  -->
+Our proposed ethical goals for each project phase are are easier to state than to measure.
+However, <!-- aspects of ethical modelling practice  -->achievement of these goals may be inferred from measurable attributes of CHEMs.
+As described in Table \@ref(tab:timelygls),  we believe that the <!-- Edit -->developers and funders<!-- /Edit --> of CHEMs that are transparent, reusable and updatable (TRU) are more likely to have achieved the ethical goals we propose.
+<!-- fulfilled a number of their ethical obligations.
+-->
+<!-- We therefore selected these CHEM attributes to use as the basis for deriving ethical assessment criteria. -->
 
@@
 
-Taking steps to make computational models *transparent* is an important part of ethical public health modelling practice [@10.3389/fpubh.2017.00068].
+Taking steps to make computational models *transparent* is an important means of reducing ethical risk<!-- part of ethical public health modelling practice  -->[@10.3389/fpubh.2017.00068].
 Guidance on transparency in health economic modelling recommend that model code and data should be clearly documented [@Eddy2012].
@@
 
-A CHEM that is *reusable* also signals ethical modelling practice.
-Although making a CHEM’s code, data and documentation publicly available is increasingly considered good practice, it is insufficient to enable model re-use.
+<!-- A CHEM that is *reusable* also signals ethical modelling practice.  -->
+Although making a CHEM’s code, data and documentation publicly available is increasingly considered good practice, it is insufficient to ensure that a model is *reusable*.
 Key concerns for health economists when considering whether to reuse a model are generalisability (application without adaptation) and transferability (selective reuse and/or modification of model components) [@RN39].
 Writing model algorithms as collections of functions (short, self-contained and reusable software routines that each perform a discrete task) is good scientific computing practice [@Wilson_2017] and promotes selective reuse.
-Computational implementations that store model code and data in distinct files and locations (as opposed to embedding data such as parameter values into source code) are easier to selectively modify.
+Computational implementations that store model code and data in distinct files and locations (as opposed to embedding data such as parameter values into source code) are also easier to selectively modify.
 Modular implementations enable models to be constructed from multiple independently reusable and replaceable sub-models (modules) [@pan2021modular].
@@
 
--	*R1*: Programming practices facilitate both generalisability and transferability.
+-	*R1*: Programming practices facilitate selective reuse of model sub-components.<!-- Replacement -->
 

### 2024-02-14 Markdown/Manuscript/Child_RMDs/MS_Intro_Short.Rmd
more edits

@@
 Health economic scientific manuscripts typically describe a model using its conceptual and mathematical representations but report results that have been generated by its computational representation (i.e., the format that allows computers to perform calculations).
-The conceptual, mathematical and computational representations of a model are assumed to be isomorphic.
+The conceptual, mathematical and computational representations of a model are assumed to be internally consistent.
 However, the ease of independently assessing the validity of this assumption depends in part on how a health economic model has been implemented computationally.
@@
 The set of software files implementing a health economic model's structure, data and algorithms can be called a computational health economic model (CHEM).
-CHEMs can be developed using commercial modelling software or open-source programming languages such as Python [@python2009] or R [@RCORE2022].
-An advantage of commercial modelling software is that users often require limited or no programming skills to develop and apply CHEMs.
-However, open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022].
-The decision about what type of software development platform to use must balance ethical considerations and resource constraints (see Table \@ref(tab:proscons)).
+CHEMs can be developed using commercial modelling <!-- Addition -->or spreadsheet<!-- /Addition --> software or open-source programming languages such as Python [@python2009] or R [@RCORE2022].
+An advantage of commercial <!-- modelling  -->software is that users often require limited or no programming skills to develop and apply CHEMs.
+<!-- Moved -->Open-source CHEMs remain rare [@Feenstra2022; @Emerson2019]<!-- Moved -->.
+However, <!-- Addition -->it has been argued that<!-- /Addition --> open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022].
+The decision about what type of software development platform to use must balance ethical considerations and resource constraints.
+<!-- ( see--><!-- Addition --> In Table \@ref(tab:proscons), we summarise some of the trade-offs we consider when undertaking modelling projects<!-- ) -->.
 
@@
 
-Open-source CHEMs remain rare [@Feenstra2022; @Emerson2019].
-The software used to develop CHEMs was identified as a barrier to open-source model implementations by participants in a 2020 survey of health economists [@Pouwels2022].
+<!--  The software used to develop CHEMs was identified as a barrier to open-source model implementations by participants in a 2020 survey of health economists [@Pouwels2022].
 Other reported barriers relate to difficulties in generalising models, updating models and transferring model data, concerns about the level of public access and lack of interest from decision-makers, and legal, confidentiality and security considerations [@Pouwels2022].
+-->
 
@@
 There are multiple challenges to overcome.
-Firstly, the ethical responsibilities of public health computational modellers remain poorly understood [@10.3389/fpubh.2017.00068] and the computational modelling landscape is changing fast [@calder2018computational].
-Secondly, software frameworks can be challenging and time consuming to create and may become excessively complex over time [@edwin2014software].
+-->
+<!-- Edit Move -->Determining how to make these trade offs can be difficult.
+The computational modelling landscape is changing fast [@calder2018computational] and the ethical <!-- responsibilities  --><!-- Edit -->dimensions of computational public health modelling <!-- /Edit -->remain poorly understood [@10.3389/fpubh.2017.00068].
+<!-- /Edit Move-->
+
+<!-- Secondly, software frameworks can be challenging and time consuming to create and may become excessively complex over time [@edwin2014software].
 Thirdly, software frameworks can be difficult to learn, often requiring model developers to undergo specialist training [@myllarniemi2018development].

### 2024-02-14 Markdown/Manuscript/Child_RMDs/MS_Tables_1.Rmd
more edits

---
title: "Predr_Cors"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---
```{r proscons, tab.cap='Implications of choosing commercial modelling or spreadsheet software or an open-source programming language for the computational implementation of health economic models', tab.id = 'proscons', echo = F, results = 'asis'}
library(kableExtra)  
df <- data.frame(`Implications` = c(rep("Resources",4), rep("Ethical",10)),
                 `Commercial modelling or spreadsheet software` = c("(+) Will include tools to efficienty accomplish many common modelling tasks.",
                                                     "(+) Requires limited or no programming skills.",
                                                     "(-) Requires model developers and users to purchase (in some cases, expensive) software licenses.",
                                                     "",
                                                     "Transparency",
                                                     "(+) Popular tools (e.g. Excel) are readily understood by many.",
                                                     "(-) Requirement to manually transfer results to manuscripts through repeated copy and paste actions or transcription from visual inspection increases risk of (undetected) reporting errors.",
                                                     "",
                                                     "Reusability",
                                                     "(+) Files will reliably open and execute correctly for years after project is completed.",
                                                     "",
                                                     "",
                                                     "Updatability",
                                                     "(-) Modifications and adaptations, particularly by third parties, may require bespoke permissions and it can be difficult to clearly itemise all changes."),
                 `Open-source programming language` = c("(-) Initial development is likely to be time and resource intensive.",
                                                        "(+) Over medium term, efficiency savings are possible if:",
                                                        "(i) artefacts (code and data) from pre-existing models are leveraged; and / or",
                                                        "(ii)	maintaining and transferring (e.g. to new decision problems / contexts) model.",
                                                        "Transparency",
                                                        "(+) Easy to publicly share testable source code while selectively restricting access to confidential data.",
                                                        "(+) Easy to make clear who has developed and tested each part.",
                                                        "(-) Novel code increases likelihood of (undetected) software errors.",
                                                        "Reusability",
                                                        "(-) Potentially fragile – if not maintained / not bundled with all required and correctly versioned dependencies.",
                                                        "(+) Facilitates modular implementations that make complex model representations more tractable.",
                                                        "(+) Facilitates customised integration with existing data / decision systems.",
                                                        "Updatability",
                                                        "(-) Facilitates collaborative development and maintenance."
                                                        ),
                 check.names = F)
if(params$X@outp_formats_chr[1] == "Word") {
  library("flextable")
  library("ftExtra")
  df %>%
    ready4show::print_table(output_type_1L_chr = params$X@outp_formats_chr[1],
                            caption_1L_chr = knitr::opts_current$get("tab.cap"),
                            mkdn_tbl_ref_1L_chr = paste0("tab:",knitr::opts_current$get("tab.id")),
                            use_rdocx_1L_lgl = ifelse(params$X@outp_formats_chr[1]=="Word",T,F),
                            add_to_row_ls = NULL,
                            sanitize_fn = NULL) %>%
    ftExtra::colformat_md()
  }else{
    df %>%
      dplyr::filter(!`Commercial modelling or spreadsheet software` %in% c("Transparency", "Reusability","Updatability")) %>%
      kableExtra::kbl(booktabs = T, 
                      escape = F,
                      caption = knitr::opts_current$get("tab.cap")) %>%
      kableExtra::kable_paper(full_width = F) %>%
      #kableExtra::pack_rows("Resource requirements implications",1,4, hline_after = T) %>%
      #kableExtra::pack_rows("Ethical implications",5,14, hline_after = T) %>%
      kableExtra::column_spec(1, width = "7em", bold=T) %>% #, latex_valign = "m"
      kableExtra::column_spec(2, width = "20em") %>%
      kableExtra::column_spec(3, width = "20em") %>%
      kableExtra::collapse_rows(columns = 1, latex_hline = "major") %>%
      kableExtra::pack_rows("Transparency",5,7, latex_align = "c") %>% #bold = F, italic = T
      kableExtra::pack_rows("Reusability",8,10, latex_align = "c") %>%
      kableExtra::pack_rows("Updatability",11,11, latex_align = "c") 
      # kableExtra::pack_rows("Accountable - it is easy to see who developed, tested and applied a model and how they did it.",1,7) %>%
      # kableExtra::pack_rows("Updatable - a model and its components are maintained and continuously improved.",17,20) %>%
      # kableExtra::pack_rows("Modular - models and their components can be combined to extend their scope",11,12) %>%
      # kableExtra::pack_rows("Epitomising - model code can be re-used in other decision contexts",13,15) %>%
      # kableExtra::pack_rows("Reusable - a model and its components can be used in other models and by other modellers.",8,16) #%>%
      # kableExtra::pack_rows("Yielding - a model can be simply, flexibly and reliably used as a decision aid",19,20)

    }
```

### 2024-02-14 Markdown/Manuscript/Child_RMDs/MS_Tables_2.Rmd
more edits

---
title: "Predr_Cors"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---
```{r timelygls, tab.cap='Transparent, reusable and updatable (TRU) computational health economic models (CHEMs) are suggestive of ethical modelling practice', tab.id = 'timelygls', echo = F, results = 'asis', warning=FALSE}
library(kableExtra)  
df <- data.frame(` ` = c("Transparent", "Reusable", "Updatable"),
                 `Social acceptability` = c("Modeller judgments, model features and verification checks can be reviewed by third parties.",
                                                                        "Models can be modified to reflect alternative value judgments.",
                                                                        "Models can be modified to reflect alternative value judgments."),
                 `Fitness for purpose` = c("Modeller judgments, model features and verification checks can be reviewed by third parties.","Third party use increases likelihood of errors being identified.","Models can be validly used for longer."),
                 `Beneficial use` = c("Decision-makers can understand the strengths and weakness of models before applying them.","Can inform more decision problems, in more contexts with less duplicative effort.","Models can be validly used for longer."),
                 check.names = F)
if(params$X@outp_formats_chr[1] == "Word") {
  library("flextable")
  library("ftExtra")
  df %>%
    ready4show::print_table(output_type_1L_chr = params$X@outp_formats_chr[1],
                            caption_1L_chr = knitr::opts_current$get("tab.cap"),
                            mkdn_tbl_ref_1L_chr = paste0("tab:",knitr::opts_current$get("tab.id")),
                            use_rdocx_1L_lgl = ifelse(params$X@outp_formats_chr[1]=="Word",T,F),
                            add_to_row_ls = NULL,
                            sanitize_fn = NULL) %>%
    ftExtra::colformat_md()
  }else{
    # library(huxtable)
    # library(dplyr)
    #flextable::set_flextable_defaults() #fonts_ignore=TRUE
    # flextable::flextable(df) %>%
    #   flextable::border_inner(border = officer::fp_border(style = "none")) %>%
    #   flextable::border_outer() %>%
    #   flextable::align(align = "center", part = "header") %>%
    #   flextable::merge_h() %>%
    #   flextable::merge_v() %>%
    #   flextable::valign(j=2, valign = "top") %>%
    #   flextable::align(j=2:4, align = "center") %>%
    #   flextable::bold(j=1, part = "all") %>%
    #   flextable::bold(part = "header") %>%
    #   flextable::width(j = 1, width = 1.2) %>%
    #   flextable::width(j = 2:4, width = 1.6)
    # %>%
    #   
      #flextable::merge_at(i=2,j=1:2)
    # huxtable::as_hux(df) %>% 
    #   #insert_row("MPG-DISP", "", "", "HP-WT", fill = "") %>% 
    #   #huxtable::merge_cells(2:3,1) %>% 
    #  # huxtable::set_header_rows(1, TRUE) %>% 
    #   huxtable::set_col_width(c(.1,.1,.1,.7)) %>%
    #   huxtable::set_wrap(1, value=T)
      #huxtable::style_headers(bold = TRUE)
    df %>%
      kableExtra::kbl(booktabs = T,
                      escape = F,
                      caption = knitr::opts_current$get("tab.cap")) %>%
      kableExtra::kable_paper(full_width = F) %>%
      kableExtra::column_spec(1, bold=T) %>%
      kableExtra::add_header_above(c(" ", "Project team responsibility" = 3), bold = T, italic = T) %>%
      kableExtra::column_spec(2, width = "13em") %>%
      kableExtra::column_spec(3, width = "13em") %>%
      kableExtra::column_spec(4, width = "13em") %>%
      kableExtra::collapse_rows(columns = 2, latex_hline = "none")
      # kableExtra::column_spec(3, width = "35em") %>%
      # kableExtra::pack_rows("Accountable - it is easy to see who developed, tested and applied a model and how they did it.",1,7) %>%
      # kableExtra::pack_rows("Updatable - a model and its components are maintained and continuously improved.",17,20) %>%
      # # kableExtra::pack_rows("Modular - models and their components can be combined to extend their scope",11,12) %>%
      # # kableExtra::pack_rows("Epitomising - model code can be re-used in other decision contexts",13,15) %>%
      # kableExtra::pack_rows("Reusable - a model and its components can be used in other models and by other modellers.",8,16) #%>%
      # kableExtra::pack_rows("Yielding - a model can be simply, flexibly and reliably used as a decision aid",19,20)
    }
```

### 2024-02-14 Markdown/Manuscript/Child_RMDs/MS_Tables_3.Rmd
more edits

---
title: "Predr_Cors"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---
```{r cpkgs, tab.cap='Software framework R libraries', tab.id = 'cpkgs', echo = F, results = 'asis'}
library(kableExtra)  
df <- tibble::tibble(Package = c("ready4",
                                       "ready4fun",
                                       "ready4class",
                                       "ready4pack",
                                       "ready4use",
                                       "ready4show"),
                     Purpose = c("Provides a template model module and novel syntax and tools for finding interoperable CHEM modules, datasets and reproducible analysis programs.",
                                 "Streamline and standardise the authoring and documenting of functions that support selective reuse of model algorithms.",
                                 "Streamline and standardise the authoring and documenting of new CHEM modules.",
                                 "Help bundle and disseminate CHEM modules as documented, licensed and quality assured  R libraries.",
                                 "Help manage the labelling and transfer of data between CHEM modules and local and remote data repositories.",
                                 "Support the authoring of programs that implement integrated and reproducible data ingest, analysis and reporting pipelines."),
                           # A = c("3","2-3","2","1-2,5,7","1-2,7","4,6"), 
                           # R = c("4","5","4-5","1,3,6","3,7",""),
                           # U = c("","","","1-3","1",""),
                           `R library dependencies` = Package %>%
                             purrr::map(~ {
                               utils::packageDescription(.x) %>%
                                 `[`(c("Depends", "Imports")) %>%
                                 purrr::map(~{
                                   if(is.null(.x)){
                                     character(0)
                                     }else{
                                       .x %>%
                                         strsplit(",\\n") %>%
                                         purrr::flatten_chr() %>%
                                         purrr::map(~strsplit(.x,", ") %>%
                                                      purrr::flatten_chr()) %>%
                                         purrr::flatten_chr() %>% sort() %>%
                                         purrr::discard(~startsWith(.x,"R "))
                                       }
                                   }) %>%
                                 purrr::flatten_chr() %>%
                                 unique() %>%
                                 sort() %>%
                                 purrr::map_chr(~{
                                   updated_1L_chr <- stringr::str_replace_all(.x,"\\n"," ")
                                   problem_idx_1L_chr <- stringr::str_locate(updated_1L_chr," ")[1,1] %>%
                                     unname()
                                   if(!is.na(problem_idx_1L_chr))
                                     updated_1L_chr <- updated_1L_chr %>%
                                     stringr::str_sub(end = problem_idx_1L_chr-1)
                                   updated_1L_chr %>% 
                                     trimws(which = "left") 
                                   })
                               }) %>% 
                             purrr::map_chr (~ .x %>% paste0(collapse = " "))) %>% 
  as.data.frame() 
# if(params$X@outp_formats_chr[1] != "Word"){
#       df$Ref <- c(rep("",6)
#        # "@ready42022","@ready4fun2022","@ready4class2022","@ready4pack2022","@ready4use2022","@ready4show2022"
#                   )
# }else{
#   df$Ref <- ""
# }

df <- df %>%
  # dplyr::mutate(Standard = purrr::pmap_chr(df,
  #                                          ~paste0("A",
  #                                                 ..3,
  #                                                 paste0(rep("&nbsp;",
  #                                                     2+max(nchar(df[,3]))-nchar(..3)),
  #                                                     collapse = ""#
  #                                                     ),
  #                                                 ifelse(..4=="","","R"),
  #                                                 ..4,
  #                                                 paste0(rep("&nbsp;",
  #                                                     2+max(nchar(df[,4]))-nchar(..4)),
  #                                                     collapse = ""#
  #                                                     ),
  #                                                 ifelse(..5=="","","U"),
  #                                                 ..5,
  #                                                 paste0(rep("&nbsp;",
  #                                                     1+max(nchar(df[,5]))-nchar(..5)),
  #                                                     collapse = ""#
  #                                                     )#,
  #                                                 ))) %>%

  dplyr::select(Package,
                #Ref,
                Purpose,
                #Standard,
                `R library dependencies`)

names_chr <- names(df)
# df <- data.frame(df[,1],
#                  rep(ifelse(params$X@outp_formats_chr[1] == "Word"," ","&nbsp; &nbsp; &nbsp;"),
#                      nrow(df)),
#                  df[,2#:3
#                     ],
#                  rep(ifelse(params$X@outp_formats_chr[1] == "Word"," ","&nbsp; &nbsp; &nbsp;")
#                      ,nrow(df)),
#                  #df[,4],
#                  #rep("&nbsp; &nbsp; &nbsp;",nrow(df)),
#                  df[,3#5
#                     ])
# colnames(df) <- c("Package",
#                   ifelse(params$X@outp_formats_chr[1] == "Word","XXX"," "),
#                   #"Ref", 
#                   "Purpose", 
#                   ifelse(params$X@outp_formats_chr[1] == "Word","YYY"," "),
#                   #"Standard     ",# " ",
#                   "R library dependencies")
#df$Package <- "![](../Data/images/ready4.png){ width=10% }"#"\\includegraphics[valign=T,scale=2.5,raise=2mm]{../Data/images/ready4.png}"
#df$`Standard     `  <- df$`Standard     ` %>% stringr::str_replace_all("&nbsp;"," ")
if(params$X@outp_formats_chr[1] == "Word") {
  library("flextable")
  library("ftExtra")
  df %>% #[,c(1,3:4,6:7)] %>%
    ready4show::print_table(output_type_1L_chr = params$X@outp_formats_chr[1],
                            caption_1L_chr = knitr::opts_current$get("tab.cap"),
                            mkdn_tbl_ref_1L_chr = paste0("tab:",knitr::opts_current$get("tab.id")),
                            use_rdocx_1L_lgl = ifelse(params$X@outp_formats_chr[1]=="Word",T,F),
                            add_to_row_ls = NULL,
                            sanitize_fn = NULL) %>%
    ftExtra::colformat_md()
  }else{
    #df$Ref<-NULL
    #df$Package <- ""
    # df$Package  <- df$Package %>% 
    #                   purrr::map_chr(~paste0("\\includegraphics[valign=M,width=1.9cm,raise=4mm]{../Data/images/",.x,"_logo.png}"))#scale=0.4,raise=3mm]
    # df$Package <- "\\includegraphics[valign=M,scale=0.08,raise=3mm]{../Data/images/ready4.png}"
    #df[,c(1,3,5:6)]   %>%
    df %>%
    # kableExtra::kbl(format = "markdown",
    #                   escape = F,
    #                 caption = knitr::opts_current$get("tab.cap"))
      kableExtra::kbl(booktabs = T, 
                      escape = F,
                      caption = knitr::opts_current$get("tab.cap")) %>%
      kableExtra::kable_paper(full_width = F) %>%
      kableExtra::column_spec(2, width = "20em", latex_valign = "m") %>%
      # kableExtra::column_spec(1, 
      #                         image = kableExtra::spec_image(rep("../Data/images/ready4.png",nrow(df)), 
      #                                                           150, 150)) %>%
      kableExtra::column_spec(3, width = "20em",
                              latex_valign = "m")
    }
```

### 2024-02-14 Markdown/Manuscript/Child_RMDs/MS_Tables_4.Rmd
more edits

---
title: "Untitled"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---

```{r checktb, tab.cap='Assessment of a CHEM implementation in youth mental health using Transparent, Reusable and Updatable (TRU) criteria', tab.id = 'checktb', echo = F, results = 'asis'}
library(kableExtra)  
df <- data.frame(Criterion = c("T1 Software files are open access", 
                               "T2 Project team contributions and judgments are easily identified", 
                               "R1 Programming practices promote selective reuse",
                               "R2 Licenses restrict only unethical reuse",
                               "U1 Maintenance infrastructure is in place",
                               "U2 New releases are systematically retested and deprecated"),
                 `Met?` = c(rep("Yes",5),"Partial"),
                 Detail = c("All source code and testing procedures are available in public GitHub repositories, with each code release persistently available in a Zenodo repository. As the study dataset contains confidential patient records, it was not published. Instead, a synthetic representation of the study dataset is persistently available in a repository in the Harvard Dataverse. Data files to support out of sample application of models are published at the same location.",
                            "All module libraries, programs, and datasets are distributed with citation information. Public GitHub repositories detail author code contributions over project development history. Model catalogues persistently available on the Harvard Dataverse describe the predictive performance of models under multiple usage regimes. Each code library is documented with worked examples of how to apply CHEM modules.  Analysis and reporting programs are self-documenting.  Sub-routines for generating reports are documented with README files.",
                            "CHEMs are written using both functional and object-oriented paradigms. Code library documentation websites include hypothetical examples of generalisability (applying study algorithm to estimate mapping models from new data with the same predictor and outcome variables) and transferability (adapting study algorithm to develop mapping models from datasets with different predictor variables and outcomes measured with a different utility instrument)." ,
                            "All code is distributed using GPL-3 licenses. Datasets use an amended version of template terms provided by Harvard Dataverse, allowing reuse of data subject to some ethical restrictions (e.g., use in efforts to reidentify study participants is prohibited).",
                            "All code is version controlled using git and GitHub, with semantic versioning. Each code library has a specified maintainer and guidance for potential code contributors is available on the project website.",
                            "Continuous integration is used for all code libraries, primarily for acceptance testing. Only limited use is made of unit testing. Retired library code is deprecated using tools from the lifecycle R library. Library documentation articles and datasets are also deprecated."),
                 check.names = F)
if(params$X@outp_formats_chr[1] == "Word") {
  library("flextable")
  library("ftExtra")
  df %>%
    ready4show::print_table(output_type_1L_chr = params$X@outp_formats_chr[1],
                            caption_1L_chr = knitr::opts_current$get("tab.cap"),
                            mkdn_tbl_ref_1L_chr = paste0("tab:",knitr::opts_current$get("tab.id")),
                            use_rdocx_1L_lgl = ifelse(params$X@outp_formats_chr[1]=="Word",T,F),
                            add_to_row_ls = NULL,
                            sanitize_fn = NULL) %>%
    ftExtra::colformat_md()
  }else{
    df %>%
      kableExtra::kbl(booktabs = T, 
                      escape = F,
                      caption = knitr::opts_current$get("tab.cap")) %>%
      kableExtra::kable_paper(full_width = F) %>%
      kableExtra::column_spec(1, width = "10em", bold=T) %>%
      kableExtra::column_spec(3, width = "35em")
    }
```

### 2024-02-14 setup.R

@@ -38,10 +38,10 @@ X <- renewSlot(X,
                {
                  args_ls <- X@abstract_args_ls
                  args_ls$abstract_ls <- list(Summary = "Most health economic analyses are undertaken with the aid of computers. However, the ethical dimensions of implementing health economic models as software (or computational health economic models (CHEMs)) are poorly understood. 
-                                                        We propose that developers and funders of CHEMs share ethical responsibilities to (i) establish socially acceptable user requirements and design specifications; (ii) ensure fitness for purpose; and (iii) support socially beneficial use. 
-                                                        We further propose that a transparent (T), reusable (R) and updatable (U) CHEM is suggestive of a project team that has largely fulfilled these responsibilities.
-                                                        We propose six criteria for assessing CHEMs: (T1) software files are open access; (T2) project team contributions and judgments are easily identified; 
-                                                        (R1) programming practices promote generalisability and transferability; (R2) licenses restrict only unethical reuse; (U1) maintenance infrastructure is in place; and 
+                                                        We propose that developers and funders of CHEMs should pursue the ethical goals of (i) socially acceptable user requirements and design specifications; (ii) fitness for purpose; and (iii) socially beneficial use. 
+                                                        We further propose that a transparent (T), reusable (R) and updatable (U) CHEM is suggestive of a project team that has largely met these goals.
+                                                        We propose six criteria for assessing TRU CHEMs: (T1) software files are open access; (T2) project team contributions and judgments are easily identified; 
+                                                        (R1) programming practices promote selective reuse; (R2) licenses restrict only unethical reuse; (U1) maintenance infrastructure is in place; and 
                                                         (U2) new releases are systematically retested and appropriately deprecated.
                                                         Our assessment criteria can help inform and improve ethical implementation of CHEMs. Resource barriers to ethical CHEM practice should be addressed by research funders."
                                              # ,

### 2024-02-15 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
for coauthor feedback draft

--- 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
@@
 ---
-
-In this paper, we <!--  MOVED_EDITED_START -->consider literature on modelling practice and draw on our prior experience with developing and using CHEMs to propose<!--  MOVED_EDITED_START -->:  (i) a set of ethical responsibilities for CHEM project teams<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.
+ <!-- Important rewordings --> 
+In this paper, we <!--  MOVED_EDITED_START -->draw on our prior experience with developing and using CHEMs and consider literature on modelling practice to propose<!--  MOVED_EDITED_END -->:  (i) a set of ethical <!--  EDITED_Start responsibilities-->goals<!--  _EDITED_END --> for CHEM <!--  EDIT -->developers and funders<!--  /EDIT --<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that suggest achievement of these goals; and (iii) criteria against which these attributes can be assessed.
 <!--  MOVED_EDITED_END -->

### 2024-02-15 Markdown/Manuscript/Child_RMDs/MS_Conclusion.Rmd
for coauthor feedback draft

@@
 `r ifelse(is.na(params$X@conclusion_1L_chr),"",params$X@conclusion_1L_chr)` 
-We have identified criteria that can be used to systematically assess the extent to which the computational implementation of health economic models adheres to transparency, reusability and updatability criteria.
+<!-- We have identified criteria that can be used to systematically assess the extent to which the computational implementation of health economic models adheres to transparency, reusability and updatability criteria.
 <!-- We have developed a prototype software framework that can support the implementation of CHEMs that largely meet these criteria.
 Our framework can be used as a prototype for developing future software frameworks to support ethical implementation of CHEMs.
---> Barriers to ethical CHEM practice arising from resource constraints should be addressed by research funders.
+-->
+<!-- Barriers to ethical CHEM practice arising from resource constraints should be addressed by research funders. -->

### 2024-02-15 Markdown/Manuscript/Child_RMDs/MS_Discussion.Rmd
for coauthor feedback draft

@@
 
-Ethical practice is a core expectation of health researchers and computational methods underpin most quantitative research, yet an understanding of what constitutes ethical public health modelling practice is underdeveloped [@10.3389/fpubh.2017.00068].
-We have addressed this gap with our proposed goals for model developers and funders, enabling CHEM attributes and CHEM implementation assessment criteria.
+Ethical practice is a core expectation of health researchers and computational methods underpin most quantitative research.
+However, an understanding of the ethical dimensions about how health economic models are implemented as software is underdeveloped.
+We have attempted to address this gap by proposing that CHEM developers and funders adopt the ethical goals of socially acceptable user requirements and design specifications, fit for purpose implementations and socially beneficial post-release use.
+We suggest that CHEM attributes of transparency, reusability and updatability are indicative of projects that meet these goals and have identified six criteria for assessing these attributes.
 
-Currently, many if not most existing CHEMs are insufficiently transparent [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], not reusable [@Feenstra2022; @Emerson2019], nor updatable [@Sampson2017; @10.3389/fpubh.2022.899874].
-Addressing these issues will require changes to the tools, practices and funding underpinning health economic model development.
+Prior literature has noted that health economic models routinely have limitations in transparency [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], reuse [@Feenstra2022; @Emerson2019], and being kept up to date [@Sampson2017; @10.3389/fpubh.2022.899874].
+These limitations not only pose ethical risks such as those we have described, but are practical barriers to model generalisability and transferability.
+Addressing these limitations will require changes to the work practices, funding and tools that underpin health economic model development.
+
+Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020] and there is a need for better and comprehensive guidelines about how health economists should implement health economic models as software.
+Research funders need to actively support initiatives to help remove barriers to better practice in this area.
+A core responsibility of research funders is reducing waste in research [@chalmers2014increase].
+Investments in modelling projects might generate greater social returns if they supported the project activities necessary to develop CHEMs that are easily understood, widely reused and well maintained.
+
+Previously recommended strategies for more beneficial health economic research investments include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019].
+Each of these strategies could benefit from enabling technical infrastructure, for example software frameworks that support the development of CHEMs that meet ethical criteria.
+We are currently engaged in developing a prototype of one such software framework [@ready4fw2024].
+
+<!-- Development of software frameworks to support ethical CHEM implementations could enable and enhance each of these strategies. -->
 
@@
 -->
-Reducing waste in research is a core responsibility of research funders [@chalmers2014increase] and funding the development of CHEMs that are not adequately understood, reused or updated is wasteful.
-Existing incentive structures for health economists generally do not promote facilitating peers to reuse their work and funding is rarely adequate to maintain and routinely update a model.
-Currently, it can take “an extraordinary amount of idealism” to commit to authoring and maintaining research software [@anzt2020environment].
-Previously recommended strategies for more beneficial health economic research investments include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019].
-
-<!-- Development of software frameworks to support ethical CHEM implementations could enable and enhance each of these strategies. --> ????
 

### 2024-02-15 Markdown/Manuscript/Child_RMDs/MS_Imp_Guidelines.Rmd
for coauthor feedback draft

@@
 <!-- support  -->
-(iii) *socially beneficial use* (of CHEM releases).
+(iii) *socially beneficial use* (post CHEM release).
 
 Boden and McKendrick’s transparency criterion underpins the ethical goals we propose for all three phases (Table \@ref(tab:timelygls)).
-Boden and McKendrick’s other <!-- ethical  -->criteria can be selectively mapped to goals for specific phases: our proposed goal for ethical CHEM planning incorporates justice and independence considerations; the goal proposed for CHEM implementation is based on independence and beneficence / non-maleficence criteria; and the goal for ethical CHEM releases reflects beneficence / non-maleficence and justice criteria.
+Boden and McKendrick’s other <!-- ethical  -->criteria can be selectively mapped to specific phases: our proposed goal for ethical CHEM planning incorporates justice and independence considerations; the goal we propose for CHEM implementation is based on independence and beneficence / non-maleficence criteria; and our proposed ethical goal post CHEM release reflects beneficence / non-maleficence and justice criteria.
 <!-- /Edit -->
@@
 These value judgments influence the assumptions, selection of model features and standards for evidence that shape health economic model projects [@HARVARD2020112975].
-For example, the process of determining CHEM *user requirements and design specifications* will involve making judgments about the competing claims of alternative software development platforms (Table \@ref(tab:timelygls)).
+For example, the process of determining CHEM *user requirements and design specifications* will involve making judgments about the competing claims of alternative software development platforms (Table \@ref(tab:proscons)).
 That decision will in turn determine how straightforward it is for a CHEM to be subsequently amended to reflect alternative value judgments relating to model conceptualisation and mathematical formalisation (Table \@ref(tab:timelygls)).
@@
 <!-- Moved -->
-<!-- Edit --><!-- All three  -->Our proposed ethical <!-- responsibilities  --> goals are most likely to be realised when there is both good<!-- have implications for both  -->professional practice by CHEM developers and <!-- Addition -->adequate <!-- Addition -->resourcing by project funders.
+<!-- Edit --><!-- All three  -->Our proposed ethical goals are most likely to be met when there is both good<!-- have implications for both  --> professional practice by CHEM developers and <!-- Addition -->adequate <!-- Addition -->resourcing by project funders.
 <!-- /Edit --> The choices CHEM developers make often need to balance ethical considerations with project resource constraints (as illustrated in Table \@ref(tab:proscons)).
-Modelling projects implemented by diligent and conscientious CHEM developers may still be exposed to ethical risks if there are inadequate resources to meaningfully engage relevant stakeholders, prepare code and data artefacts for independent assessment and provide ongoing model maintenance and user support.
+Modelling projects implemented by diligent and conscientious CHEM developers may still be exposed to ethical risks if there are inadequate resources to meaningfully engage important stakeholder groups, prepare code and data artefacts for independent assessment and provide ongoing model maintenance and user support.
+Existing incentive structures for research health economists generally do not promote facilitating peers to reuse their work and funding is rarely adequate to maintain and routinely update a model.
+Currently, it can take “an extraordinary amount of idealism” to commit to authoring and maintaining research software [@anzt2020environment].
 For these reasons, we believe responsibility for achieving our proposed ethical goals is shared between CHEM developers and funders.
@@
 -->
+
 <!-- For example, as illustrated in Table \@ref(tab:proscons), project resource constraints may require model developers to make trade-offs -->
@@
 <!-- The responsibilities of model project teams over a CHEM lifecycle  -->
-Our proposed ethical goals for each project phase are are easier to state than to measure.
+Our proposed ethical goals for each project phase are easier to state than to measure.
 However, <!-- aspects of ethical modelling practice  -->achievement of these goals may be inferred from measurable attributes of CHEMs.
@@
 
-Taking steps to make computational models *transparent* is an important means of reducing ethical risk<!-- part of ethical public health modelling practice  -->[@10.3389/fpubh.2017.00068].
+Taking steps to make computational models *transparent* is an important means of reducing ethical risk <!-- part of ethical public health modelling practice  -->[@10.3389/fpubh.2017.00068].
 Guidance on transparency in health economic modelling recommend that model code and data should be clearly documented [@Eddy2012].
@@
 Modular implementations enable models to be constructed from multiple independently reusable and replaceable sub-models (modules) [@pan2021modular].
-The programming concept inheritance [@8717448], deployed in object-oriented programming approaches, can facilitate the duplication and selective modification of models.
+The programming concept inheritance [@8717448], deployed in object-oriented programming approaches, can facilitate the selective reuse and modification of model components.
 Granting permissions to others to use, test and adapt models and their components, can be facilitated by two broad categories of open source licenses.
@@
 <!-- # -->
-#	Criteria for assessing ethical computational implementations
-For each CHEM attribute that suggests ethical CHEM modelling practice, we identified assessment criteria.
+#	Criteria for assessing TRU attributes
+We suggest six criteria (two for each attribute) against which for assessing the TRU attributes of CHEMs.
 

### 2024-02-15 Markdown/Manuscript/Parent_PDF/Parent_PDF.Rmd
for coauthor feedback draft

---
bibliography: ../Data/references.bib
output: 
  bookdown::pdf_book: 
    base_format: rticles::springer_article # rticles::arxiv_article
    includes:
      in_header: preamble.tex
    keep_tex: true
    pandoc_args: 
      - "--metadata-file=../Header/header_common.yaml"
      - '--lua-filter=../Filters/scholarly-metadata.lua'
      - '--lua-filter=../Filters/author-info-blocks.lua'
  # bookdown::pdf_document2: 
  #   includes:
  #     in_header: preamble.tex
  #   pandoc_args: 
  #     - "--metadata-file=../Header/header_common.yaml"
  #     - '--lua-filter=../Filters/scholarly-metadata.lua'
  #     - '--lua-filter=../Filters/author-info-blocks.lua'
  #   toc: no
  #   keep_tex: true

params:
  X: NULL
abstract: "TO BE REPLACED"
# abstract: "`r paste0(readLines('../Header/abstract.txt'),collapse=' ')`"
graphics: yes
#csl: plos.csl
title: REPLACE ME
titlerunning: TEMPORARY
authorrunning: Hamilton, Gao, Karnon, et al.
thanks: | 
    The study was funded by ADD DETAILS HERE. 
keywords: one, two, three, four, five, six
#PACS: 
#- PAC1
#- superPAC

# MSC:
# - MSC code 1
# - MSC code 2
classoptions: 
  - sn-basic      
  # - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
  # - referee       # Optional: Use double line spacing 
  # - lineno        # Optional: Add line numbers
  # - iicol         # Optional: Double column layour
#biblio-style: spbasic
# bibstyle options spbasic(default), spphys, spmpsci
# header-includes:
# - \usepackage{booktabs}
# - \usepackage{longtable}
# - \usepackage{array}
# - \usepackage{multirow}
# - \usepackage{wrapfig}
# - \usepackage{float}
# - \usepackage{colortbl}
# - \usepackage{pdflscape}
# - \usepackage{tabu}
# - \usepackage{threeparttable}
# - \usepackage[normalem]{ulem}
# - \usepackage{threeparttablex}
# - \usepackage{makecell}
---

```{r echo = F}
## NOTE: EDITED TEMPLATE FILE To Add rotating and multirow packages as per:
# https://stackoverflow.com/questions/31503418/how-to-add-a-latex-package-to-r-journal-template-from-rticle-package-in-r
library(ready4show)
options(tinytex.verbose = TRUE)
library(xtable)
options(xtable.floating = FALSE)
options(xtable.timestamp = "")
```

```{r child="../Child_RMDs/All_Content.Rmd", echo=FALSE}
```

### 2024-02-15 Markdown/Manuscript/Parent_PDF/Springer.Rmd
for coauthor feedback draft

@@
   #   affiliation: [6,3]
+  - firstname: Caroline
+    lastname: Gao
+    #     email: caroline.gao@orygen.org.au
+    affiliation: [2,4,1]
 
@@
     affiliation: [3]
-
-  - firstname: Caroline
-    lastname: Gao
-    #     email: caroline.gao@orygen.org.au
-    affiliation: [2,4,1]
 

### 2024-02-15 setup.R

@@ -38,12 +38,13 @@ X <- renewSlot(X,
                {
                  args_ls <- X@abstract_args_ls
                  args_ls$abstract_ls <- list(Summary = "Most health economic analyses are undertaken with the aid of computers. However, the ethical dimensions of implementing health economic models as software (or computational health economic models (CHEMs)) are poorly understood. 
-                                                        We propose that developers and funders of CHEMs should pursue the ethical goals of (i) socially acceptable user requirements and design specifications; (ii) fitness for purpose; and (iii) socially beneficial use. 
+                                                        We propose that developers and funders of CHEMs should pursue the ethical goals of: (i) socially acceptable user requirements and design specifications; (ii) fit for purpose implementations; and (iii) socially beneficial post-release use. 
                                                         We further propose that a transparent (T), reusable (R) and updatable (U) CHEM is suggestive of a project team that has largely met these goals.
                                                         We propose six criteria for assessing TRU CHEMs: (T1) software files are open access; (T2) project team contributions and judgments are easily identified; 
                                                         (R1) programming practices promote selective reuse; (R2) licenses restrict only unethical reuse; (U1) maintenance infrastructure is in place; and 
                                                         (U2) new releases are systematically retested and appropriately deprecated.
-                                                        Our assessment criteria can help inform and improve ethical implementation of CHEMs. Resource barriers to ethical CHEM practice should be addressed by research funders."
+                                                        Few existing CHEMs would meet all TRU criteria. Addressing these limitations will require changes to the work practices, funding and tools that underpin health economic model development.
+                                                        Priority actions include the development of new and updated good practice guidelines and research funding to support activities that help make CHEMs more easily understood, widely reused and kept up to date."
                                              # ,
                                              # Objectives = "We aimed to develop a conceptual and technical framework for implementing open source mental health systems models and illustrate how it can be applicatied to economic topics in youth mental health.",
                                              # Methods = "We created a software development kit in R to support standardised implementation of mental health systems models. As a worked example, we applied the toolkit to develop open source youth mental health utility mapping models, which we assessed against five principles using 17 standards.",

### 2024-03-19 Markdown/Manuscript/Child_RMDs/Clippings.Rmd
v4

@@
 <!-- We have designed this framework to enable ethical computational implementations of health economic models that are both open-source (for transparency and reusability) and modular (for reusability and updatability).
-The framework aims to support the standardization and interoperability that are core enablers of implementing a modular and open source approaches.
+The framework aims to support the standardization and interoperability that are core enablers of implementing a modular and open-source approaches.
 -->
@@
 Modular models provide an opportunity for multiple modelling teams to contribute to, test and reuse models.
-To enhance this capacity, modular models may be implemented as open source projects that give others liberal permissions to access and use model source code and data [@pan2021modular; @barros2023empowering; @frazier2022robust].
-Although there appears to be in principle support from many health economists for greater use of open source health economic models (OSHEMs)  [@Pouwels2022], actual implementations are rare [@Emerson2019; @Michalczyk2018; @Feenstra2022].
-Barriers to adoption of OSHEMs include concerns about intellectual property, confidentiality, model misuse and the resources required to support open source implementations [@Pouwels2022; @Wu2019].
+To enhance this capacity, modular models may be implemented as open-source projects that give others liberal permissions to access and use model source code and data [@pan2021modular; @barros2023empowering; @frazier2022robust].
+Although there appears to be in principle support from many health economists for greater use of open-source health economic models (OSHEMs)  [@Pouwels2022], actual implementations are rare [@Emerson2019; @Michalczyk2018; @Feenstra2022].
+Barriers to adoption of OSHEMs include concerns about intellectual property, confidentiality, model misuse and the resources required to support open-source implementations [@Pouwels2022; @Wu2019].
 Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020], but guidelines for implementing OSHEMs remain scarce, piecemeal and need improving [@Sampson2019].
 
-Our interest in modular and open source approaches developed when we began seeking am appropriate framework for undertaking and validly synthesising diverse types of economic research in mental health.
+Our interest in modular and open-source approaches developed when we began seeking am appropriate framework for undertaking and validly synthesising diverse types of economic research in mental health.
 Mental disorders impose high health, social and economic burdens worldwide [@RN8;@GBD2019].
@@
 
-Standardisation is an essential enabler of automation, collaboration, interoperability and transferability in modular and open source approaches.
+Standardisation is an essential enabler of automation, collaboration, interoperability and transferability in modular and open-source approaches.
 An early requirement to implement ready4 is therefore to identify an appropriate set of standards.
@@
 To make model code and data widely re-usable by others, it is important to provide users with appropriate and explicit permissions.
-In the context of open source models, there are two broad categories of licensing options.
+In the context of open-source models, there are two broad categories of licensing options.
 Some guidance strongly recommends the use of permissive licensing [@Wilson_2017] that provides users with great flexibility as to the purposes (including commercial) for which the content could be re-used.
-An alternative approach is to use copyleft licenses [@copyleft2022] that can require content users to distribute any derivative works they create under similar open source arrangements.
-For code, it may be appropriate to adopt the prevailing open source licensing practice within the programming language being used.
+An alternative approach is to use copyleft licenses [@copyleft2022] that can require content users to distribute any derivative works they create under similar open-source arrangements.
+For code, it may be appropriate to adopt the prevailing open-source licensing practice within the programming language being used.
 For data, it may not be sufficient to simply choose between a permissive license like the Public Domain Dedication (CC0) [@cc02022] or a copyleft option such as the Attribution-Share Alike (CC-BY-SA) [@bysa2022].
@@
 Statistical models are a common output of health economic evaluations, but they are often not reported in a format that enables others to confidently and reliably re-use them for out of sample prediction [@Kearns2013].
-Open source approaches can help address this by disseminating code artefacts that enable easy and appropriate use of a statistical model to make predictions with new data.
+open-source approaches can help address this by disseminating code artefacts that enable easy and appropriate use of a statistical model to make predictions with new data.
 However, great care must be exercised when publicly releasing model artefacts derived from data on human subjects as they may by default embed a copy of the source dataset.
@@
 Without reuse, errors in model artefacts are more likely to remain undetected.
-Even when errors are detected, they still need to be fixed, but maintaining code and data requires ongoing resourcing through a combination of centralised infrastructure and an active open source community.
+Even when errors are detected, they still need to be fixed, but maintaining code and data requires ongoing resourcing through a combination of centralised infrastructure and an active open-source community.
 
@@
 ## Implications for economic modelling in mental health
-Open source approaches have been recommended to help develop the mental health modelling field [@RN73] but only one mental health related model (in Alcohol Use Disorder [@Basu2018]) is currently indexed in the Open Source Models Clearinghouse [@OSMC_20xx;@Emerson2019].
-We are aware of just one other open source mental health model - a reference model in Major Depressive Disorder - that is currently in development [@IVIMDD2022].
-Of the known barriers to adoption of open source models by health economists (including issues like intellectual property and confidentiality [@Pouwels2022; @Wu2019]), our experience suggests that the biggest challenges may be the enormous effort required to first prepare model code and data for public release in formats that facilitate appropriate reuse by third parties and to then maintain and continually improve potentially large numbers of digital artefacts.
+open-source approaches have been recommended to help develop the mental health modelling field [@RN73] but only one mental health related model (in Alcohol Use Disorder [@Basu2018]) is currently indexed in the open-source Models Clearinghouse [@OSMC_20xx;@Emerson2019].
+We are aware of just one other open-source mental health model - a reference model in Major Depressive Disorder - that is currently in development [@IVIMDD2022].
+Of the known barriers to adoption of open-source models by health economists (including issues like intellectual property and confidentiality [@Pouwels2022; @Wu2019]), our experience suggests that the biggest challenges may be the enormous effort required to first prepare model code and data for public release in formats that facilitate appropriate reuse by third parties and to then maintain and continually improve potentially large numbers of digital artefacts.
 
 Automated tools such as those we developed in our modelling toolkit can help reduce the burden associated with some of these tasks.
-However, we think the current low rates of adoption by health economists of open source approaches [@Emerson2019; @Michalczyk2018; @Feenstra2022] (which in turn facilitate the collaboration that make modular models more attractive) will only change slowly unless there is significant and strategic investments made by research funders.
+However, we think the current low rates of adoption by health economists of open-source approaches [@Emerson2019; @Michalczyk2018; @Feenstra2022] (which in turn facilitate the collaboration that make modular models more attractive) will only change slowly unless there is significant and strategic investments made by research funders.
 Currently, incentive structures for health economists do not promote the dedication of large quantities of time to enable peers to reuse their work.<!-- MH Research priorities (WP3 - Research, WP6 - Economic) https://doi.org/10.1016/S2215-0366(15)00332-6 -->
@@
 The intellectual asset represented by this literature could be enhanced if many of the models described in these reports could be brought and kept up to date and made available in formats that maximised transferability to diverse decision contexts.
-We believe that modular and open source approaches would be well suited to accomplishing this goal and that the framework we have developed could act as an early prototype for solving some of the technical challenges of this task.
+We believe that modular and open-source approaches would be well suited to accomplishing this goal and that the framework we have developed could act as an early prototype for solving some of the technical challenges of this task.
 Ideally such a program of research would be resourced to be sustained over the medium to long term and to engage a diverse network of investigators, contributors and advisers from high, middle and low income countries.
@@
 Ideally health economists would explore these complex topics in partnership with modellers from other disciplines (in particular epidemiology and health services research) and a wide range of stakeholders such as other researchers, policymakers, service planners and community members.
-Modular and open source approaches would facilitate such investigations by breaking down ambitious and long term goals into manageable time-bound discrete projects, each progressed by different teams.
+Modular and open-source approaches would facilitate such investigations by breaking down ambitious and long term goals into manageable time-bound discrete projects, each progressed by different teams.
 To facilitate such an approach, a common framework of standards and tools would be needed.
-To be suitable for such a task our framework would need additional development, with its overall architecture reviewed for scalability and suitability and to provide better integration with and use of other open source languages (particularly python) and repositories.
-Whatever MOSHEM infrastructure is developed, its resilience would depend on a community of open source contributors sufficiently large and active to ensure that all core modules are maintained even after their original authors cease their involvement.
+To be suitable for such a task our framework would need additional development, with its overall architecture reviewed for scalability and suitability and to provide better integration with and use of other open-source languages (particularly python) and repositories.
+Whatever MOSHEM infrastructure is developed, its resilience would depend on a community of open-source contributors sufficiently large and active to ensure that all core modules are maintained even after their original authors cease their involvement.
 
@@
 Not all efforts by health economists to promote MOSHEMs need to depend on the decisions of research funders.
-Releasing selected subsets of unmaintained model artefacts in open source repositories is still better than not providing access to any code and data and can typically be accomplished within existing project budgets.
+Releasing selected subsets of unmaintained model artefacts in open-source repositories is still better than not providing access to any code and data and can typically be accomplished within existing project budgets.
 Developing knowledge and skills of MOSHEMS can be advanced by making small contributions (e.g.
-improvements to documentation, code contributions) to open source projects.
+improvements to documentation, code contributions) to open-source projects.
 Our project website [@rfwn2022] includes details of multiple ways to contribute to ready4.
@@
 <!-- Estimating the potential uptake of these tools is difficult.  -->
-<!-- Open source is more than open access (limited discussion). -->
+<!-- open-source is more than open access (limited discussion). -->
  <!-- A coding framework for OSHEMs developed in the language R includes standardised approaches to directory structure and naming conventions [@Alarid2019]. -->
@@
  <!-- but, as with health economics more generally, OSHEMs remain rare.
-Greater use of open source approaches could help improve the scope, validity and usefulness of mental health economic models.
+Greater use of open-source approaches could help improve the scope, validity and usefulness of mental health economic models.
 -->

### 2024-03-19 Markdown/Manuscript/Child_RMDs/Doodles.Rmd
v4

---
title: "Doodles"
author: "Matthew Hamilton"
date: '2022-04-19'
output: html_document
---

Libraries:
describing and validating input data [@hamilton_matthew_2022_6084467], 
scoring health utility data [@hamilton_matthew_2022_6084824], 
specifying and testing mapping models [@hamilton_matthew_2022_6116701] 
reporting mapping studies [@gao_caroline_2022_6130155].
make predictions with new data using another R package [@matthew_p_hamilton_2021_5646669] that we developed

Programs:
CSP [@hamilton_matthew_2022_6129906]
Predict [@hamilton_matthew_2022_6416330]
Make Fakes [@hamilton_matthew_p_2022_6321821]

Sub-routines:
Ctlg: [@hamilton_matthew_2022_6116385]
SS: [@matthew_p_hamilton_2022_5976988]

Data
results / replication [@DVN/DKDIB0_2021]  
Fake [@DVN/HJXYKQ_2021]

To facilitate appropriate use of the mapping models we developed we created catalogues with detailed information about the performance of all models generated by our study under multiple types of regime (e.g. whether: mean or sampled parameter values are used; minimum and maximum constraints are applied to prediction results; the original R model objects or just tables of model coefficients are used).

10 rules [@Erdemir2020]
.....DARTH

[@Sampson2019] - 
development of networks of health economists that are working on a particular disease area. 
f reference models, open-source modelling, or model registries are to be part of the model development process, then standards need to be developed to guide analysts in achieving best practice. The availability of guidance can help to create confidence in researchers to pursue transparency. Yet, selecting the best standards to use can also be challenging [56]. Wider uptake—and appropriate use—of standards, by both modellers and journals, should be encouraged, and research should evaluate the suitability of guidance relating to transparency.
However, transparency depends on the code being well structured and clearly commented and a user or reviewer is required to have knowledge of the language in order to validate, use or adapt the model. There is a steep learning curve for analysts unfamiliar with script-based models
There may also be barriers in the communication of models due to a lack of knowledge for end users. However, this can be substantially mitigated with the use of open access interface packages, like the R package Shiny, which can be used to turn an R model into an interactive web application [58]. This allows a user without any R knowledge to use or, in part, validate the model.
Collaboration
----
@Erdemir2020] - 
We recommend the use of existing repositories for disseminating code and models.

there is a need for guidance, frameworks and templates
Guidance needed for implementation. Current transparency guidance not good enough.

ome resistence to adoption centers on.... [@Padula2017] . 
HE commercial...[@Feenstra2022],
   author]

A major shift in how health economic models are implemented.....They may also have educational value in developing the health systems modelling field [@Dunlop2017]. However, open-source computational models also present a range of conceptual, technical, legal and resourcing issues compared to more traditional approaches to model development.

The health economics field i

cooperation [@Arnold2010]
coding framework / guidance [@Alarid2019]
10 rules [@Erdemir2020]
Sys review??

Maybe [@Feenstra2022] - 
Model complexity increases validation needs as does multi-purpose modelling
conceptual model, input data, computational model, operational model
International collaboration in validation
Need for documentation
Commercial ownership of HE models reduce scope for open-source
finances, accountability, and ownership barriers to OS
still not very widespread in HE decision model
would require more resources, could greatly increase transparency
distinguishing grades of open-source
Model validation should hence become more structured and standardized, as well as more transparent.

Modular validation - Ghabri, Justifies extensive validation
Citing Dunlop - HE lagging behind other modelling fields

Australian
poor validation - ghabri

Over the last decade there have been
the potential to increase the transparency, reproducibility and reuse of health economic analyses.

open-source frameworks, which have been recommended for the development of mental health modelling field [@RN73], have the potential to help address a number of these challenges. open-source approaches can promote common standards and interoperability, improve model transparency and facilitate routine updates to rectify errors, update data and add features.

- Our project - readyforwhatsnext : open-source model of the systems shaping the mental health of young people
- Lack of guidance on open-source health economic models and in particular on recommended frameworks for implementing the computational model.
- Focus computational model. Most in need of novel guidance. Also: Of the three components that comprise a mental health computer simulation model - the conceptual model, the mathematical model and the computational model - only the first two are typically developed and reported in a manner that facilitates reuse, refinement and extension by others. This is problematic as the code implementation is where the theory represented by a simulation model is most fully elaborated.

ISPOR T & V: 
a multiapplication model, transparency, validation,
For a multiapplication model, transparency, validation,
and reporting are ongoing processes. The multiapplica-
tion model is described (transparency)24 and validated,25
and the descriptions and validations are continuously
updated as science and the model evolve.26 In addition,
each instantiation of the model is described, validated,
and reported as each application is done.27

CLIPPINGS
It is frequently claimed that better approaches to the organising and financing of mental health service systems can generate significant gains in population health (Saxena, 2007; Chisholm, 2016).  Over the last two decades, policymakers in Australia have acted on these claims by investing in a series of major systemic mental health reforms, with improving the mental health of young people a major focus (McGorry, 2017). The scope and complexity of these reforms has increased over time, with the Victorian Government’s current plan to overhaul mental health service delivery (Victoria, 2021) arguably the most ambitious to-date.

 Although Australian mental health policy and system design processes has a formally specified role for a robust model-based decision aid, this tool,  the National Mental Health Services Planning Framework, has limited capacity to meet many of the core decision-support requirements of systems planners (Commission, 2020). These requirements are diverse and extensive, but are likely to include models that can:

•	anticipate the distribution and evolution of risk factors and need for mental health care by region and population group;

•	identify the budget, equity and efficiency impacts of alternative resource allocations between mental health programs;

•	provide guidance about how services can be optimally configured to achieve outcomes such as minimising wait times, maximising health gain, meeting patient preferences and improving societal welfare; and

•	predict how the supply, location and behaviour of mental healthcare providers change in response to alternative incentives and conditions.

Although a number of these or similar types of models have been developed for the Australian context, the validity, utility and accessibility of these models vary considerably.

and to anticipate and mitigate risks to the successful implementation of these reforms

Developing theories that produce valid and useful predictions about the potential impacts and feasibility of phased, multi-component programs of policy and service system reform is most likely a major and long term challenge. This undertaking might become more tractable if it could be chunked into smaller tasks and progressed by a research community that, even when working independently, routinely test, re-use and extend each other's work.

However, developing  This leads to computational model issue. This leads to open-source.

The absence of common standards and taxonomies make it hard to develop these more extensive models by simply combining existing mental health models.

 which can be used to generate credible predictions about how these systems might respond to phased, multi-component redesign strategies

Common standards and frameworks for how computational models are developed could address some of these challenges [@RN73].

Outputs from discrete projects should be reproducible, generalisable and able to be linked like Lego bricks to progressively assemble more complex and realistic representations of the systems being modelled.

 Similarly,

These knowledge gaps may partly explain why debates within the mental health policymaking community extend beyond normative issues (what should policy aim to achieve) to technical questions (which system financing and organising choices are most likely to achieve specified objectives). A more mature mental health systems modelling field might see the technical aspects of re-engineering mental health systems subject to the same stringent scientific standards as those required for the development of therapeutic goods. Yet,

If these issues can be successfully addressed, there are potential benefits to both decision makers (who will have access to more relevant and trustworthy model-based decision aids) and modellers (for whom complex models will become more tractable through improved access to relevant code and data).

Access to some model-based decision aids and their constituent algorithms and data is highly restricted. The relevance of models to current and future decision contexts can also be limited as mental health models are rarely routinely updated with new data and improved algorithms.

In addition to issues relating to improving the quality and usefulness of models that address narrowly defined topics, there are additional challenges concerning the development of models with much broader scope.

A number of high income countries have begun to increase public expenditures on mental health services[@RN26], and in some cases have embraced ambitious programs of major structural reforms to mental health service systems[REFSGOHERE]. However

In the limited cases where code is made publicly available it is typically in the form of programs (an algorithm that performs one task) rather than code libraries (where the primary algorithm is decomponsed into multiple component algorithms, that can be re-used for a potentially large number of other purposes).

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
v4

--- 2024-01-24 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
@@
 ---
-
-In this paper, we <!--  MOVED_EDITED_START -->consider literature on modelling practice and draw on our prior experience with developing and using CHEMs to propose<!--  MOVED_EDITED_START -->:  (i) a set of ethical responsibilities for CHEM project teams<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.
+<!-- Important rewordings --> 
+In this paper, we <!--  MOVED_EDITED_START -->draw on our prior experience with developing and using CHEMs and consider literature on modelling practice to propose<!--  MOVED_EDITED_END -->:  (i) a set of research ethics <!--  EDITED_Start responsibilities-->goals<!--  _EDITED_END --> for CHEM <!--  EDIT -->developers and funders<!--  /EDIT --<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that suggest achievement of these goals; and (iii) criteria against which these attributes can be assessed.
 <!--  MOVED_EDITED_END -->

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Cndl_Tabs_Not_In_Body.Rmd
v4

@@
 
-<!-- # ```{r, child = '../Child_RMDs/MS_Tables_3.Rmd', echo=FALSE} -->
-<!-- # ``` -->
-<!-- # \newpage -->
+```{r, child = '../Child_RMDs/MS_Tables_3.Rmd', echo=FALSE}
+```
+\newpage
 <!-- # ```{r, child = '../Child_RMDs/MS_Tables_4.Rmd', echo=FALSE} -->

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Conclusion.Rmd
v4

@@
 <!-- Barriers to ethical CHEM practice arising from resource constraints should be addressed by research funders. -->
+ We propose that CHEM developers and funders adopt the ethical goals of socially acceptable user requirements and design specifications, fit for purpose implementations and socially beneficial post-release use.
+We suggest that CHEM attributes of transparency, reusability and updatability are indicative of projects that meet these goals and have identified six criteria for assessing these attributes.
+We hope our proposals contribute to a timely and much needed debate within the health economics community about the ethical dimensions of implementing health economic models as software.

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Discussion.Rmd
v4

@@
 
-Ethical practice is a core expectation of health researchers and computational methods underpin most quantitative research.
-However, an understanding of the ethical dimensions about how health economic models are implemented as software is underdeveloped.
-We have attempted to address this gap by proposing that CHEM developers and funders adopt the ethical goals of socially acceptable user requirements and design specifications, fit for purpose implementations and socially beneficial post-release use.
-We suggest that CHEM attributes of transparency, reusability and updatability are indicative of projects that meet these goals and have identified six criteria for assessing these attributes.
+Prior literature has noted that health economic models routinely have limitations in transparency [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], reuse [@Feenstra2022; @Emerson2019], and being kept up to date [@Sampson2017; @10.3389/fpubh.2022.899874].
+These limitations not only pose ethical risks such as those we have described, but are practical barriers to achieving potential exciting enhancements to health economic modelling practice such as improved model transferability [@munthe2019systematic; @barbieri2010international], greater use of reference models [@Afzali2013], living health technology assessments [@thokala2023living] and leveraging the analytic capabilities of machine learning [@padula2022machine].
 
-Prior literature has noted that health economic models routinely have limitations in transparency [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], reuse [@Feenstra2022; @Emerson2019], and being kept up to date [@Sampson2017; @10.3389/fpubh.2022.899874].
-These limitations not only pose ethical risks such as those we have described, but are practical barriers to model generalisability and transferability.
-Addressing these limitations will require changes to the work practices, funding and tools that underpin health economic model development.
+It is important that the health economics community engages with the ethical dimensions of how health economic models are implemented as software.
+Recent developments in Large Language Models (LLMs) mean software developers are confronted with novel ethical issues including: increased risk of security flaws; a potential need to add restrictions to open-source licensing; environmentally harmful levels of computing power; and concerns about model explainability and unintended consequences [@ozkaya2023application].
+Early work exploring the use of LLMs in CHEM development has already been undertaken [@poirrier2023msr26].
+Many of the active ethics debates in the computer science community are therefore likely to become increasing relevant to health economists and public health researchers.
+Documents such as the World Health Organisation's recent guidance on generative artificial intelligence [@world2024ethics] will be important reference points for how the health economics discipline navigates emerging ethical themes in CHEM development.
 
-Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020] and there is a need for better and comprehensive guidelines about how health economists should implement health economic models as software.
-Research funders need to actively support initiatives to help remove barriers to better practice in this area.
-A core responsibility of research funders is reducing waste in research [@chalmers2014increase].
+Meeting all six TRU criteria we propose is more likely when a CHEM has been implemented as an open-source project<!-- and model transferability -->.
+It should be noted that the ethical dimensions of open-source software development are broader than the issues we have discussed here, including issues relating to social contracts, contributor autonomy and public goods [@grodzinsky2003ethical].
+There are also important barriers to growing the (currently paltry) number of open-source CHEMs.
+These barriers include health economist concerns about intellectual property, confidential data, the risk of model misuse and the resources required for model maintenance [@Pouwels2022; @Wu2019].
+Commercial considerations, for example when health economic models are owned by pharmaceutical companies and consultancies, may also limit the public availability of models and their constituent code and data [@Feenstra2022].
+Other barriers relate to <!-- model generalisability and transferability and  -->the software platforms in which CHEMs are authored [@Pouwels2022].
+
+Recent efforts to develop the open-source CHEMs field include an initiative to develop a suite of open source CHEMs addressing different health conditions in the United States [@Jansen2019] and a Canadian platform facilitating access to and validation of multiple CHEMs [@harvard2022developing].
+It has also been recommended that training curricula for current and future health economists need to develop some foundational software development skills [@kindilien2021role].
+Education efforts to develop such skills could be reinforced with best-practice scientific computing guidance for health economists.
+Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020], but guidelines for implementing open source CHEMs remain scarce, piecemeal and need improving [@Sampson2019].
+Health economists also need to be able to access reliable and up to date legal advice about the implications of alternative approaches to software licensing.
+
+<!-- Move to discussion--><!-- In health economics the related and enabling concept of reference models has been recommended [@Afzali2013].  -->
+<!-- Transferring a health economic model developed for one jurisdiction for application in another typically involves retaining some features and updating others [@barbieri2010international].
+However, models do not always make implementation choices (for example, use of concepts with standardised meanings across jurisdictions)<!-- [@garcia2021cost] -->  <!-- facilitate such selective re-use.
+-->
+<!-- Open-source CHEMs remain rare [@Feenstra2022; @Emerson2019].
+The software used to develop CHEMs was identified as a barrier to open-source model implementations by participants in a 2020 survey of health economists [@Pouwels2022].
+Other reported barriers relate to difficulties in generalising models, updating models and transferring model data, concerns about the level of public access and lack of interest from decision-makers, and legal, confidentiality and security considerations [@Pouwels2022].
+-->
+
+Many of these barriers require significant investments by governments and other research funders in enabling infrastructure and human capital.
+<!-- A core responsibility of research funders is reducing waste in research [@chalmers2014increase].
 Investments in modelling projects might generate greater social returns if they supported the project activities necessary to develop CHEMs that are easily understood, widely reused and well maintained.
+--> Investment strategies that have previously been recommended for development of the health economic research field include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open-source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019].
 
-Previously recommended strategies for more beneficial health economic research investments include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019].
-Each of these strategies could benefit from enabling technical infrastructure, for example software frameworks that support the development of CHEMs that meet ethical criteria.
+<!-- Each of these strategies could benefit from enabling technical infrastructure, for example software frameworks that support the development of CHEMs that meet ethical criteria.
 We are currently engaged in developing a prototype of one such software framework [@ready4fw2024].
+-->
+
+<!-- Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020] and there is a need for better and comprehensive guidelines about how health economists should implement health economic models as software.
+-->
 
@@
 <!-- However, our prototype has a number of features that future work on CHEM software frameworks may find useful to incorporate.
-Firstly, developing a software framework to work within an existing and widely used open source programming language such as R, can keep framework scope relatively narrow.
+Firstly, developing a software framework to work within an existing and widely used open-source programming language such as R, can keep framework scope relatively narrow.
 This makes it more tractable to develop, maintain and learn, and facilitates integration with other modelling tools written in that language (e.g., the dependency libraries we list in Table \@ref(tab:cpkgs)).
@@
 
-<!-- Our software framework provides the outline of a technical solution to some of the challenges of open source CHEM development.
+<!-- Our software framework provides the outline of a technical solution to some of the challenges of open-source CHEM development.
 However, better tools alone are insufficient and policy and institutional changes are also required.

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Imp_Guidelines.Rmd
v4

@@
 <!-- responsibilities to goals -->
-# Ethical goals for CHEM developers and funders
-Boden and McKendrick [@10.3389/fpubh.2017.00068] propose a framework for ethical public health modelling based on the criteria of independence (concerning how modeller subjectivity shapes model design), beneficence / non-maleficence (concerning model quality and utility), transparency (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and justice (concerning the social obligations of modellers to consider and communicate ethical issues about model use).
-<!-- However, t -->These criteria are not specific to the computational model, but <!-- are designed to  -->apply to a modelling project as a whole, including a model's conceptual and mathematical representations.
-The authors propose 13 questions to evaluate ethical risk across the four criteria, but of these, only "is the model code open source or available on request" (for transparency) and "has the model been verified, i.e., does it do what the modeller wants it to do?" (for beneficence / non-maleficence) are specific to model's representation as software.
+# Research ethics goals for CHEM developers and funders
+Resnik [@resnik2020ethics] identified eighteen ethical principles that underpin influential codes of research good practice.
+We suggest that six of these principles are particularly relevant to CHEM implementation decisions: *accountability* (researchers declaring and taking responsibility for their specific contributions); *confidentiality* (protecting patient records and other confidential information), *intellectual property* (honoring copyrights and acknowledging contributors), *openness* (sharing project data, tools and resources), *transparency* (disclosing the methods and materials needed to evaluate research) and *social responsibility* (promoting social good).
 
-To identify ethical issues that are specific to CHEMs, we <!-- first  -->considered <!-- Edit -->how Boden and McKendrick's four criteria might apply to <!-- /Edit -->the distinct phases in a software development project.
+Boden and McKendrick [@10.3389/fpubh.2017.00068] propose a framework for ethical public health modelling based on the criteria of *independence* (concerning how modeller subjectivity shapes model design), *beneficence / non-maleficence* (concerning model quality and utility), *transparency* (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and *justice* (concerning the social obligations of modellers to consider and communicate ethical issues about model use).
+<!-- However, t -->This framework is not specific to a model's implementation as software, but <!-- are designed to  -->apply to a modelling project as a whole, including a model's conceptual and mathematical representations.
+Boden and McKendrick proposed 13 questions to evaluate ethical risk across the four criteria, but of these, only two pertain to computational models: "is the model code open-source or available on request" (for transparency) and "has the model been verified, i.e., does it do what the modeller wants it to do?" (for beneficence / non-maleficence).
+
+To identify ethical issues that are specific to CHEMs, we <!-- first  -->considered <!-- Edit -->how Resnick's 18 principles and Boden and McKendrick's four criteria might apply to <!-- /Edit -->the distinct phases in a software development project.
 Software development lifecycle (SDLC) models take different approaches to the inclusion, naming, definition, sequencing and iteration of project phases [@ruparelia2010software].
-However, SDLC models typically have components that can principally map to the concepts of *planning* (e.g., identification of system / user requirements and design specifications), *implementation* (e.g., development and testing) and *release* (e.g., deployment, system integration, maintenance and support).
+However, SDLC models typically have components that can also principally map to the concepts of *planning* (e.g., identification of system / user requirements and design specifications), *implementation* (e.g., development and testing) and *release* (e.g., deployment, system integration, maintenance and support).
 For simplicity, we propose one overarching ethical <!-- responsibility  for CHEM project team-->goal for each phase.
@@
 
-Boden and McKendrick’s transparency criterion underpins the ethical goals we propose for all three phases (Table \@ref(tab:timelygls)).
-Boden and McKendrick’s other <!-- ethical  -->criteria can be selectively mapped to specific phases: our proposed goal for ethical CHEM planning incorporates justice and independence considerations; the goal we propose for CHEM implementation is based on independence and beneficence / non-maleficence criteria; and our proposed ethical goal post CHEM release reflects beneficence / non-maleficence and justice criteria.
-<!-- /Edit -->
+Transparency (as a Resnik principle and Boden and McKendrick criterion) and Resnik's social responsibility principle underpin all three of the ethical goals we propose (Table \@ref(tab:timelygls)).
+Additionally: <!-- Boden and McKendrick’s other <!-- ethical  --><!-- criteria can be selectively mapped to specific goals:  -->(i) the social acceptability goal incorporates Resnik's accountability, confidentiality and intellectual property principles and Boden and McKendrick's justice and independence criteria;
+(ii) the fitness for purpose goal relates to Resnik's accountability principle and Boden and McKendrick's independence and beneficence / non-maleficence criteria; and 
+(iii) the socially beneficial use goal reflects Resnik's openness principle and Boden and McKendrick's beneficence / non-maleficence and justice criteria. <!-- /Edit -->
 
@@
 These value judgments influence the assumptions, selection of model features and standards for evidence that shape health economic model projects [@HARVARD2020112975].
-For example, the process of determining CHEM *user requirements and design specifications* will involve making judgments about the competing claims of alternative software development platforms (Table \@ref(tab:proscons)).
+For example, the process of determining CHEM *user requirements and design specifications* will involve making judgments about the competing claims of open-source and commercial development platforms (Table \@ref(tab:proscons)).
 That decision will in turn determine how straightforward it is for a CHEM to be subsequently amended to reflect alternative value judgments relating to model conceptualisation and mathematical formalisation (Table \@ref(tab:timelygls)).
@@
 
-Modellers need to do more than self-certify that their <!-- have duties both to ensure a  -->computational model is *fit for purpose* - <!-- Edit -->they also need to<!-- /Edit --> provide potential third-party users with the means of assessing its adequacy
-<!-- for their proposed uses  -->[@Erdemir2020; @Feenstra2022; @thompson2019escape; @10.3389/fpubh.2017.00068].
+Modellers need to do more than self-certify that their <!-- have duties both to ensure a  -->computational model is *fit for purpose* - <!-- Edit -->they also need to<!-- /Edit --> provide potential third-party users with the means of assessing its adequacy, appropriateness and usefulness <!-- for their proposed uses  -->[@Erdemir2020; @Feenstra2022; @thompson2019escape; @10.3389/fpubh.2017.00068].
 However, it is common for health economic models to have serious methodological flaws [@carletto_zanuzzi_sammarco_russo_2020; @WONDER2015467]; insufficient validation [@Ghabri2019; @kolovos2017model; @haji2013model], poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017]; and undeclared errors [@Radeva2020].
-Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and <!-- Edit -->providing executable code for  <!--facilitating manual  -->review and testing <!-- /Edit -->by third parties.
+Appropriate computational implementation choices can help address many of these shortcomings, for example by automating and publicly documenting quality assurance checks and <!-- Edit -->providing executable code for  <!--facilitating manual  -->review and testing <!-- /Edit -->by third parties.
 
@@
 Currently, health economic models are rarely implemented computationally in a manner that facilitates routine updates [@Sampson2017], thus limiting the temporal window within which a CHEM can be validly applied.
+
 <!-- Moved -->
@@
 Modelling projects implemented by diligent and conscientious CHEM developers may still be exposed to ethical risks if there are inadequate resources to meaningfully engage important stakeholder groups, prepare code and data artefacts for independent assessment and provide ongoing model maintenance and user support.
+Currently, it can take “an extraordinary amount of idealism” to commit to authoring and maintaining research software [@anzt2020environment].
 Existing incentive structures for research health economists generally do not promote facilitating peers to reuse their work and funding is rarely adequate to maintain and routinely update a model.
-Currently, it can take “an extraordinary amount of idealism” to commit to authoring and maintaining research software [@anzt2020environment].
+Economists may therefore make pragmatic judgments that proprietary licensing approaches provide the most feasible pathway to mobilising the resources to support model maintenance.
 For these reasons, we believe responsibility for achieving our proposed ethical goals is shared between CHEM developers and funders.
@@
 Guidance on transparency in health economic modelling recommend that model code and data should be clearly documented [@Eddy2012].
-Notably, the same guidelines, published over ten years ago, do not recommend sharing model code and data.
+These transparency guidelines, published twelve years ago, did not recommend sharing model code and data.
 However, more recent healthcare modelling guidance does recommend public dissemination of such artefacts [@Erdemir2020].
@@
 The extensiveness of such verification checks can be reported using the concept of code coverage [@ERICWONG2010188] - the proportion of model code that has been explicitly tested.
-Tests should ideally combine both unit tests (to verify that small, isolated sections of code produce the correct output when run independently) and acceptance tests (to verify that the correct output is produced when multiple code components are run together to perform tasks that meet core user-requirements [@martin2003agile]).
-The nature and extent of individual model authorship contributions can become unclear when models are implemented over longer time-frames with a large and changing group of collaborators [@thompson2022escape].
-This issue can be addressed by use of online repository services such as GitHub [@github2007], that provide citation tools and can transparently record all individual code contributions to a modelling project over its lifecycle.
+<!-- Tests should ideally combine both unit tests (to verify that small, isolated sections of code produce the correct output when run independently) and acceptance tests (to verify that the correct output is produced when multiple code components are run together to perform tasks that meet core user-requirements [@martin2003agile]).
+-->Testing procedures and results can be publicly shared through online code development services like GitHub [@github2007].
+GitHub also provides citation tools and can transparently record all individual code contributions to a modelling project over its lifecycle.
+These tools may be particularly useful in complex modelling projects implemented over longer time-frames with a large and changing group of collaborators, where the nature and extent of individual model authorship contributions can become unclear [@thompson2022escape].
 
@@
 Although making a CHEM’s code, data and documentation publicly available is increasingly considered good practice, it is insufficient to ensure that a model is *reusable*.
-Key concerns for health economists when considering whether to reuse a model are generalisability (application without adaptation) and transferability (selective reuse and/or modification of model components) [@RN39].
-Writing model algorithms as collections of functions (short, self-contained and reusable software routines that each perform a discrete task) is good scientific computing practice [@Wilson_2017] and promotes selective reuse.
+<!-- Key concerns for health economists when considering whether to reuse a model are generalisability (application without adaptation) and transferability (selective reuse and/or modification of model components) [@RN39].
+-->Writing model algorithms as collections of functions (short, self-contained and reusable software routines that each perform a discrete task) has been recommended as good scientific computing practice [@Wilson_2017] and promotes reuse of model sub-components.
 Computational implementations that store model code and data in distinct files and locations (as opposed to embedding data such as parameter values into source code) are also easier to selectively modify.
-Modular implementations enable models to be constructed from multiple independently reusable and replaceable sub-models (modules) [@pan2021modular].
-The programming concept inheritance [@8717448], deployed in object-oriented programming approaches, can facilitate the selective reuse and modification of model components.
-Granting permissions to others to use, test and adapt models and their components, can be facilitated by two broad categories of open source licenses.
+Modular implementations enable models to be constructed from replaceable sub-models (modules) [@pan2021modular] that can share inputs and outputs with each other or be run as independent models [@barros2023empowering].
+Modularity can <!-- Advantages of modular models include feasibility (large projects are broken into smaller tasks, with each component independently developed and tested) and flexibility ( -->make it easier to selectively replace or update specific parts of a model and to scale up or down the level of granularity [@pan2021modular].
+<!-- The programming concept inheritance [@8717448], deployed in object-oriented programming approaches, can facilitate the selective reuse and modification of model components.
+-->Granting permissions to others to use, test and adapt models and their components, can be facilitated by two broad categories of open-source licenses.
 Some guidance strongly recommends the use of permissive licensing [@Wilson_2017] that provides users with great flexibility as to the purposes (including commercial) for which content can be re-used.
@@
 Ensuring that a model is regularly reviewed to identify and implement required improvements is a recommended defence against model validity decay [@calder2018computational].
-Sustainable maintenance of open source research software requires both a core development team and an active user community [@info:doi/10.2196/20028].
+Sustainable maintenance of open-source research software requires both a core development team and an active user community [@info:doi/10.2196/20028].
 Online communities can be an efficient means of engaging model users in testing each version of a model, identifying issues and suggesting improvements.
 Services such as GitHub [@github2007] provide collaborative code development tools [@MERGEL2015464] that help elicit, integrate and reconcile contributions from multiple contributors and to ensure each update is uniquely identifiable and retrievable.
-It is also important that verification checks are rerun with each model update, a task that can be automated using the software development practice of continuous integration [@CI2017].
+Verification checks should be rerun with each model update, a task that can be automated using the software development practice of continuous integration [@CI2017].
 The risk of software revisions having unintended consequences for third party users can be mitigated through the use of deprecation conventions [@zhou2016api] that take an informative and staged approach to retiring outdated CHEMs.
@@
 #	Criteria for assessing TRU attributes
-We suggest six criteria (two for each attribute) against which for assessing the TRU attributes of CHEMs.
+We suggest six criteria (two for each attribute) for assessing the TRU attributes of CHEMs. These criteria are summarised in Table \@ref(tab:trucrit).
 
-How *transparent* a CHEM is can be assessed against the criteria:
+```{r, child = child_docs_ls$three, echo=FALSE}
+```
 
--	*T1*: Software files (executable code, testing procedures and outcomes and non-confidential data) are available in open access repositories.
+<!-- How *transparent* a CHEM is can be assessed against the criteria: -->
 
--	*T2*: It is easy to see who funded, developed and tested each part of the CHEM and to identify the modelling team’s assumptions, judgments and theories about CHEM development and use.
+<!-- -	*T1*: Software files (executable code, testing procedures and outcomes and non-confidential data) are publicly available. -->
 
-How *reusable* a CHEM is can be assessed against the criteria:
+<!-- -	*T2*: It is easy to see who funded, developed and tested each part of the CHEM and to identify the modelling team’s assumptions, judgments and theories about CHEM development and use. -->
 
--	*R1*: Programming practices facilitate selective reuse of model sub-components.<!-- Replacement -->
+<!-- How *reusable* a CHEM is can be assessed against the criteria: -->
 
--	*R2*: Licenses restrict only unethical use and allow anyone to ethically reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works.
+<!-- -	*R1*: Programming practices facilitate selective reuse of model sub-components.<!-- Replacement -->
 
-How *updatable* a CHEM is can be assessed against the criteria:
+<!-- -	*R2*: Licenses allow anyone to ethically reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works. -->
 
--	*U1*: Maintenance infrastructure is in place to support version control and collaboration with model users.
+<!-- How *updatable* a CHEM is can be assessed against the criteria: -->
 
--	*U2*: Each new release of a CHEM is systematically retested, with changes implemented to minimise disruptions for existing model users.
+<!-- -	*U1*: Maintenance infrastructure is in place to support version control and collaboration with model users. -->
+
+<!-- -	*U2*: Each new release of a CHEM is systematically retested, with changes implemented to minimise disruptions for existing model users. -->

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Intro_Short.Rmd
for coauthor feedback draft
v4

@@
 A health economic model should be capable of being described using words and figures (a conceptual model), equations (a mathematical model) and software (a computational model).
-Health economic scientific manuscripts typically describe a model using its conceptual and mathematical representations but report results that have been generated by its computational representation (i.e., the format that allows computers to perform calculations).
-The conceptual, mathematical and computational representations of a model are assumed to be internally consistent.
-However, the ease of independently assessing the validity of this assumption depends in part on how a health economic model has been implemented computationally.
+Health economic scientific manuscripts typically involve description of a model using its conceptual and mathematical representations and report results that have been generated by its computational representation (i.e., the format that allows computers to perform calculations).
+Authors of these manuscripts assume (rarely explicitly) that the conceptual, mathematical and computational representations of a model are internally consistent and externally valid.
+The feasibility of independently assessing the validity of this assumption depends in part on how a health economic model has been implemented computationally.
 
 The set of software files implementing a health economic model's structure, data and algorithms can be called a computational health economic model (CHEM).
-CHEMs can be developed using commercial modelling <!-- Addition -->or spreadsheet<!-- /Addition --> software or open-source programming languages such as Python [@python2009] or R [@RCORE2022].
+CHEMs can be developed using commercial modelling <!-- Addition -->or spreadsheet<!-- /Addition --> software (for example, Microsoft Excel [@msexcel]) or open-source programming languages such as Python [@python2009] or R [@RCORE2022].
 An advantage of commercial <!-- modelling  -->software is that users often require limited or no programming skills to develop and apply CHEMs.
-<!-- Moved -->Open-source CHEMs remain rare [@Feenstra2022; @Emerson2019]<!-- Moved -->.
-However, <!-- Addition -->it has been argued that<!-- /Addition --> open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022].
-The decision about what type of software development platform to use must balance ethical considerations and resource constraints.
+<!-- Addition -->It has been argued that<!-- /Addition --> open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022].
+<!-- Moved -->However, open-source CHEMs remain rare [@Feenstra2022; @Emerson2019]<!-- Moved -->.
+
+Health economists must balance ethical considerations and resource constraints when deciding what type of software development platform to use for authoring CHEMs.
 <!-- ( see--><!-- Addition --> In Table \@ref(tab:proscons), we summarise some of the trade-offs we consider when undertaking modelling projects<!-- ) -->.
+<!-- Edit Move -->Determining how to make these trade-offs can be difficult.
+The computational modelling landscape is changing fast [@calder2018computational] and the ethical <!-- responsibilities  --><!-- Edit -->dimensions of computational public health modelling <!-- /Edit -->remain poorly understood [@10.3389/fpubh.2017.00068].
+<!-- /Edit Move-->
 
@@
 -->
-<!-- Edit Move -->Determining how to make these trade offs can be difficult.
-The computational modelling landscape is changing fast [@calder2018computational] and the ethical <!-- responsibilities  --><!-- Edit -->dimensions of computational public health modelling <!-- /Edit -->remain poorly understood [@10.3389/fpubh.2017.00068].
-<!-- /Edit Move-->
 

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Tables.Rmd
v4

@@
   child_docs_ls <- list(one = "../Child_RMDs/MS_Tables_1.Rmd",
-                        two = "../Child_RMDs/MS_Tables_2.Rmd"
-                        # ,
-                        # three = "../Child_RMDs/MS_Tables_3.Rmd",
+                        two = "../Child_RMDs/MS_Tables_2.Rmd",
+                        three = "../Child_RMDs/MS_Tables_3.Rmd"
+                        #,
                         # four = "../Child_RMDs/MS_Tables_4.Rmd"

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Tables_2.Rmd
v4

@@
                  `Fitness for purpose` = c("Modeller judgments, model features and verification checks can be reviewed by third parties.","Third party use increases likelihood of errors being identified.","Models can be validly used for longer."),
-                 `Beneficial use` = c("Decision-makers can understand the strengths and weakness of models before applying them.","Can inform more decision problems, in more contexts with less duplicative effort.","Models can be validly used for longer."),
+                 `Beneficial use` = c("Decision-makers can understand the strengths and weakness of models before applying them.","Can inform more decision problems, in more contexts with less duplicative effort.",""#"Models can be validly used for longer."
+                                      ),
                  check.names = F)
-if(params$X@outp_formats_chr[1] == "Word") {
  library("flextable")
   library("ftExtra")
+if(params$X@outp_formats_chr[1] == "Word") {
+
   df %>%
@@
     # library(dplyr)
-    #flextable::set_flextable_defaults() #fonts_ignore=TRUE
-    # flextable::flextable(df) %>%
+   # flextable::set_flextable_defaults() #fonts_ignore=TRUE
+    flextable::flextable(df) %>%
       # flextable::border_inner(border = officer::fp_border(style = "none")) %>%
       # flextable::border_outer() %>%
-    #   flextable::align(align = "center", part = "header") %>%
+      flextable::align(align = "center", part = "header") 
+    # %>%
+    #   flextable::merge_at(i = 3, j = 3:4)
       # flextable::merge_h() %>%
@@
       #huxtable::style_headers(bold = TRUE)
-    df %>%
-      kableExtra::kbl(booktabs = T,
-                      escape = F,
-                      caption = knitr::opts_current$get("tab.cap")) %>%
-      kableExtra::kable_paper(full_width = F) %>%
-      kableExtra::column_spec(1, bold=T) %>%
-      kableExtra::add_header_above(c(" ", "Project team responsibility" = 3), bold = T, italic = T) %>%
-      kableExtra::column_spec(2, width = "13em") %>%
-      kableExtra::column_spec(3, width = "13em") %>%
-      kableExtra::column_spec(4, width = "13em") %>%
-      kableExtra::collapse_rows(columns = 2, latex_hline = "none")
-      # kableExtra::column_spec(3, width = "35em") %>%
-      # kableExtra::pack_rows("Accountable - it is easy to see who developed, tested and applied a model and how they did it.",1,7) %>%
-      # kableExtra::pack_rows("Updatable - a model and its components are maintained and continuously improved.",17,20) %>%
-      # # kableExtra::pack_rows("Modular - models and their components can be combined to extend their scope",11,12) %>%
-      # # kableExtra::pack_rows("Epitomising - model code can be re-used in other decision contexts",13,15) %>%
-      # kableExtra::pack_rows("Reusable - a model and its components can be used in other models and by other modellers.",8,16) #%>%
-      # kableExtra::pack_rows("Yielding - a model can be simply, flexibly and reliably used as a decision aid",19,20)
+    # df %>%
+    #   kableExtra::kbl(booktabs = T,
+    #                   escape = F,
+    #                   caption = knitr::opts_current$get("tab.cap")) %>%
+    #   kableExtra::kable_paper(full_width = F) %>%
+    #   kableExtra::column_spec(1, bold=T) %>%
+    #   kableExtra::add_header_above(c(" ", "Project team responsibility" = 3), bold = T, italic = T) %>%
+    #   kableExtra::column_spec(2, width = "13em") %>%
+    #   kableExtra::column_spec(3, width = "13em") %>%
+    #   kableExtra::column_spec(4, width = "13em") %>%
+    #   kableExtra::collapse_rows(columns = 2, latex_hline = "none")
+
     }

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Tables_3.Rmd
v4

@@
 ---
-```{r cpkgs, tab.cap='Software framework R libraries', tab.id = 'cpkgs', echo = F, results = 'asis'}
+
+\renewcommand{\arraystretch}{2}
+```{r trucrit, tab.cap='Criteria for assessing the transparency, reusability and updatability of CHEMs', tab.id = 'cpkgs', echo = F, results = 'asis'}
 library(kableExtra) 
-df <- tibble::tibble(Package = c("ready4",
-                                       "ready4fun",
-                                       "ready4class",
-                                       "ready4pack",
-                                       "ready4use",
-                                       "ready4show"),
-                     Purpose = c("Provides a template model module and novel syntax and tools for finding interoperable CHEM modules, datasets and reproducible analysis programs.",
-                                 "Streamline and standardise the authoring and documenting of functions that support selective reuse of model algorithms.",
-                                 "Streamline and standardise the authoring and documenting of new CHEM modules.",
-                                 "Help bundle and disseminate CHEM modules as documented, licensed and quality assured  R libraries.",
-                                 "Help manage the labelling and transfer of data between CHEM modules and local and remote data repositories.",
-                                 "Support the authoring of programs that implement integrated and reproducible data ingest, analysis and reporting pipelines."),
-                           # A = c("3","2-3","2","1-2,5,7","1-2,7","4,6"), 
-                           # R = c("4","5","4-5","1,3,6","3,7",""),
-                           # U = c("","","","1-3","1",""),
-                           `R library dependencies` = Package %>%
-                             purrr::map(~ {
-                               utils::packageDescription(.x) %>%
-                                 `[`(c("Depends", "Imports")) %>%
-                                 purrr::map(~{
-                                   if(is.null(.x)){
-                                     character(0)
-                                     }else{
-                                       .x %>%
-                                         strsplit(",\\n") %>%
-                                         purrr::flatten_chr() %>%
-                                         purrr::map(~strsplit(.x,", ") %>%
-                                                      purrr::flatten_chr()) %>%
-                                         purrr::flatten_chr() %>% sort() %>%
-                                         purrr::discard(~startsWith(.x,"R "))
-                                       }
-                                   }) %>%
-                                 purrr::flatten_chr() %>%
-                                 unique() %>%
-                                 sort() %>%
-                                 purrr::map_chr(~{
-                                   updated_1L_chr <- stringr::str_replace_all(.x,"\\n"," ")
-                                   problem_idx_1L_chr <- stringr::str_locate(updated_1L_chr," ")[1,1] %>%
-                                     unname()
-                                   if(!is.na(problem_idx_1L_chr))
-                                     updated_1L_chr <- updated_1L_chr %>%
-                                     stringr::str_sub(end = problem_idx_1L_chr-1)
-                                   updated_1L_chr %>% 
-                                     trimws(which = "left") 
-                                   })
-                               }) %>% 
-                             purrr::map_chr (~ .x %>% paste0(collapse = " "))) %>% 
-  as.data.frame() 
-# if(params$X@outp_formats_chr[1] != "Word"){
-#       df$Ref <- c(rep("",6)
-#        # "@ready42022","@ready4fun2022","@ready4class2022","@ready4pack2022","@ready4use2022","@ready4show2022"
-#                   )
-# }else{
-#   df$Ref <- ""
-# }
-
-df <- df %>%
-  # dplyr::mutate(Standard = purrr::pmap_chr(df,
-  #                                          ~paste0("A",
-  #                                                 ..3,
-  #                                                 paste0(rep("&nbsp;",
-  #                                                     2+max(nchar(df[,3]))-nchar(..3)),
-  #                                                     collapse = ""#
-  #                                                     ),
-  #                                                 ifelse(..4=="","","R"),
-  #                                                 ..4,
-  #                                                 paste0(rep("&nbsp;",
-  #                                                     2+max(nchar(df[,4]))-nchar(..4)),
-  #                                                     collapse = ""#
-  #                                                     ),
-  #                                                 ifelse(..5=="","","U"),
-  #                                                 ..5,
-  #                                                 paste0(rep("&nbsp;",
-  #                                                     1+max(nchar(df[,5]))-nchar(..5)),
-  #                                                     collapse = ""#
-  #                                                     )#,
-  #                                                 ))) %>%
-
-  dplyr::select(Package,
-                #Ref,
-                Purpose,
-                #Standard,
-                `R library dependencies`)
-
-names_chr <- names(df)
-# df <- data.frame(df[,1],
-#                  rep(ifelse(params$X@outp_formats_chr[1] == "Word"," ","&nbsp; &nbsp; &nbsp;"),
-#                      nrow(df)),
-#                  df[,2#:3
-#                     ],
-#                  rep(ifelse(params$X@outp_formats_chr[1] == "Word"," ","&nbsp; &nbsp; &nbsp;")
-#                      ,nrow(df)),
-#                  #df[,4],
-#                  #rep("&nbsp; &nbsp; &nbsp;",nrow(df)),
-#                  df[,3#5
-#                     ])
-# colnames(df) <- c("Package",
-#                   ifelse(params$X@outp_formats_chr[1] == "Word","XXX"," "),
-#                   #"Ref", 
-#                   "Purpose", 
-#                   ifelse(params$X@outp_formats_chr[1] == "Word","YYY"," "),
-#                   #"Standard     ",# " ",
-#                   "R library dependencies")
-#df$Package <- "![](../Data/images/ready4.png){ width=10% }"#"\\includegraphics[valign=T,scale=2.5,raise=2mm]{../Data/images/ready4.png}"
-#df$`Standard     `  <- df$`Standard     ` %>% stringr::str_replace_all("&nbsp;"," ")
+df <- data.frame(Attribute = c(rep("Transparent",2), rep("Reusable",2), rep("Updtatable",2)),
+                 Item = c("T","R","U") %>% purrr::map(~paste0(.x,1:2)) %>% purrr::flatten_chr(),
+                 Criterion = c("Software files (executable code, testing procedures and outcomes and non-confidential data) are publicly available.",
+                               "It is easy to see who funded, developed and tested each part of the CHEM and to identify the modelling team’s assumptions, judgments and theories about CHEM development and use.",
+                               "Programming practices facilitate selective reuse of model sub-components.",
+                               "Licenses allow anyone to ethically reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works.",
+                               "Maintenance infrastructure is in place to support version control and collaboration with model users.",
+                               "Each new release of a CHEM is systematically retested, with changes implemented to minimise disruptions for existing model users."))
 if(params$X@outp_formats_chr[1] == "Word") {
@@
   library("ftExtra")
-  df %>% #[,c(1,3:4,6:7)] %>%
+  df %>%
     ready4show::print_table(output_type_1L_chr = params$X@outp_formats_chr[1],
@@
   }else{
-    #df$Ref<-NULL
-    #df$Package <- ""
-    # df$Package  <- df$Package %>% 
-    #                   purrr::map_chr(~paste0("\\includegraphics[valign=M,width=1.9cm,raise=4mm]{../Data/images/",.x,"_logo.png}"))#scale=0.4,raise=3mm]
-    # df$Package <- "\\includegraphics[valign=M,scale=0.08,raise=3mm]{../Data/images/ready4.png}"
-    #df[,c(1,3,5:6)]   %>%
     df %>%
-    # kableExtra::kbl(format = "markdown",
-    #                   escape = F,
-    #                 caption = knitr::opts_current$get("tab.cap"))
       kableExtra::kbl(booktabs = T, 
                       escape = F,
+                      linesep = "\\addlinespace",
                       caption = knitr::opts_current$get("tab.cap")) %>%
       kableExtra::kable_paper(full_width = F) %>%
-      kableExtra::column_spec(2, width = "20em", latex_valign = "m") %>%
-      # kableExtra::column_spec(1, 
-      #                         image = kableExtra::spec_image(rep("../Data/images/ready4.png",nrow(df)), 
-      #                                                           150, 150)) %>%
-      kableExtra::column_spec(3, width = "20em",
-                              latex_valign = "m")
+      #kableExtra::pack_rows("Resource requirements implications",1,4, hline_after = T) %>%
+      #kableExtra::pack_rows("Ethical implications",5,14, hline_after = T) %>%
+      kableExtra::column_spec(1, width = "8em", bold=T) %>% #, latex_valign = "m"
+      kableExtra::column_spec(2, width = "3em") %>%
+      kableExtra::column_spec(3, width = "28em") %>%
+      kableExtra::collapse_rows(columns = 1, latex_hline = "major") 
+    # %>%
+    #   kableExtra::kable_styling(latex_options = "striped")
+      # kableExtra::pack_rows("Accountable - it is easy to see who developed, tested and applied a model and how they did it.",1,7) %>%
+      # kableExtra::pack_rows("Updatable - a model and its components are maintained and continuously improved.",17,20) %>%
+      # kableExtra::pack_rows("Modular - models and their components can be combined to extend their scope",11,12) %>%
+      # kableExtra::pack_rows("Epitomising - model code can be re-used in other decision contexts",13,15) %>%
+      # kableExtra::pack_rows("Reusable - a model and its components can be used in other models and by other modellers.",8,16) #%>%
+      # kableExtra::pack_rows("Yielding - a model can be simply, flexibly and reliably used as a decision aid",19,20)
+
     }

### 2024-03-19 Markdown/Manuscript/Child_RMDs/MS_Tables_4.Rmd
v4

@@
 df <- data.frame(Criterion = c("T1 Software files are open access", 
-                               "T2 Project team contributions and judgments are easily identified", 
+                               "T2 Developer contributions and judgments on appropriate use are easily identified", 
                                "R1 Programming practices promote selective reuse",
-                               "R2 Licenses restrict only unethical reuse",
+                               "R2 Licenses permit reuse and derivative works",
                                "U1 Maintenance infrastructure is in place",
-                               "U2 New releases are systematically retested and deprecated"),
+                               "U2 Releases are systematically retested and deprecated"),
                  `Met?` = c(rep("Yes",5),"Partial"),

### 2024-03-19 setup.R

@@ -8,12 +8,12 @@ X <- Ready4showSynopsis(#background_1L_chr = "Our study is entirely fictional.",
                         #conclusion_1L_chr = "We have identified criteria that can be used to systematically assess extent to which the computational implementation of health economic models adheres to the ethical goals of transparency, reusability and updatability. We have developed an open-source software framework that can support the ethical computational implementation of economic models in youth mental health. Our framework can be used as a prototype for developing future software frameworks to support ethical implementation of CHEMs.",
                         digits_int = 3L,
                         ethics_1L_chr = "None.",
-                        funding_1L_chr = "Funded by an Australian Government Research Training Program (RTP) Scholarship. ",
+                        funding_1L_chr = "Funded by an Australian Government Research Training Program (RTP) Scholarship to Matthew Hamilton. ",
                         #interval_chr = "three months",
                         keywords_chr = c("computational models", "ethics of modelling", "health economics", "open-source models"),
                         outp_formats_chr = "PDF",
                         #sample_desc_1L_chr = "The study sample is fake data that pretends to be young people aged 12 to 25 years who attended Australian primary care services for mental health related needs between November 2019 to August 2020.",
-                        title_1L_chr = "Ethical considerations for computational implementation of health economic models")
+                        title_1L_chr = "Identifying and assessing ethical risks when implementing health economic models as software")
 X <- renewSlot(X,
                "authors_r3", 
                {
@@ -37,17 +37,17 @@ X <- renewSlot(X,
                "abstract_args_ls",
                {
                  args_ls <- X@abstract_args_ls
-                 args_ls$abstract_ls <- list(Summary = "Most health economic analyses are undertaken with the aid of computers. However, the ethical dimensions of implementing health economic models as software (or computational health economic models (CHEMs)) are poorly understood. 
-                                                        We propose that developers and funders of CHEMs should pursue the ethical goals of: (i) socially acceptable user requirements and design specifications; (ii) fit for purpose implementations; and (iii) socially beneficial post-release use. 
+                 args_ls$abstract_ls <- list(Summary = "Most health economic analyses are undertaken with the aid of computers. However, the research ethics of implementing health economic models as software (or computational health economic models (CHEMs)) are poorly understood. 
+                                                        We propose that developers and funders of CHEMs should adhere to research ethics principles and pursue the goals of: (i) socially acceptable user requirements and design specifications; (ii) fit for purpose implementations; and (iii) socially beneficial post-release use. 
[…]

### 2024-03-20 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
ready for preprint

--- 2024-02-14 Markdown/Manuscript/Child_RMDs/MS_Aims.Rmd
@@
 <!-- Important rewordings --> 
-In this paper, we <!--  MOVED_EDITED_START -->draw on our prior experience with developing and using CHEMs and consider literature on modelling practice to propose<!--  MOVED_EDITED_END -->:  (i) a set of ethical <!--  EDITED_Start responsibilities-->goals<!--  _EDITED_END --> for CHEM <!--  EDIT -->developers and funders<!--  /EDIT --<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that can reduce ethical risks; and (iii) criteria against which these attributes can be assessed.
+In this paper, we <!--  MOVED_EDITED_START -->draw on our experience with developing and using CHEMs and consider literature on modelling and software development practice to propose<!--  MOVED_EDITED_END -->:  (i) an initial set of research ethics <!--  EDITED_Start responsibilities-->goals<!--  _EDITED_END --> for CHEM <!--  EDIT -->developers and funders<!--  /EDIT --<!-- and criteria for assessing ethical CHEM implementations -->;<!--  MOVED_EDITED_START --> (ii) attributes of CHEMs that suggest achievement of these goals; and (iii) criteria against which these attributes can be assessed.
 <!--  MOVED_EDITED_END -->

### 2024-03-20 Markdown/Manuscript/Child_RMDs/MS_Discussion.Rmd
ready for preprint

@@
 
-Prior literature has noted that health economic models routinely have limitations in transparency [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], reuse [@Feenstra2022; @Emerson2019], and being kept up to date [@Sampson2017; @10.3389/fpubh.2022.899874].
-These limitations not only pose ethical risks such as those we have described, but are practical barriers to achieving potential exciting enhancements to health economic modelling practice such as improved model transferability [@munthe2019systematic; @barbieri2010international], greater use of reference models [@Afzali2013], living health technology assessments [@thokala2023living] and leveraging the analytic capabilities of machine learning [@padula2022machine].
+Health economic models routinely have limitations in transparency [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], reusability [@Feenstra2022; @Emerson2019], and updatability [@Sampson2017; @10.3389/fpubh.2022.899874].
+Few current CHEMs would meet all six TRU criteria that we propose, which collectively require that a CHEM is implemented as an open-source project<!-- and model transferability -->.
+Despite in principle support from many in the health economics field [@Pouwels2022], open-source CHEMs remain rare [@Feenstra2022; @Emerson2019].
 
-It is important that the health economics community engages with the ethical dimensions of how health economic models are implemented as software.
-Recent developments in Large Language Models (LLMs) mean software developers are confronted with novel ethical issues including: increased risk of security flaws; a potential need to add restrictions to open-source licensing; environmentally harmful levels of computing power; and concerns about model explainability and unintended consequences [@ozkaya2023application].
-Early work exploring the use of LLMs in CHEM development has already been undertaken [@poirrier2023msr26].
-Many of the active ethics debates in the computer science community are therefore likely to become increasing relevant to health economists and public health researchers.
-Documents such as the World Health Organisation's recent guidance on generative artificial intelligence [@world2024ethics] will be important reference points for how the health economics discipline navigates emerging ethical themes in CHEM development.
-
-Meeting all six TRU criteria we propose is more likely when a CHEM has been implemented as an open-source project<!-- and model transferability -->.
-It should be noted that the ethical dimensions of open-source software development are broader than the issues we have discussed here, including issues relating to social contracts, contributor autonomy and public goods [@grodzinsky2003ethical].
-There are also important barriers to growing the (currently paltry) number of open-source CHEMs.
-These barriers include health economist concerns about intellectual property, confidential data, the risk of model misuse and the resources required for model maintenance [@Pouwels2022; @Wu2019].
+Barriers to open-source CHEMs include health economist concerns about intellectual property, confidential data, the risk of model misuse and the resources required for model maintenance [@Pouwels2022; @Wu2019].
 Commercial considerations, for example when health economic models are owned by pharmaceutical companies and consultancies, may also limit the public availability of models and their constituent code and data [@Feenstra2022].
@@
 
-Recent efforts to develop the open-source CHEMs field include an initiative to develop a suite of open source CHEMs addressing different health conditions in the United States [@Jansen2019] and a Canadian platform facilitating access to and validation of multiple CHEMs [@harvard2022developing].
-It has also been recommended that training curricula for current and future health economists need to develop some foundational software development skills [@kindilien2021role].
-Education efforts to develop such skills could be reinforced with best-practice scientific computing guidance for health economists.
+<!-- Recent efforts to develop the open-source CHEMs field include an initiative to develop a suite of open source CHEMs addressing different health conditions in the United States [@Jansen2019] and a Canadian platform facilitating access to and validation of multiple CHEMs [@harvard2022developing].
+-->
 Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020], but guidelines for implementing open source CHEMs remain scarce, piecemeal and need improving [@Sampson2019].
+Training curricula for current and future health economists that develop foundational software development skills have been recommended [@kindilien2021role].
 Health economists also need to be able to access reliable and up to date legal advice about the implications of alternative approaches to software licensing.
+The ethical dimensions of open-source software development are broader than the issues we have discussed here and include issues relating to social contracts, contributor autonomy and public goods [@grodzinsky2003ethical].
 
@@
 
-Many of these barriers require significant investments by governments and other research funders in enabling infrastructure and human capital.
+Achieving the ethical goals we propose will require significant investments by governments and other research funders in enabling infrastructure and human capital.
 <!-- A core responsibility of research funders is reducing waste in research [@chalmers2014increase].
 Investments in modelling projects might generate greater social returns if they supported the project activities necessary to develop CHEMs that are easily understood, widely reused and well maintained.
---> Investment strategies that have previously been recommended for development of the health economic research field include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open-source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019].
+--> Related investment strategies that have previously been recommended for development of the health economic research field include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open-source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019].
 

### 2024-03-20 Markdown/Manuscript/Child_RMDs/MS_Imp_Guidelines.Rmd
ready for preprint

@@
 Boden and McKendrick [@10.3389/fpubh.2017.00068] propose a framework for ethical public health modelling based on the criteria of *independence* (concerning how modeller subjectivity shapes model design), *beneficence / non-maleficence* (concerning model quality and utility), *transparency* (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and *justice* (concerning the social obligations of modellers to consider and communicate ethical issues about model use).
-<!-- However, t -->This framework is not specific to a model's implementation as software, but <!-- are designed to  -->apply to a modelling project as a whole, including a model's conceptual and mathematical representations.
+<!-- However, t -->This framework is not specific to a model's implementation as software, but <!-- are designed to  -->applies to a modelling project as a whole, including a model's conceptual and mathematical representations.
 Boden and McKendrick proposed 13 questions to evaluate ethical risk across the four criteria, but of these, only two pertain to computational models: "is the model code open-source or available on request" (for transparency) and "has the model been verified, i.e., does it do what the modeller wants it to do?" (for beneficence / non-maleficence).
@@
 Software development lifecycle (SDLC) models take different approaches to the inclusion, naming, definition, sequencing and iteration of project phases [@ruparelia2010software].
-However, SDLC models typically have components that can also principally map to the concepts of *planning* (e.g., identification of system / user requirements and design specifications), *implementation* (e.g., development and testing) and *release* (e.g., deployment, system integration, maintenance and support).
+However, SDLC models typically have components that can principally map to the concepts of *planning* (e.g., identification of system / user requirements and design specifications), *implementation* (e.g., development and testing) and *release* (e.g., deployment, system integration, maintenance and support).
 For simplicity, we propose one overarching ethical <!-- responsibility  for CHEM project team-->goal for each phase.
@@
 These value judgments influence the assumptions, selection of model features and standards for evidence that shape health economic model projects [@HARVARD2020112975].
-For example, the process of determining CHEM *user requirements and design specifications* will involve making judgments about the competing claims of open-source and commercial development platforms (Table \@ref(tab:proscons)).
-That decision will in turn determine how straightforward it is for a CHEM to be subsequently amended to reflect alternative value judgments relating to model conceptualisation and mathematical formalisation (Table \@ref(tab:timelygls)).
+For example, the process of determining CHEM *user requirements and design specifications* will involve making judgments about the competing claims of proprietary and open-source development platforms (Table \@ref(tab:proscons)).
+An advantage of commercial <!-- modelling  -->software is that users often require limited or no programming skills to develop and apply CHEMs.
+<!-- Addition --><!-- /Addition --> Open-source software programming languages, on the other hand, may facilitate the development of CHEMs that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022].
+The choice of software development platform can influence how easily a CHEM can be amended by others to reflect alternative value judgments relating to model conceptualisation and mathematical formalisation (Table \@ref(tab:timelygls)).
 <!-- Moved -->
@@
 <!-- Edit --><!-- All three  -->Our proposed ethical goals are most likely to be met when there is both good<!-- have implications for both  --> professional practice by CHEM developers and <!-- Addition -->adequate <!-- Addition -->resourcing by project funders.
-<!-- /Edit --> The choices CHEM developers make often need to balance ethical considerations with project resource constraints (as illustrated in Table \@ref(tab:proscons)).
+<!-- /Edit --> The choices CHEM developers make often need to balance ethical considerations with project resource constraints (Table \@ref(tab:proscons)).
 Modelling projects implemented by diligent and conscientious CHEM developers may still be exposed to ethical risks if there are inadequate resources to meaningfully engage important stakeholder groups, prepare code and data artefacts for independent assessment and provide ongoing model maintenance and user support.
 Currently, it can take “an extraordinary amount of idealism” to commit to authoring and maintaining research software [@anzt2020environment].
-Existing incentive structures for research health economists generally do not promote facilitating peers to reuse their work and funding is rarely adequate to maintain and routinely update a model.
+Existing incentive structures for research health economists generally do not promote facilitating peers to reuse their work and funding is rarely adequate to routinely update a model.
 Economists may therefore make pragmatic judgments that proprietary licensing approaches provide the most feasible pathway to mobilising the resources to support model maintenance.

### 2024-03-20 Markdown/Manuscript/Child_RMDs/MS_Intro_Short.Rmd
ready for preprint

@@
 
-The set of software files implementing a health economic model's structure, data and algorithms can be called a computational health economic model (CHEM).
+The set of software tools and files implementing a health economic model's structure, data and algorithms can be called a computational health economic model (CHEM).
 CHEMs can be developed using commercial modelling <!-- Addition -->or spreadsheet<!-- /Addition --> software (for example, Microsoft Excel [@msexcel]) or open-source programming languages such as Python [@python2009] or R [@RCORE2022].
-An advantage of commercial <!-- modelling  -->software is that users often require limited or no programming skills to develop and apply CHEMs.
-<!-- Addition -->It has been argued that<!-- /Addition --> open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022].
-<!-- Moved -->However, open-source CHEMs remain rare [@Feenstra2022; @Emerson2019]<!-- Moved -->.
+Training and professional practice guidance for health economists typically provide only superficially coverage of software development topics.
+However, these capabilities will be critical for achieving proposed methodological innovations at the frontier of the health economics field.
+Opportunities for field development that closely overlap with CHEM implementation approaches include living health technology assessments [@thokala2023living], machine learning powered health economic analyses [@padula2022machine], open source health-economic models [@Pouwels2022] and utilising Large Language Models (LLMs) to program health economic models [@poirrier2023msr26].
+<!-- Potential enhancements to health economic modelling practice include improved model transferability [@munthe2019systematic; @barbieri2010international], greater use of reference models [@Afzali2013], living health technology assessments [@thokala2023living] and leveraging the analytic power of machine learning [@padula2022machine].
+-->
 
-Health economists must balance ethical considerations and resource constraints when deciding what type of software development platform to use for authoring CHEMs.
-<!-- ( see--><!-- Addition --> In Table \@ref(tab:proscons), we summarise some of the trade-offs we consider when undertaking modelling projects<!-- ) -->.
-<!-- Edit Move -->Determining how to make these trade-offs can be difficult.
+There is a significant ethical dimension to software development.
+For example, recent developments in LLMs mean software developers may confront novel ethical issues including: new security risks; more restrictive open-source licensing; environmentally harmful levels of computing power; and concerns about model explainability and unintended consequences [@ozkaya2023application].
+The health economics community should engage more directly with these ethical debates.
 The computational modelling landscape is changing fast [@calder2018computational] and the ethical <!-- responsibilities  --><!-- Edit -->dimensions of computational public health modelling <!-- /Edit -->remain poorly understood [@10.3389/fpubh.2017.00068].
+Documents such as the World Health Organisation's recent guidance on generative artificial intelligence [@world2024ethics] can be important reference points in helping the health economics discipline navigate relevant ethical topics in software development.
+
+  <!-- Moved -->
+ <!-- However, open-source CHEMs remain rare [@Feenstra2022; @Emerson2019], despite in principle support from many in the health economics field [@Pouwels2022]. -->
+ <!-- Moved -->
+
+<!-- Health economists must balance ethical considerations and resource constraints when deciding what type of software development platform to use for authoring CHEMs. <!-- ( see--><!-- Addition --> 
+<!--In Table \@ref(tab:proscons), we summarise some of the trade-offs we consider when undertaking modelling projects<!-- ) --> <!-- Edit Move --><!--Determining how to make these trade-offs can be difficult.
+-->
+
  <!-- /Edit Move-->

### 2024-03-20 Markdown/Manuscript/Child_RMDs/MS_Tables_2.Rmd
ready for preprint

@@
 date: "`r Sys.Date()`"
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
 ---
-```{r timelygls, tab.cap='Transparent, reusable and updatable (TRU) computational health economic models (CHEMs) are suggestive of ethical modelling practice', tab.id = 'timelygls', echo = F, results = 'asis', warning=FALSE}
-library(kableExtra)  
+
+```{r echo=FALSE}
 df <- data.frame(` ` = c("Transparent", "Reusable", "Updatable"),
@@
                                                                         "Models can be modified to reflect alternative value judgments.",
-                                                                        "Models can be modified to reflect alternative value judgments."),
-                 `Fitness for purpose` = c("Modeller judgments, model features and verification checks can be reviewed by third parties.","Third party use increases likelihood of errors being identified.","Models can be validly used for longer."),
-                 `Beneficial use` = c("Decision-makers can understand the strengths and weakness of models before applying them.","Can inform more decision problems, in more contexts with less duplicative effort.",""#"Models can be validly used for longer."
+                                                                        ""#"Models can be modified to reflect alternative value judgments."
+                                            ),
+                 `Fitness for purpose` = c("",#"Modeller judgments, model features and verification checks can be reviewed by third parties.",
+                                           "Third party use increases likelihood of errors being identified.",
+                                           "Models can be validly used for longer."),
+                 `Beneficial use` = c("Decision-makers can understand the strengths and weakness of models before applying them.",
+                                      "Can inform more decision problems, in more contexts with less duplicative effort.",
+                                      ""#"Models can be validly used for longer."
                                       ),
                  check.names = F)
- library("flextable")
-  library("ftExtra")
-if(params$X@outp_formats_chr[1] == "Word") {
-
-  df %>%
-    ready4show::print_table(output_type_1L_chr = params$X@outp_formats_chr[1],
-                            caption_1L_chr = knitr::opts_current$get("tab.cap"),
-                            mkdn_tbl_ref_1L_chr = paste0("tab:",knitr::opts_current$get("tab.id")),
-                            use_rdocx_1L_lgl = ifelse(params$X@outp_formats_chr[1]=="Word",T,F),
-                            add_to_row_ls = NULL,
-                            sanitize_fn = NULL) %>%
-    ftExtra::colformat_md()
+```
+```{r echo=FALSE}
+if(params$X@outp_formats_chr[1] == "Word"){
+  conditional_1L_chr <- "../Child_RMDs/MS_Tables_2_WORD.Rmd"
 }else{
-    # library(huxtable)
-    # library(dplyr)
-   # flextable::set_flextable_defaults() #fonts_ignore=TRUE
-    flextable::flextable(df) %>%
-      # flextable::border_inner(border = officer::fp_border(style = "none")) %>%
-      # flextable::border_outer() %>%
-      flextable::align(align = "center", part = "header") 
-    # %>%
-    #   flextable::merge_at(i = 3, j = 3:4)
-      # flextable::merge_h() %>%
-      # flextable::merge_v() %>%
-      # flextable::valign(j=2, valign = "top") %>%
-      # flextable::align(j=2:4, align = "center") %>%
-      # flextable::bold(j=1, part = "all") %>%
-      # flextable::bold(part = "header") %>%
-      # flextable::width(j = 1, width = 1.2) %>%
-      # flextable::width(j = 2:4, width = 1.6)
-    # %>%
-    #   
-      #flextable::merge_at(i=2,j=1:2)
-    # huxtable::as_hux(df) %>% 
-    #   #insert_row("MPG-DISP", "", "", "HP-WT", fill = "") %>% 
-    #   #huxtable::merge_cells(2:3,1) %>% 
-    #  # huxtable::set_header_rows(1, TRUE) %>% 
-    #   huxtable::set_col_width(c(.1,.1,.1,.7)) %>%
-    #   huxtable::set_wrap(1, value=T)
-      #huxtable::style_headers(bold = TRUE)
-    # df %>%
-    #   kableExtra::kbl(booktabs = T,
-    #                   escape = F,
-    #                   caption = knitr::opts_current$get("tab.cap")) %>%
-    #   kableExtra::kable_paper(full_width = F) %>%
-    #   kableExtra::column_spec(1, bold=T) %>%
-    #   kableExtra::add_header_above(c(" ", "Project team responsibility" = 3), bold = T, italic = T) %>%
-    #   kableExtra::column_spec(2, width = "13em") %>%
-    #   kableExtra::column_spec(3, width = "13em") %>%
-    #   kableExtra::column_spec(4, width = "13em") %>%
-    #   kableExtra::collapse_rows(columns = 2, latex_hline = "none")
-
+    conditional_1L_chr <- "../Child_RMDs/MS_Tables_2_LATEX.Rmd"
 }
 ```
+```{r child = conditional_1L_chr, echo=FALSE}
+```

### 2024-03-20 Markdown/Manuscript/Child_RMDs/MS_Tables_2_LATEX.Rmd
ready for preprint

---
title: "Untitled"
output:
  pdf_document: default
  html_document: default
date: "2024-03-20"
---

\begin{table}
\centering
\caption{\label{tab:timelygls}Transparent, reusable and updatable (TRU) computational health economic models (CHEMs) are suggestive of ethical modelling practice}
\centering
\begin{tabularx}{14cm}{|l|X|X|X|}
\toprule
  & \textbf{Social acceptability} & \textbf{Fitness for purpose} & \textbf{Beneficial use}\\
\midrule
\textbf{Transparent} & \multicolumn{2}{>{\hsize=\dimexpr 2\hsize+2\tabcolsep+\arrayrulewidth}X|}{\centering Modeller judgments, model features and verification checks can be reviewed by third parties.}  & Decision-makers can understand the strengths and weakness of models before applying them.\\[20pt] \cmidrule(l{7mm}r{7mm}){2-4}

\textbf{Reusable} & \multirow{2}{3cm}{\centering Models can be modified to reflect alternative value judgments.} & Third party use increases likelihood of errors being identified. & Can inform more decision problems, in more contexts with less duplicative effort.\\[20pt] \cmidrule(l{7mm}r{7mm}){3-4}

\textbf{Updatable} & & \multicolumn{2}{>{\hsize=\dimexpr 2\hsize+2\tabcolsep+\arrayrulewidth}X|}{\centering Models can be validly used for longer.}  \\[20pt]
\bottomrule
\end{tabularx}
\end{table}

### 2024-03-20 Markdown/Manuscript/Child_RMDs/MS_Tables_2_LATEX.tex
ready for preprint

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Untitled},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Untitled}
\author{}
\date{\vspace{-2.5em}2024-03-20}

\begin{document}
\maketitle

\begin{table}
\centering
\caption{\label{tab:timelygls}Transparent, reusable and updatable (TRU) computational health economic models (CHEMs) are suggestive of ethical modelling practice}
\centering
\begin{tabular}[t]{>{}l>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{13em}}
\toprule
  & Social acceptability & Fitness for purpose & Beneficial use\\
\midrule
\textbf{Transparent} & Modeller judgments, model features and verification checks can be reviewed by third parties. &  & Decision-makers can understand the strengths and weakness of models before applying them.\\

\textbf{Reusable} & Models can be modified to reflect alternative value judgments. & Third party use increases likelihood of errors being identified. & Can inform more decision problems, in more contexts with less duplicative effort.\\

\textbf{Updatable} &  & Models can be validly used for longer. & \\
\bottomrule
\end{tabular}
\end{table}

\end{document}

### 2024-03-20 Markdown/Manuscript/Child_RMDs/MS_Tables_2_WORD.Rmd
ready for preprint

---
title: "Untitled"
output:
  pdf_document: default
  html_document: default
date: "2024-03-20"
---

```{r timelygls, tab.cap='Transparent, reusable and updatable (TRU) computational health economic models (CHEMs) are suggestive of ethical modelling practice', tab.id = 'timelygls', echo = F, results = 'asis', warning=FALSE}
library(kableExtra)

 library("flextable")
  library("ftExtra")
if(params$X@outp_formats_chr[1] == "Word") {

  df %>%
    ready4show::print_table(output_type_1L_chr = params$X@outp_formats_chr[1],
                            caption_1L_chr = knitr::opts_current$get("tab.cap"),
                            mkdn_tbl_ref_1L_chr = paste0("tab:",knitr::opts_current$get("tab.id")),
                            use_rdocx_1L_lgl = ifelse(params$X@outp_formats_chr[1]=="Word",T,F),
                            add_to_row_ls = NULL,
                            sanitize_fn = NULL) %>%
    ftExtra::colformat_md()
  }else{
    #library(huxtable)
    # library(gt)
    # library(dplyr)
   # flextable::set_flextable_defaults() #fonts_ignore=TRUE
    # flextable::flextable(df) %>%
      # flextable::border_inner(border = officer::fp_border(style = "none")) %>%
      # flextable::border_outer() %>%
      # flextable::align(align = "center", part = "header") 
    # %>%
    #   flextable::merge_at(i = 3, j = 3:4)
      # flextable::merge_h() %>%
      # flextable::merge_v() %>%
      # flextable::valign(j=2, valign = "top") %>%
      # flextable::align(j=2:4, align = "center") %>%
      # flextable::bold(j=1, part = "all") %>%
      # flextable::bold(part = "header") %>%
      # flextable::width(j = 1, width = 1.2) %>%
      # flextable::width(j = 2:4, width = 1.6)
    # %>%
    #   
      #flextable::merge_at(i=2,j=1:2)
    # gt::gt(df) %>%
    #    gt::cols_width(
    # ` ` ~ gt::px(50),
    # `Social acceptability` ~ gt::px(120),
    # `Fitness for purpose` ~ gt::px(120),
    # `Beneficial use` ~ gt::px(120)#,
    #gt::everything() ~ gt::px(60)
       #) # %>%
    #   gt::cols_merge(columns = `Social acceptability`,
    #                  rows = 2:3) %>%
      # gt::cols_merge(columns = c(`Fitness for purpose`, `Beneficial use`),
      #                rows = 1)

    # huxtable::as_hux(df) %>%
    #   huxtable::to_latex(tabular_only = TRUE) 
    #%>%
      #insert_row("MPG-DISP", "", "", "HP-WT", fill = "") %>%
     #  huxtable::merge_cells(2:3,1) %>%
     # huxtable::set_header_rows(1, TRUE) %>%
      # huxtable::set_col_width(c(.1,.1,.1,.7)) %>%
      #huxtable::set_wrap(1, value=T) %>%
   # huxtable::style_headers(bold = TRUE)
    # df %>%
    #   kableExtra::kbl(booktabs = T,
    #                   escape = F,
    #                   caption = knitr::opts_current$get("tab.cap")) %>%
    #   kableExtra::kable_paper(full_width = F) %>%
    #   kableExtra::column_spec(1, bold=T) %>%
    #   #kableExtra::add_header_above(c(" ", "Project team responsibility" = 3), bold = T, italic = T) %>%
    #   kableExtra::column_spec(2, width = "13em") %>%
    #   kableExtra::column_spec(3, width = "13em") %>%
    #   kableExtra::column_spec(4, width = "13em") %>%
    #   kableExtra::collapse_rows(columns = 2, latex_hline = "none")

    }
```

### 2024-03-20 Markdown/Manuscript/Parent_PDF/Springer.Rmd
v4
ready for preprint

@@
 
-title: Ethical considerations for computational implementation of health economic models
-titlerunning: Ethical computational modelling
+title: "Ethical considerations when planning, implementing and releasing health economic model software: a new proposal"
+titlerunning: Ethical modelling software
 
@@
     info:
-#      orgdiv: XXXXXXXXXXX
+      orgdiv: Flinders Health and Medical Research Institute
       orgname: Flinders University   
@@
     info:
-#      orgdiv: XXXXXXXXXXX
+      orgdiv: Health Research Institute
       orgname: University of Canberra  
@@
   \usepackage{makecell}
+  \usepackage{array}
+  \newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
+  \newcolumntype{M}[1]{>{\centering\arraybackslash}m{#1}}
+  \newcolumntype{N}{@{}m{0pt}@{}}
+  \usepackage{caption}
+  \usepackage{graphicx}
+  \usepackage{siunitx}
+  \usepackage{hhline}
+  \usepackage{calc}
+  \usepackage{tabularx}
+  \renewcommand\tabularxcolumn[1]{m{#1}}% for vertical centering text in X column
+
+  \usepackage{threeparttable}
+  \usepackage{wrapfig}
+  \usepackage{adjustbox}
+  \usepackage{hyperref}
 

### 2024-03-20 Markdown/Manuscript/Parent_PDF/Test_Tb.Rmd
ready for preprint

---
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    # includes:
    #   in_header: preamble.tex
    keep_tex: true
#     pandoc_args: 
#       - '--lua-filter=../Filters/scholarly-metadata.lua'
#       - '--lua-filter=../Filters/author-info-blocks.lua'
classoptions: 
  - sn-vancouver      
  - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
  # - referee       # Optional: Use double line spacing 
  # - lineno        # Optional: Add line numbers
  # - iicol         # Optional: Double column layour

title: Identifying and assessing ethical risks when implementing health economic models as software
titlerunning: Ethical computational modelling

authors: 
  - firstname: Matthew
    particle: P 
    lastname: Hamilton
    email: matthew.hamilton1@monash.edu
    affiliation: [1,2]
    corresponding: TRUE
  - firstname: Caroline
    lastname: Gao
    affiliation: [2,4,1]

  - firstname: Jonathan  
    lastname: Karnon
    affiliation: [3]

  - firstname: Luis  
    lastname: Salvador-Carulla
    affiliation: [5]

  - firstname: Sue
    particle: M
    lastname: Cotton
    affiliation: [2,4]

  - firstname: Cathrine
    lastname: Mihalopoulos
    affiliation: [1]

affiliations:
  - number: 1
    corresponding: TRUE
    info:
      orgdiv: School of Public Health and Preventive Medicine
      orgname: Monash University
    address:
        city: Melbourne
        country: Australia
  - number: 2
    corresponding: FALSE
    info:
      orgname: Orygen
    address:
        city: Parkville
        country: Australia
  - number: 3
    corresponding: FALSE
    info:
      orgname: Flinders University   
    address:
        city: Adelaide
        country: Australia 
  - number: 4
    corresponding: FALSE
    info:
      orgdiv: Centre for Youth Mental Health
      orgname: The University of Melbourne
    address:
        city: Parkville
        country: Australia

  - number: 5
    corresponding: FALSE
    info:
#      orgdiv: XXXXXXXXXXX
      orgname: University of Canberra  
    address:
    #     # street: "182 Victoria Square"     
        city: Canberra
    #     postcode: 5000
    #  #   state: South Australia
        country: Australia 
keywords:  
  - computational models
  - ethics of modelling
  - health economics
  - open-source models

pacs: 
  jel:
    - "C63"
    - "C88"
    - "I10"
  msc:
    - "91-08"
# title: "REPLACE ME"
params:
  X: null
abstract: TO BE REPLACED
# output: 
#   pdf_document:
#     keep_tex: true
graphics: true
header-includes: |
  %% Per the spinger doc, new theorem styles can be included using built in style, 
  %% but it seems the don't work so commented below
  %\theoremstyle{thmstyleone}%
  \newtheorem{theorem}{Theorem}%  meant for continuous numbers
  %%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
  %% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
  \newtheorem{proposition}[theorem]{Proposition}%
  %%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

  %% \theoremstyle{thmstyletwo}%
  \theoremstyle{remark}
  \newtheorem{example}{Example}%
  \newtheorem{remark}{Remark}%

  %% \theoremstyle{thmstylethree}%
  \theoremstyle{definition}
  \newtheorem{definition}{Definition}%

  \usepackage{float}
  \let\origfigure\figure
  \let\endorigfigure\endfigure
  \renewenvironment{figure}[1][2] {
      \expandafter\origfigure\expandafter[H]
  } {
      \endorigfigure
  }
  \usepackage{lscape}
  \usepackage{pdflscape}
  \newcommand{\blandscape}{\begin{landscape}}
  \newcommand{\elandscape}{\end{landscape}}
  \usepackage[none]{hyphenat}
  \usepackage[export]{adjustbox}
  \usepackage{multirow}
  \usepackage{longtable}
  \usepackage{array}
  \usepackage{wrapfig}
  \usepackage{colortbl}
  \usepackage{booktabs} 
  \usepackage{colortbl}
  \usepackage{pdflscape}
  \usepackage{tabu}
  \usepackage{threeparttable}
  \usepackage[normalem]{ulem}
  \usepackage{threeparttablex}
  \usepackage{makecell}
  \expandafter\def\csname
  ver@l3backend.sty\endcsname{}
---

```{r echo = F}
## NOTE: EDITED TEMPLATE FILE To Add rotating and multirow packages as per:
# https://stackoverflow.com/questions/31503418/how-to-add-a-latex-package-to-r-journal-template-from-rticle-package-in-r
library(ready4show)
options(tinytex.verbose = TRUE)
# library(xtable)
# options(xtable.floating = FALSE)
# options(xtable.timestamp = "")
```
```{r}
params_ls <- list(X = ready4show::Ready4showSynopsis())
```

```{r timelygls, tab.cap='Transparent, reusable and updatable (TRU) computational health economic models (CHEMs) are suggestive of ethical modelling practice', tab.id = 'timelygls', echo = F, results = 'asis', warning=FALSE, message=FALSE}
library(kableExtra)  
df <- data.frame(` ` = c("Transparent", "Reusable", "Updatable", "Blank"),
                 `Social acceptability` = c("Modeller judgments, model features and verification checks can be reviewed by third parties.",
                                                                        "Models can be modified to reflect alternative value judgments.",
                                                                        "",#"Models can be modified to reflect alternative value judgments.",
                                            "XXXXXXXXXXXXXXXXX"
                                            ),
                 `Fitness for purpose` = c("",#"Modeller judgments, model features and verification checks can be reviewed by third parties.",
                                           "Third party use increases likelihood of errors being identified.",
                                           "Models can be validly used for longer.",
                                           "XXXXXXXXXXXXXXXXX"),
                 `Beneficial use` = c("Decision-makers can understand the strengths and weakness of models before applying them.",
                                      "Can inform more decision problems, in more contexts with less duplicative effort.",
                                      "",#"Models can be validly used for longer."
                                      "XXXXXXXXXXXXXXXXX"),
                 #Blank = c(rep("XXXXXXXXXXXXXXXXXXXXXXXXXXXX",3)),
                 check.names = F)
 library("flextable")
  library("ftExtra")
if(params_ls$X@outp_formats_chr[1] == "Word") {

  df %>%
    ready4show::print_table(output_type_1L_chr = params_ls$X@outp_formats_chr[1],
                            caption_1L_chr = knitr::opts_current$get("tab.cap"),
                            mkdn_tbl_ref_1L_chr = paste0("tab:",knitr::opts_current$get("tab.id")),
                            use_rdocx_1L_lgl = ifelse(params_ls$X@outp_formats_chr[1]=="Word",T,F),
                            add_to_row_ls = NULL,
                            sanitize_fn = NULL) %>%
    ftExtra::colformat_md()
  }else{
    # library(huxtable)
    # library(gt)
    # library(dplyr)
   #flextable::set_flextable_defaults() #fonts_ignore=TRUE
   flextable::flextable(df) %>%
   flextable::border_inner(border = officer::fp_border(style = "none")) %>%
   flextable::border_outer() %>%
   flextable::align(align = "center", part = "header") %>%
   #flextable::padding(j=2:4, part = "all", padding.top = 40, padding.bottom = 40) %>%
   # flextable::merge_h() %>%
   # flextable::merge_v() %>%
   flextable::align(j=2:4, align = "center") %>%
   flextable::bold(j=1, part = "all") %>%
   flextable::bold(part = "header") %>%
   flextable::height(i = 1:3, height = 1.2) %>%
   flextable::hrule(rule = "exact", part = "body") %>% 
   flextable::width(j = 1, width = 1) %>%
   flextable::width(j = 2:4, width = 1.8) %>%
   flextable::merge_at(i=1,j=2:3) %>%
   flextable::merge_at(i=2:4,j=2) %>%
   flextable::merge_at(i=3:4,j=1) %>%
   flextable::merge_at(i=3:4,j=3:4)  
    }
```

### 2024-03-20 setup.R

@@ -13,7 +13,7 @@ X <- Ready4showSynopsis(#background_1L_chr = "Our study is entirely fictional.",
                         keywords_chr = c("computational models", "ethics of modelling", "health economics", "open-source models"),
                         outp_formats_chr = "PDF",
                         #sample_desc_1L_chr = "The study sample is fake data that pretends to be young people aged 12 to 25 years who attended Australian primary care services for mental health related needs between November 2019 to August 2020.",
-                        title_1L_chr = "Identifying and assessing ethical risks when implementing health economic models as software")
+                        title_1L_chr = "Ethical considerations when planning, implementing and releasing health economic model software - a new proposal")
 X <- renewSlot(X,
                "authors_r3", 
                {
@@ -40,11 +40,10 @@ X <- renewSlot(X,
                  args_ls$abstract_ls <- list(Summary = "Most health economic analyses are undertaken with the aid of computers. However, the research ethics of implementing health economic models as software (or computational health economic models (CHEMs)) are poorly understood. 
                                                         We propose that developers and funders of CHEMs should adhere to research ethics principles and pursue the goals of: (i) socially acceptable user requirements and design specifications; (ii) fit for purpose implementations; and (iii) socially beneficial post-release use. 
                                                         We further propose that a transparent (T), reusable (R) and updatable (U) CHEM is suggestive of a project team that has largely met these goals.
-                                                        We propose six criteria for assessing TRU CHEMs: (T1) software files are open access; (T2) developer contributions and judgments on appropriate use are easily identified; 
-                                                        (R1) programming practices promote selective reuse; (R2) licenses permit reuse and derivative works; (U1) maintenance infrastructure is in place; and 
+                                                        We propose six criteria for assessing TRU CHEMs: (T1) software files are publicly available; (T2) developer contributions and judgments on appropriate use are easily identified; 
+                                                        (R1) programming practices promote subcomponent reuse; (R2) licenses permit reuse and derivative works; (U1) maintenance infrastructure is in place; and 
                                                         (U2) releases are systematically retested and deprecated.
-                                                        Few existing CHEMs would meet all TRU criteria. Addressing these limitations will require changes to the work practices, funding and tools that underpin health economic model development.
-                                                        Priority actions include the development of new and updated good practice guidelines and research infrastructure and human capital investments."
+                                                        Few existing CHEMs would meet all TRU criteria. Addressing these limitations will require the development of new and updated good practice guidelines and investments by governments and other research funders in enabling infrastructure and human capital."
[…]

### 2024-03-25 Markdown/Manuscript/Child_RMDs/MS_Intro_Short.Rmd
pre submission polish

@@
 CHEMs can be developed using commercial modelling <!-- Addition -->or spreadsheet<!-- /Addition --> software (for example, Microsoft Excel [@msexcel]) or open-source programming languages such as Python [@python2009] or R [@RCORE2022].
-Training and professional practice guidance for health economists typically provide only superficially coverage of software development topics.
-However, these capabilities will be critical for achieving proposed methodological innovations at the frontier of the health economics field.
-Opportunities for field development that closely overlap with CHEM implementation approaches include living health technology assessments [@thokala2023living], machine learning powered health economic analyses [@padula2022machine], open source health-economic models [@Pouwels2022] and utilising Large Language Models (LLMs) to program health economic models [@poirrier2023msr26].
+Training and professional practice guidance for health economists typically provide only superficial coverage of software development topics.
+However, these capabilities may be important for achieving a number of proposed methodological innovations for advancing the health economics field.
+Living health technology assessments [@thokala2023living], machine learning powered health economic analyses [@padula2022machine], open source health-economic models [@Pouwels2022] and the use of Large Language Models (LLMs) to program health economic models [@poirrier2023msr26] all require careful consideration of CHEMs as software development projects.
 <!-- Potential enhancements to health economic modelling practice include improved model transferability [@munthe2019systematic; @barbieri2010international], greater use of reference models [@Afzali2013], living health technology assessments [@thokala2023living] and leveraging the analytic power of machine learning [@padula2022machine].
@@
 
-There is a significant ethical dimension to software development.
-For example, recent developments in LLMs mean software developers may confront novel ethical issues including: new security risks; more restrictive open-source licensing; environmentally harmful levels of computing power; and concerns about model explainability and unintended consequences [@ozkaya2023application].
-The health economics community should engage more directly with these ethical debates.
-The computational modelling landscape is changing fast [@calder2018computational] and the ethical <!-- responsibilities  --><!-- Edit -->dimensions of computational public health modelling <!-- /Edit -->remain poorly understood [@10.3389/fpubh.2017.00068].
-Documents such as the World Health Organisation's recent guidance on generative artificial intelligence [@world2024ethics] can be important reference points in helping the health economics discipline navigate relevant ethical topics in software development.
-
+There are important ethical dimensions specific to software development that the health economics community should engage with more directly.
+For example, recent developments in LLMs pose novel ethical issues for software developers including: new security risks; more restrictive open-source licensing; environmentally harmful levels of computing power; and concerns about model explainability and unintended consequences [@ozkaya2023application].
+The ethical <!-- responsibilities  --><!-- Edit -->dimensions of computational public health modelling <!-- /Edit -->remain poorly understood [@10.3389/fpubh.2017.00068].
+However, documents such as the World Health Organisation's recent guidance on generative artificial intelligence [@world2024ethics] can be reference points to help identify ethical topics relevant to health economic model software development.
+<!-- computational modelling landscape is changing fast [@calder2018computational] and the -->
   <!-- Moved -->

### 2024-03-25 Markdown/Manuscript/Child_RMDs/MS_Tables_1.Rmd
for coauthor feedback draft
pre submission polish

@@
 ---
-```{r proscons, tab.cap='Implications of choosing commercial modelling or spreadsheet software or an open-source programming language for the computational implementation of health economic models', tab.id = 'proscons', echo = F, results = 'asis'}
+```{r proscons, tab.cap='Implications of choosing proprietary software or an open-source programming language for the computational implementation of health economic models', tab.id = 'proscons', echo = F, results = 'asis'}
 library(kableExtra)  
 df <- data.frame(`Implications` = c(rep("Resources",4), rep("Ethical",10)),
-                 `Commercial modelling or spreadsheet software` = c("(+) Will include tools to efficienty accomplish many common modelling tasks.",
+                 `Proprietary modelling or spreadsheet software` = c("(+) Will include tools to efficiently accomplish many common modelling tasks.",
                                                      "(+) Requires limited or no programming skills.",
@@
                                                      "(+) Popular tools (e.g. Excel) are readily understood by many.",
-                                                     "(-) Requirement to manually transfer results to manuscripts through repeated copy and paste actions or transcription from visual inspection increases risk of (undetected) reporting errors.",
+                                                     "(-) Manual transfer of results to manuscripts through copy and paste or transcription from visual inspection increases risk of (undetected) reporting errors.",
                                                      "",
@@
                                                         "(i) artefacts (code and data) from pre-existing models are leveraged; and / or",
-                                                        "(ii)	maintaining and transferring (e.g. to new decision problems / contexts) model.",
+                                                        "(ii)	models are to be maintained and transferred (e.g. to new decision problems / contexts).",
                                                         "Transparency",
@@
     df %>%
-      dplyr::filter(!`Commercial modelling or spreadsheet software` %in% c("Transparency", "Reusability","Updatability")) %>%
+      dplyr::filter(!`Proprietary modelling or spreadsheet software` %in% c("Transparency", "Reusability","Updatability")) %>%
       kableExtra::kbl(booktabs = T, 

### 2024-03-25 Markdown/Manuscript/Child_RMDs/MS_Tables_3.Rmd
pre submission polish

@@
                                "It is easy to see who funded, developed and tested each part of the CHEM and to identify the modelling team’s assumptions, judgments and theories about CHEM development and use.",
-                               "Programming practices facilitate selective reuse of model sub-components.",
+                               "Programming practices facilitate independent reuse of model components.",
                                "Licenses allow anyone to ethically reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works.",

### 2024-03-25 setup.R

@@ -41,7 +41,7 @@ X <- renewSlot(X,
                                                         We propose that developers and funders of CHEMs should adhere to research ethics principles and pursue the goals of: (i) socially acceptable user requirements and design specifications; (ii) fit for purpose implementations; and (iii) socially beneficial post-release use. 
                                                         We further propose that a transparent (T), reusable (R) and updatable (U) CHEM is suggestive of a project team that has largely met these goals.
                                                         We propose six criteria for assessing TRU CHEMs: (T1) software files are publicly available; (T2) developer contributions and judgments on appropriate use are easily identified; 
-                                                        (R1) programming practices promote subcomponent reuse; (R2) licenses permit reuse and derivative works; (U1) maintenance infrastructure is in place; and 
+                                                        (R1) programming practices facilitate independent reuse of model components; (R2) licenses permit reuse and derivative works; (U1) maintenance infrastructure is in place; and 
                                                         (U2) releases are systematically retested and deprecated.
                                                         Few existing CHEMs would meet all TRU criteria. Addressing these limitations will require the development of new and updated good practice guidelines and investments by governments and other research funders in enabling infrastructure and human capital."
                                              # ,
